\section{Summary}

%% \subsection{Ongoing and future work}
Here is the list of ongoing and future work:
\begin{itemize}
\item \emph{Embedding Nax into strongly normalizing calculus}
	As mentioned in \S\ref{sec:theory}, we can prove the desired properties
	of Nax by embedding the features of Nax, most importantly,
	the Mendler-style recursion combinators into the calculus
	equipped with those desired properties. We have already done this for
	some of the combinators: \textsf{MIt} and \textsf{MsfIt} into \Fi, and,
	\textsf{MPr} into \Fixi. We believe that the course-of-values
	combinators \textsf{McvIt} and \textsf{McvPr} are also embeddable
	into \Fixi.
\item \emph{Correctness of the type inference algorithm}\\
	We are describing a type inference algorithm for Nax,
	which extends the type inference algorithm $W$ for HM.
	We will prove its correctness with respect to the typing rules of Nax.
\item \emph{Large eliminations}\\
	The current prototype of Nax only supports limited
	form of large elimination of mapping indices from
	argument types to result types due to the limited syntax
	of the index transformer. We are trying to enrich the index
	transformer syntax for more expressive large eliminations
	(\eg, if-then-else or more generally case expressions
	in index transformers) and still maintain our design goals of
	having type inference and logically consistent type system.
	Again, we will make sure that such new features are safe by embedding
	the new feature into the calculus Nax is based on, such as System \Fi.
\item \emph{Non-logical language fragments}\\
	The prototype of the Nax language described by this paper
	only contains strongly normalizing features for logical consistency.
	Our eventual goal is include non-logical features into the language
	and still maintain the desired features within the subset of
	the language distingiusehd by the type system. We plan to
	adopt ideas of the Trellys project
	\cite{plpv12TrellysEq,msfp12TrellysHEq,msfp12TrellysSN4GenRec}
	to approach our eventual goal.
\end{itemize}

