%\documentclass[preliminary,copyright,creativecommons]{eptcs}
%\documentclass[submission,copyright,creativecommons]{eptcs}
\documentclass[adraft,copyright,creativecommons]{eptcs}
\providecommand{\event}{FICS or MSFP 2012} % Name of the event to submit
\usepackage{breakurl}              % Not needed if you use pdflatex only.
\usepackage{amssymb}
\usepackage[fleqn]{amsmath}
\usepackage{amsthm}
\usepackage{semantic}
\usepackage{color}
\usepackage{fancybox}
\usepackage{framed}

\definecolor{grey}{rgb}{0.8,0.8,0.8}

\title{Nax theory}

\author{Ki Yung Ahn \qquad Tim Sheard
\institute{Portland State University\thanks{Thanks here if needed}\\
           Portland, Oergon, USA}
\email{kya@cs.pdx.edu \qquad sheard@cs.pdx.edu}
\and
Marcelo P. Fiore \quad\qquad Andrew M. Pitts
\institute{University of Cambridge\thanks{Another thanks here if needed}\\
           Cambridge, UK}
\email{\{Marcelo.Fiore,Andrew.Pitts\}@cl.cam.ac.uk}
}
\def\titlerunning{title running}
\def\authorrunning{KY. Ahn, T. Sheard \& M. P. Fiore, A. M. Pitts}

%%%%% cammands for comments
\newcommand{\KYA}[1]{\textcolor{magenta}{#1 --KYA}}

\theoremstyle{plain}
\newtheorem{proposition}{Proposition}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}

\theoremstyle{remark}
\newtheorem{remark}{Remark}

\theoremstyle{definition}
\newtheorem{definition}{Definition}


\newcommand{\Fig}[1]{Figure\,\ref{fig:#1}}
\newcommand{\newFi}[1]{\colorbox{grey}{\ensuremath{#1}}}

\newcommand{\eg}{{e.g.}}
\newcommand{\ie}{{i.e.}}

\newcommand{\Fi}{\ensuremath{\mathsf{F}_i}}
\newcommand{\Fw}{\ensuremath{\mathsf{F}_\omega}}
\newcommand{\fix}{\mathsf{fix}}
\newcommand{\Fix}{\mathsf{Fix}}
\newcommand{\Fixw}{\ensuremath{\Fix_{\omega}}}
\newcommand{\Fixi}{\ensuremath{\Fix_{i}}}

\newcommand{\Nat}{\ensuremath{\mathsf{Nat}}}
\newcommand{\Bool}{\ensuremath{\mathsf{Bool}}}
\newcommand{\sfList}{\ensuremath{\mathsf{List}}}
\newcommand{\sfVec}{\ensuremath{\mathsf{Vec}}}
\newcommand{\SAT}{\ensuremath{\mathsf{SAT}}}

\newcommand{\oz}{\oldstylenums{0}}
\newcommand{\ka}{{\check\kappa}}

\newcommand{\calS}{\mathcal{S}}
\newcommand{\calA}{\mathcal{A}}
\newcommand{\calR}{\mathcal{R}}
\newcommand{\dom}{\mathop{\mathsf{dom}}}
\newcommand{\FV}{\mathop{\mathrm{FV}}}
\newcommand{\MIt}{\mathsf{MIt}}
\newcommand{\In}{\mathsf{In}}
\newcommand{\case}{\mathsf{case}}
\newcommand{\sflet}{\mathsf{let}}
\newcommand{\sfin}{\mathsf{in}}
\newcommand{\Let}[3]{\sflet\;#1 = #2\;\sfin\; #3}

\newcommand{\defeq}{\stackrel{\mathrm{def}}{=}}

\begin{document}
\maketitle

\begin{abstract}
This is a sentence in the abstract.
This is another sentence in the abstract.
This is yet another sentence in the abstract.
This is the final sentence in the abstract.
\end{abstract}

\input{intro}

\section{Features of Nax (informal introduction with examples??)}\label{sec:bg}
\subsection{Indexed types with static indices}\label{sec:bg:ixty}

A representative example of a term
indexed type is the length indexed list type (often called the vector type).
A regular polymorphic list type ($\sfList\;a$) is parametrized by a type
parameter ($a$), which can be instantiated to a specific type, while the
vector type ($\sfVec\;a\;\{n\}$) has an additional term index ($n$), which
can be instantiated to a specific natural number value. The curly brackets
($\{\cdots\}$) around $n$ is to syntactically distinguish term indices from
other type arguments. 

\subsection{Recursive types of unrestricted polarity but restricted elimination}
\label{sec:bg:recty}
TODO

\subsection{Mendler style iteration and recursion combinators}
\label{sec:bg:Mendler}

%% \section{TODO}

\input{nax}

\input{fi}

%% \input{fixi}

%% \section{TODO}

\section{TODO}
let's write a paper for maybe one of the following venues?
\begin{itemize}
\item FICS \url{http://www.inf.u-szeged.hu/fics2012/}\\
Abstract submission: 	4 Dec 2011\\
Paper submission: 	11 Dec 2011\\
8 pages using eptcs style
\item MSFP \url{http://cs.ioc.ee/msfp/msfp2012/}\\
Submission of papers 16 December\\
There is no specific page limit but authors should strive for brevity.
\end{itemize}

~\\~\\
\cite{article,bookA} just dummy citation %% \nocite{*}
\bibliographystyle{eptcs}
\bibliography{main}

\end{document}
