%This is a template for producing LIPIcs articles.
%See lipics-manual.pdf for further information.

\documentclass[a4paper,UKenglish]{lipics}
  %for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
  %for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
 % for section-numbered lemmas etc., use "numberwithinsect"
 
\usepackage{microtype}%if unwanted, comment out or use option "draft"

%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory

\bibliographystyle{plain}% the recommended bibstyle

% Author macros %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Mendler-style Recursion Schemes for Mixed-Variant Datatypes\footnote{
	This work was partially supported by someone.} }
%% \titlerunning{A Sample LIPIcs Article} %optional, in case that the title is too long; the running title should fit into the top page column

\author[1]{Ki Yung Ahn}
\author[1]{Tim Sheard}
\author[2]{Marcelo Fiore}
\affil[1]{Department of Computer Science, Portland State University\\
  Oregon, USA\\
  \texttt{\{kya,sheard\}@cs.pdx.edu}}
\affil[2]{Computer Laboratory, University of Cambridge\\
  Cambridge, UK\\
  \texttt{Marcelo.Fiore@cl.cam.ac.uk}}
\authorrunning{KY. Ahn, T. Sheard and M. Fiore} %mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et. al.'

\Copyright{Ki Yung Ahn, Tim Sheard and Marcelo Fiore}%mandatory. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

\subjclass{
D.3.3 [Programming Languages]: Language Constructs and Features
--- Data types and structures, Polymorphism, and Recursion;
F.3.3 [Logics and Meanings of Programs]: Studies of Program Constructs
-- Functional constructs, Program and recursion schemes, and Type structure;
F.4.1 [Mahtematical Logic and Formal Systems]: Mathematical Logic
--- Lambda calculus and related systems
}
\keywords{
Mendler-style recursion, Higher-Order Abstract Syntax (HOAS),
mixed-variant datatypes, negative datatypes, termination}
% mandatory: Please provide 1-5 keywords
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%Editor-only macros (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\serieslogo{}%please provide filename (without suffix)
\volumeinfo%(easychair interface)
  {Billy Editor, Bill Editors}% editors
  {2}% number of editors: 1, 2, ....
  {Conference title on which this volume is based on}% event
  {1}% volume
  {1}% issue
  {1}% starting page number
\EventShortName{}
\DOI{10.4230/LIPIcs.xxx.yyy.p}% to be completed by the volume editor
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newcommand{\cf}[0]{{cf.}}
\newcommand{\eg}[0]{{e.g.}}
\newcommand{\ie}[0]{{i.e.}}
\newcommand{\aka}[0]{{a.k.a.}}

\newcommand{\F}[0]{{\ensuremath{\mathsf{\uppercase{F}}}}}
\newcommand{\Fw}[0]{{\ensuremath{\mathsf{\uppercase{F}}_{\!\omega}}}}
\newcommand{\Fixw}[0]{{\ensuremath{\mathsf{\uppercase{F}\lowercase{ix}}_{\omega}}}}
\newcommand{\Fwhat}[0]{{\ensuremath{\mathsf{\uppercase{F}}_{\!\omega}\!\!\char`\^}}}

\newcommand{\MIt}[1]{\ensuremath{\textbf{\textit{mit}}_{#1}}}
\newcommand{\MPr}[1]{\ensuremath{\textbf{\textit{mpr}}_{#1}}}
\newcommand{\mpr}[1]{\ensuremath{\textbf{\textit{mpr}}_{#1}}}
\newcommand{\MsfIt}[1]{\ensuremath{\textbf{\textit{msfit}}_{#1}}}
\newcommand{\msfit}[1]{\ensuremath{\textbf{\textit{msfit}}_{#1}}}
\newcommand{\mprsi}[1]{\ensuremath{\textbf{\textit{mprsi}}_{#1}}}
\newcommand{\lift}[0]{\textit{lift}}
\newcommand{\In}[1]{\ensuremath{\textbf{\textit{In}}_{#1}}}
\newcommand{\inL}[0]{\textit{in}_{L}}
\newcommand{\inR}[0]{\textit{in}_{R}}

\newcommand{\textcomment}[1]{\rm\color{blue}#1}

%% \setlength{\thickmuskip}{0mu}
%% \setlength{\medmuskip}{0mu}
%% \setlength{\thinmuskip}{0mu}
\lstset{captionpos=t,
	float,
	keepspaces=false,
	abovecaptionskip=-\medskipamount,
	language=Haskell,
 	basicstyle=\it,
 	keywordstyle=\rm\bfseries,
	commentstyle=\color{blue},
	escapechar={\%},
	stringstyle=\rm\ttfamily,
	deletekeywords={Bool,Int,Integer,String,show,const,id,length,List},
 	literate=
 		{forall}{{$\forall$}}1
 		{:}{{$:$}}1
 		{;}{{$;$}}1
 		{::}{{$::$}}1
 		{->}{{$\to$}}2
 		{=}{{$=$}}1
 		{=>}{{$\Rightarrow$}}2
 		{*}{{$*$}}1
 		{\\}{{$\lambda$}}1
		{Phi0}{{Phi$_{*}$}}3
		{Phi1}{{Phi$_{*\to*}$}}5
		{Phi0'}{{Phi$_{*}^{\phantom{.}\prime}$}}3
		{Phi1'}{{Phi$_{*\to*}^{\phantom{.}\prime}$}}5
		{Mu0}{{$\mu_{*}$}}1
		{In0}{{\In{*}}}2
		{out0}{{out$_{*}$}}3
		{Mu1}{{$\mu_{*\to*}$}}3
		{In1}{{\In{*\to*}}}4
		{out1}{{out$_{*\to*}$}}5
		{Rec0}{{$\mu_{*}^\prime$}}1
		{Roll0}{{$\In{*}^\prime$}}2
		{unRoll0}{{unIn$_{*}^\prime$}}4
		{Inverse0}{{Inverse$_{*}$}}6
		{Rec1}{{$\mu_{*\to*}^\prime$}}3
		{Roll1}{{$\In{*\to*}^\prime$}}4
		{unRoll1}{{unIn$_{*\to*}^\prime$}}6
		{Inverse1}{{Inverse$_{*\to*}$}}8
 		{phi}{{$\varphi$}}1
 		{t1}{{$t_1$}}1
 		{t1}{{$t_1$}}1
 		{t2}{{$t_2$}}1
 		{e1}{{$e_1$}}1
 		{e2}{{$e_2$}}1
		{mpr}{{mpr}}3
		{mpr1}{{mpr$_{* \to *}$}}6
		{mprim}{{\mpr{}}}3
		{mprim1}{{\mpr{* \to *}}}6
		{mprsi}{{\mprsi{}}}5
		{mprsi1}{{\mprsi{* \to *}}}7
		{uncast}{{uncast}}6
		{cast}{{cast}}4
		{call}{{call}}4
		{unVal}{{un\!Val}}4
		{mcata}{{\MIt{}}}3
		{mcata0}{{\MIt{*}}}3
		{mcata1}{{\MIt{* \to *}}}5
		{msfcata}{{\msfit{}}}4
		{msfcata'}{{\msfit{}$^\prime$}}5
		{msfcata0}{{\msfit{*}}}5
		{msfcata1}{{\msfit{* \to *}}}7
		{++}{{$+\!\!\!\!+$}}2
 		{+}{{$+$}}1
 		{(}{{$($}}1
 		{)}{{$)$}}1
 		{[}{{$[$}}1
 		{]}{{$]$}}1
 		{\{}{{$\{$}}1
 		{\}}{{$\}$}}1
 		{|}{{$\mid$}}1
		{Exp}{{Exp}}3
		{Expr}{{Expr}}4
		{Exp_u}{{Exp$_u$}}3
		{ExpF}{{ExpF}}4
		{ExpF_u}{{ExpF$_u$}}4
		{App_u}{{App$_u$}}4
		{Lam_u}{{Lam$_u$}}4
		{V_u}{{V\!$_u$}}1
		{Val}{{Val}}3
		{VFun_u}{{VFun$_u$}}4
		{Exp'}{{Exp$\,^\prime$}}4
		{show'}{{show$\,^\prime$}}4
		{showExp}{{showExp}}7
		{const}{const}4
		{String}{String}5
		{Bool}{{Bool}}4
		{Int}{{Int}}3
		{inv}{{inv}}3
		{ev}{{ev}}2
		{val}{{val}}2
		{eval}{{eval}}3
		{veval}{{veval}}4
		{vs}{{vs}}2
		{vars}{{vars}}3
		{lam}{{lam}}3
		{app}{{app}}3
		{Id}{{Id}}1
		{MkId}{{MkId}}4
		{unId}{{unId}}4
		{exp2expr}{{exp2expr}}7
		{expr2exp'}{{expr2exp$^\prime$}}7
		{exp'2expr}{{exp$^\prime$2expr}}7
		{p2r}{{p2r}}2
}


\begin{document}

\maketitle

\begin{abstract}
Some concepts, such as Higher-Order Abstract Syntax (HOAS),
are most naturally expressed by \emph{mixed-variant datatypes}
(\aka\ negative (recursive) datatypes). Unfortunately,
mixed-variant datatypes are often outlawed in formal reasoning systems
based on the Curry--Howard correspondence (\eg, Coq, Agda), because
the conventional recursion schemes (or, induction principles) supported in
such systems cannot guarantee termination for mixed-variant datatypes.

There is an alternative style of formulating recursion schemes,
known as the Mendler style, that can guarantee termination for
arbitrary datatypes. Ahn and Sheard \cite{AhnShe11} formulated
a Mendler-style recursion scheme (\msfit{}), and demonstrated
examples involving regular (\ie, non-indexed) mixed-variant datatypes
(\eg, untyped $\lambda$-calculus in HOAS). Their examples demonstrate
an advantage of the Mender style -- termination guarantee for
arbitrary datatypes, including mixed-variant ones. They proved
termination of the examples via embedding into System~\Fw.

Another advantage of the Mendler style is that recursion schemes
naturally extend to non-regular (\ie, indexed) datatypes.
Our type-preserving evaluator example for the simply-typed HOAS, defined as
a type-indexed mixed-variant datatype, demonstrates both advantages of
the Mendler style. This example has its own theoretical value, which is
a novel discovery that the simply-typed HOAS evaluation is expressible
within System~\Fw. To our knowledge, this is the first report of
a simply-typed HOAS evaluator (without translation through first-order syntax)
that is equipped with correct-by-construction proofs
(in the Curry--Howard sense) of both type-preservation and normalization.
We also develop further theoretical discussions on the \Fw-embedding of \msfit{}
and propose yet another recursion scheme (\mprsi{}), which is also useful for
mixed-variant datatypes.

Our study promotes the usefulness of the recursion schemes,
which are useful over mixed-variant datatypes. We hope our
work motivates designing logical reasoning systems that support
wider range of datatypes including mixed-variant ones.
\end{abstract}

\input{intro}
\input{mendler}
\input{eval}
\input{murec}
\input{theory}
\input{related}
\input{mprsi}
\input{summary}

\section*{Acknowledgements}
%% \subparagraph*{Acknowledgements}
I want to thank \dots

\appendix
\section{TODO Appendix Section}
TODO

\bibliography{main}

\end{document}
