%-----------------------------------------------------------------------------
%
%               Template for sigplanconf LaTeX Class
%
% Name:         sigplanconf-template.tex
%
% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
%               file for SIGPLAN conference proceedings.
%
% Guide:        Refer to "Author's Guide to the ACM SIGPLAN Class,"
%               sigplanconf-guide.pdf
%
% Author:       Paul C. Anagnostopoulos
%               Windfall Software
%               978 371-2316
%               paul@windfall.com
%
% Created:      15 February 2005
%
%-----------------------------------------------------------------------------


\documentclass[preprint]{sigplanconf}

% The following \documentclass options may be useful:
%
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.

\usepackage[fleqn]{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{semantic}
\usepackage{color}
\usepackage{fancybox}
\usepackage{framed}
\usepackage{listings}
\usepackage{todonotes}
%% \usepackage{comment}

\newcommand{\KYA}[1]{\todo[inline,author=KiYung,size=\small,color=pink]{#1}}

\definecolor{grey}{rgb}{0.8,0.8,0.8}

\newcommand{\Fig}[1]{Figure\,\ref{fig:#1}}
\newcommand{\newFi}[1]{\colorbox{grey}{\ensuremath{#1}}}

\newcommand{\eg}{{e.g.}}
\newcommand{\ie}{{i.e.}}

\newcommand{\Fi}{\ensuremath{\mathsf{F}_i}}
\newcommand{\Fw}{\ensuremath{\mathsf{F}_\omega}}
\newcommand{\fix}{\mathsf{fix}}
\newcommand{\Fix}{\mathsf{Fix}}
\newcommand{\Fixw}{\ensuremath{\Fix_{\omega}}}
\newcommand{\Fixi}{\ensuremath{\Fix_{i}}}

\renewcommand{\l}{\lambda}

\newcommand{\dom}{\mathop{\mathsf{dom}}}
\newcommand{\FV}{\mathop{\mathrm{FV}}}

\theoremstyle{plain}
\newtheorem{proposition}{Proposition}
\newtheorem*{proposition*}{Proposition}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}

\theoremstyle{remark}
\newtheorem{remark}{Remark}

\theoremstyle{definition}
\newtheorem{definition}{Definition}



\begin{document}

\conferenceinfo{WXYZ '05}{date, City.} 
\copyrightyear{2005} 
\copyrightdata{[to be supplied]} 

\titlebanner{banner above paper title}        % These are ignored unless
\preprintfooter{short description of paper}   % 'preprint' option specified.

\title{System \Fi}
\subtitle{a higher-order polymorphic $\lambda$-calculus with erasable term indices}

\authorinfo{Ki Yung Ahn\and Tim Sheard}
           {Portland State University}
	   {\{kya,sheard\}@cs.pdx.edu}

\authorinfo{Marcelo Fiore\and Andrew M. Pitts}
           {University of Cambridge}
	   {\{Marcelo.Fiore,Andrew.Pitts\}@cl.cam.ac.uk}

\maketitle

\begin{abstract}

We wish to incorporate dependent types into ordinary programming
languages. We are interested in two kinds of dependent types.
Full dependency, where the type of a function can depend upon the value
of its run-time parameters, and static dependency, where the type
of a function can depend only upon static (or compile-time) parameters.
Static dependency is sometimes referred to as indexed typing. The first
is very expressive, while the second is often much easier to learn and 
use. Indexed types come in two flavors: type-indexed and term-indexed
types. Type indexing includes parametric polymorphism, but it also
includes more sophisticated typing such as Generalized Algebraic 
Datatypes~(GADTs). Term-indexed types include indices that range over data
structures, such as the Natural Numbers~(like {\tt Z}, {(\tt S
Z})) or Lists~(like {\tt Nil} or ({\tt Cons 1 Nil})) rather than types
such as {\tt Int}. The classic example of a term index is the second
parameter to the length-indexed list type {\tt Vec}~(as in ({\tt Vec Int
(S Z)})).  In languages such as Haskell, which support GADTs,
term-indices are not first-class; they are faked by reflecting data at the
type level with uninhabited type constructors.

The purpose of this paper is to introduce a foundational type system,
System~\Fi, for the design of programming languages with first-class
term-indexed datatypes; thus avoiding reflecting data. To do this, we have
devised a minimal extension of System \Fw\ that incorporates term indices.
While term-indexed datatypes are expressible in richer type systems, like
the Implicit Calculus Constructions~(ICC), such systems come coupled with
orthogonal features such as large eliminations and dependent types.  We
argue that there are important pedagogical benefits of isolating the
minimal features to support term-indexing. We show that System~\Fi\
provides a theory for analysing programs with term-indexed types and also
argue that it constitutes a basis for the design of light-weight
logically-sound dependent programming languages.

In terms of expressivity, System~\Fi\ sits in between System~\Fw\ (the
prototypical logical calculus for functional programming) and ICC~(a
full-featured dependent type theory).  Indeed, we relate System~\Fi\ to
System~\Fw\ and ICC as follows.  We establish erasure properties of
System~\Fi\ that capture the idea that term indices are discardable in
that they are irrelevant for computation.  Erased \mbox{\Fi-terms} are
\Fw-terms; so System~\Fi\ inherits the strong-normalisation property from
System~\Fw.  On the other hand, the logical consistency of System~\Fi\ is
established by embedding it into a subset of ICC.
\end{abstract}

\category{D.3.3}{Programming Languages}{Language Constructs and Features}[data types and structures]
\category{F.4.1}{Mathematical Logic and Formal Languages}{Mathematical Logic}[lambda calculus and related systems]

\terms
Languages, Theory

\keywords
term-indexed data types, generalized algebraic data types,
higher-order polymorphism, type-constructor polymorphism, higher-kinded types,
impredicative encoding

\input{intro}
\input{fi}
\input{data}
\input{theory}
\input{relwork}
\input{concl}




\appendix
\section{Appendix Title}

This is the text of the appendix, if you need one.

\acks

Acknowledgments, if needed.

% We recommend abbrvnat bibliography style.

\bibliographystyle{abbrvnat}

% The bibliography should be embedded for final submission.

\bibliography{main}

\end{document}
