\documentclass[12pt]{report}
%% \usepackage[bbgreekl]{mathbbol}
\usepackage{psu-thesis}
%% \usepackage[T1]{fontenc}
%% \usepackage[latin1]{inputenc}
\usepackage{amssymb}
\usepackage{amsthm}
%% \usepackage{txfonts}
%% \usepackage{pxfonts}
%% \usepackage{eulervm}
\usepackage{lmodern}
\usepackage[sort&compress,square,comma,numbers,longnamesfirst]{natbib}
\usepackage{semantic}
\usepackage[hang]{caption}
\usepackage{color}
\usepackage{tcolorbox}
\usepackage{comment}
\usepackage{stmaryrd}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{textcomp}
\usepackage{dashbox}
%% \usepackage{ulem} %% http://tex.stackexchange.com/questions/27258/how-do-i-write-underline-text-but-with-a-dotted-line
\usepackage{lscape}
\usepackage{afterpage}
\usepackage{url}
\usepackage{listings}
\usepackage[normal]{subfigure}
\usepackage[all]{xy}
\usepackage{enumerate}
\usepackage[shortlabels]{enumitem}
\usepackage{makeidx}
% \usepackage[backref]{hyperref}
\usepackage{hyperref}

\usepackage{mylhs2tex}

\makeindex

\definecolor{grey}{rgb}{0.8,0.8,0.8}
\definecolor{gray}{rgb}{0.57,0.57,0.57}

\setlist{partopsep=-3em}
\setlist{topsep=-3em}
\setlist{parsep=-2em}
\setlist{itemsep=-.2em}

\newcommand{\KYA}[1]{\textcolor{blue}{#1 -- Ki Yung}}
\newcommand{\TS}[1]{\textcolor{red}{#1 -- Tim}}

%% http://www.latex-community.org/forum/viewtopic.php?f=37&t=6637

%% \usepackage{blindtext}
%% %% http://www.latex-community.org/forum/viewtopic.php?f=48&t=6262
%% %% http://www.phil.cam.ac.uk/teaching_staff/Smith/logicmatters/l4llogiciansnew.html
%% \DeclareSymbolFont{symbolsC}{U}{txsyc}{m}{n}
%% \DeclareMathSymbol{\strictif}{\mathrel}{symbolsC}{74}
%% \DeclareMathSymbol{\boxright}{\mathrel}{symbolsC}{128}

\newcommand{\Fig}[1]{Figure~\ref{fig:#1}}

\newcommand{\cf}[0]{{cf.}}
\newcommand{\eg}[0]{{e.g.}}
\newcommand{\ie}[0]{{i.e.}}
\newcommand{\aka}[0]{{a.k.a.}}
%% \newcommand{\wrt}[0]{{w.r.t.}}

\renewcommand{\l}{\lambda}
\newcommand{\bbl}{\boldsymbol\lambda}
\renewcommand{\L}{\Lambda}
\renewcommand{\S}{Section~}

\newcommand{\fix}{\ensuremath{\mathsf{fix}}}
\newcommand{\dom}{\mathop{\mathrm{dom}}}
\newcommand{\ran}{\mathop{\mathrm{ran}}}
\newcommand{\FV}{\mathop{\mathrm{FV}}}
\newcommand{\unify}{\mathop{\mathrm{unify}}}
\newcommand{\occurs}{\mathop{\mathrm{occurs}}}

\newcommand{\F}[0]{{\ensuremath{\mathsf{\uppercase{F}}}}}
\newcommand{\Fw}[0]{{\ensuremath{\mathsf{\uppercase{F}}_{\!\omega}}}}
\newcommand{\Fwhat}[0]{{\ensuremath{\mathsf{\uppercase{F}}_{\!\omega}\!\!\char`\^}}}
\newcommand{\Fi}[0]{{\ensuremath{\mathsf{\uppercase{F}}_{\lowercase{i}}}}}
\newcommand{\Fixw}[0]{{\ensuremath{\mathsf{\uppercase{F}\lowercase{ix}}_{\omega}}}}
\newcommand{\Fixi}[0]{{\ensuremath{\mathsf{\uppercase{F}\lowercase{ix}}_{\lowercase{i}}}}}

\newcommand{\karrow}[1]{\stackrel{#1\;}{\pmb{\pmb{->}}}}
\newcommand{\rrarrow}[0]{\rightrightarrows}

\newcommand{\SN}[0]{\mathsf{SN}}
\newcommand{\SAT}[0]{\mathsf{SAT}}

\newcommand{\unIn}{\ensuremath{\mathsf{unIn}}}
\newcommand{\In}{\ensuremath{\mathsf{In}}}
\newcommand{\It}[0]{\ensuremath{\mathbf{It}}}
\newcommand{\PR}[0]{\ensuremath{\mathbf{Pr}}}
\newcommand{\MIt}{\ensuremath{\mathbf{mit}}}
\newcommand{\McoIt}{\ensuremath{\mathbf{mcoit}}}
\newcommand{\MPr}{\ensuremath{\mathbf{mpr}}}
\newcommand{\MsfIt}{\ensuremath{\mathbf{msfit}}}
\newcommand{\McvIt}{\ensuremath{\mathbf{mcvit}}}
\newcommand{\MsfcvIt}{\ensuremath{\mathbf{msfcvit}}}
\newcommand{\McvPr}{\ensuremath{\mathbf{mcvpr}}}
\newcommand{\McfvPr}{\ensuremath{\mathbf{msfcvpr}}}
\newcommand{\mprsi}[0]{\ensuremath{\mathbf{mprsi}}}

\newcommand{\rulename}[1]{{\small\sc #1}}

%% \mathlig{:.}{\,\text{:}\,} %%% for compact layout of the colon for (\(x:A).t)
\mathlig{=?=}{\stackrel{?}{=}}
\mathlig{|-s}{\vdash\stackrel{\!\mathsf{s}}{}}
\mathlig{~>}{\rightsquigarrow}
\mathlig{[|}{\llbracket} %% since semantic defines |[ as \llbracket :(
\reservestyle[\mathinner]{\command}{\mathbf}
\command{case[case\;], data[data\;], let[let\;], of[\;of\;], in[\;in\;], if[if\;], then[\;then\;], else[\;else\;], where[\;where\;], when[\;when\;]}

\newtheorem{proposition}{Proposition}
\newtheorem*{proposition*}{Proposition}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\newtheorem{conjecture}{Conjecture}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}
\theoremstyle{remark}
\newtheorem{remark}{Remark}

\numberwithin{definition}{section}
\numberwithin{equation}{section}
\numberwithin{proposition}{section}
\numberwithin{conjecture}{section}
\numberwithin{theorem}{section}
\numberwithin{lemma}{section}
\numberwithin{corollary}{section}
\numberwithin{example}{section}
\numberwithin{remark}{section}


\newcommand{\Eq}{\mathtt{Eq}}
\newcommand{\LEq}{\mathtt{LEq}}
\newcommand{\Ran}{\mathtt{Ran}}
\newcommand{\Lan}{\mathtt{Lan}}
\newcommand{\s}[1]{\{#1\}}
%% \renewcommand\qedsymbol{$\blacksquare$}

\begin{document}
\title{The Nax language:\\
       unifying functional programming and logical reasoning in a language\\
       based on Mendler-style recursion schemes and term-indexed types}
\titleline{The Nax language}

\author{Ki Yung Ahn}
\principaladviser{Tim Sheard}{\ }
\firstreader{James Hook}
\secondreader{Mark P. Jones}
\thirdreader{Andrew Tolmach}
\graduaterepresentative{Douglas V. Hall}
\departmenthead{Warren Harrison}
\grantdate{June}{30}{2014}

%%%% %% case (PAPER DISSERTATION)
%%%% \abstractpage
%%%%     Your abstract goes here...
%%%% \titlep
%%%% \prefatory
%%%% \dedication
%%%%     I dedicate this thesis to...
%%%% \prefacesection{Acknowledgments}
%%%%     I would like to thank...
%% case (ELECTRONIC DISSERTATION)
\titlep
\prefatory
\prefacesection{Abstract}
\vskip-5.5ex
$~~~~~~$Two major applications of lambda calculi in computer science are
functional programming languages and mechanized reasoning systems
(or, proof assistants). According to the Curry--Howard correspondence,
it is possible, in principle, to design a unified language based on
a typed lambda calculus for both logical reasoning and programming.
However, the different requirements of programming languages and
reasoning systems make it difficult to design such a unified language
useful for both purposes. Programming languages usually
extend lambda calculi with programming-friendly features
(\eg, recursive datatypes, general recursion) for supporting the flexibility to
model various computations, while sacrificing logical consistency.
Logical reasoning systems usually extend lambda calculi with
logic-friendly features (\eg, induction principles, dependent types)
for paradox-free inference over fine-grained properties,
while being more restrictive in modeling computations.

In this dissertation, we design and implement a language called Nax
that balances between the benefits of both. Nax accepts
all recursive datatypes, thus, allowing the same flexibility of defining
recursive datatypes as functional languages. Nax supports
a number of Mendler-style recursion schemes that can express various kinds of
recursive computations and also grantee termination. Nax supports
term-indexed types to support specifications of fine-grained properties.
In addition, Nax supports a conservative extension of
Hindley--Milner type inference.

The theoretical contributions of this dissertation include theories for
Mendler-style recursion schemes and term-indexed types, which we
developed to establish strong normalization and logical consistency of Nax.

\prefacesection{Dedication} 
To the Logos from the beginning of the universe, who was with the creator
and the creator himself, the source of all beings, and the light of life
shines in the darkness.

\prefacesection{Acknowledgments}
I would like to thank my advisor Tim Sheard, who has always been patient and
open to experimenting with new ideas, the members of the dissertation committee
for their useful feedback, and special thanks to Andrew M. Pitts and
Marcelo Fiore for the discussions greatetly improved my theiss research
during the visit to Cambridge, UK, and also to Gabor Grief, who gave feedback on
my thesis related research. Many thanks to my parents,
Rev. Sam Whan Kim, and friends in Myungsung Presbyterian Church for their
love and prayers from the other side of the Pacific; the young adults' group
in Oregon Eden Presbyterian Church for being a warm and dynamic community; and,
to coach Ruth Aquino and friends in the Lloyd Center Ice Rink, where I chilled
out in the evenings.
%%%% %% esac

\afterpreface
\body

%% BEGIN PART1
\part{Prelude}\label{part:Prelude}
\include{intro}   %% Introduction                          {ch:intro}
\include{poly}    %% Polymorphic type systems              {ch:poly}
%% END PART1

%% BEGIN PART2
\part{Mendler style}\label{part:Mendler}
\include{mendler} %% Mendler-style recursion schemes       {ch:mendler}
%% END PART2

%% BEGIN PART3
\part{Term-Indexed Lambda Calculi}\label{part:Calculi}
\include{fi}      %% System Fi                             {ch:fi}
\include{fixi}    %% System Fixi                           {ch:fixi}
%% END PART3

%% BEGIN PART4
\part{Nax Language}\label{part:Nax}
\include{nax}     %% The Nax language                      {ch:nax}
%%%% \include{casestd} %% Case studies                          {ch:casestd}
%% END PART4

%% BEGIN PART5
\part{Postlude}\label{part:Postlude}
\include{relwork} %% Related work                          {ch:relwork}
\include{futwork} %% Future work                           {ch:futwork}
\include{concl}   %% Conclusion                            {ch:concl}
%% END PART5

\clearpage
\phantomsection
\addcontentsline{toc}{part}{Appendix}
\appendix
\include{appendix}

\clearpage
\phantomsection
\addcontentsline{toc}{part}{Bibliography}
\bibliographystyle{plainnat}
\bibliography{main}

\clearpage
\phantomsection
\addcontentsline{toc}{part}{Index}
{\small
\begin{singlespace}
\printindex
\end{singlespace}
}



\end{document}


