\documentclass[11pt,twocolumn]{article}

\usepackage{amsfonts,euler}
\usepackage{MnSymbol}

\usepackage[all]{xy}
\UseComputerModernTips
\xyoption{knot}

\usepackage[a4paper,includehead,includefoot,headheight=13.6pt,
  headsep=4mm,top=1cm,bottom=1cm,left=2cm,right=2cm]{geometry}

\usepackage{float}
\floatstyle{ruled}
\restylefloat{figure}

\usepackage{dashrule}

\usepackage{multirow}

\usepackage{color}
\definecolor{grey}{gray}{.85}

\setcounter{secnumdepth}{4}
\renewcommand{\theparagraph}{\arabic{subsection}.\arabic{subsubsection}.\arabic{paragraph}}

%%% BEGIN macros
\newenvironment{myitemize}
  {\begin{list}{$\bullet$}
  {\setlength{\topsep}{1pt}
   \setlength{\partopsep}{1pt}
   \setlength{\itemsep}{0pt}
   \setlength{\parsep}{0pt}
   \setlength{\leftmargin}{1em}
   \setlength{\labelwidth}{.5em}}}
  {\end{list}}
\newenvironment{myindentitemize}
  {\begin{list}{$-$}
  {\setlength{\topsep}{2pt}
   \setlength{\partopsep}{2pt}
   \setlength{\itemsep}{2.5pt}
   \setlength{\parsep}{2.5pt}
   \setlength{\leftmargin}{2.125em}
   \setlength{\labelwidth}{1.625em}}}
  {\end{list}}
\newenvironment{myquote}
  {\begin{list}{}
  {\setlength{\topsep}{2pt}
   \setlength{\partopsep}{2pt}
   \setlength{\itemsep}{2.5pt}
   \setlength{\parsep}{2.5pt}
   \setlength{\rightmargin}{1em}
   \setlength{\leftmargin}{1em}
   \setlength{\labelwidth}{.5em}}}
  {\end{list}}
\newenvironment{mybigitemize}
  {\begin{list}{$\bullet$}
  {\setlength{\topsep}{2pt}
   \setlength{\partopsep}{2pt}
   \setlength{\itemsep}{2.5pt}
   \setlength{\parsep}{2.5pt}
   \setlength{\leftmargin}{1em}
   \setlength{\labelwidth}{.5em}}}
  {\end{list}}
\newenvironment{btritemize}
  {\begin{list}{\btr}
  {\setlength{\topsep}{2pt}
   \setlength{\partopsep}{2pt}
   \setlength{\itemsep}{2.5pt}
   \setlength{\parsep}{2.5pt}
   \setlength{\leftmargin}{1em}
   \setlength{\labelwidth}{.5em}}}
  {\end{list}}
\newcounter{CC}
\newenvironment{resenumerate}
  {\begin{list}{[\textbf{\arabic{CC}]}}
  {\usecounter{CC}
   \setlength{\topsep}{2pt}
   \setlength{\partopsep}{2pt}
   \setlength{\itemsep}{2.5pt}
   \setlength{\parsep}{2.5pt}
   \setlength{\leftmargin}{1.65em}
   \setlength{\labelwidth}{1.15em}
 }}
  {\end{list}}

\newcommand{\mysf}{\small\sf}
\newcommand{\mytextsf}[1]{\textsf{\small #1}}
\newcommand{\erc}{{\small\sf MaStrPLan}}
\newcommand{\ERC}{Mathematical Structures for Type Theories,\\[-.5mm] Logical
  Systems, and Programming Languages}

\newcommand{\hide}[1]{}
\newcommand{\hidewiring}[1]{}
\newcommand{\hideconservativity}[1]{}
\newcommand{\hideLId}[1]{}
\newcommand{\hideMathematicalUniverses}[1]{#1}
\newcommand{\note}[1]{\begin{quote}{\color{blue}$\leadsto$ \bf\em
      #1}\end{quote}}
\newcommand{\hidenote}{\hide}

\newcommand{\emgoal}[1]{#1}%{\em #1}
\newcommand{\bfem}[1]{\textbf{\em #1}}

\newcommand{\pref}[1]{\,(\ref{#1})}
\newcommand{\itemref}[1]{\textbf{[\ref{#1}]}}

\newcommand{\eg}{\emph{eg.}}
\newcommand{\Eg}{\emph{Eg.}}
\newcommand{\vs}{\emph{vs.}}
\newcommand{\ie}{\emph{ie.}}
\newcommand{\viz}{\emph{viz.}}
\newcommand{\etc}{\emph{etc.}}
\newcommand{\etal}{\emph{et al.}}
\newcommand{\cf}{\emph{cf.}}

\newcommand{\lcalculus}{\mbox{$\lambda$-calculus}}
\newcommand{\SysL}{$L$}%{$\boldsymbol L$}
\newcommand{\SystemF}{\mbox{System~$F$}}
\newcommand{\SystemFi}{\mbox{System~$F_i$}}
\newcommand{\SystemFomega}{\mbox{System~$F_\omega$}}
\newcommand{\LC}{\mbox{$LC$}}

\newcommand{\btr}{$\blacktriangleright$}

\newcommand{\reqpsize}{8.113395cm}%{\columnwidth}

\newcommand{\req}[2]{\begin{center}\colorbox{grey}{\begin{minipage}{\reqpsize} 
  \mytextsf{Research question}\hfill$^{\mbox{\scriptsize see #1 }}$\\[-5.5mm]
  \begin{btritemize}
  \item #2
  \end{btritemize}
\end{minipage}}\end{center}}

\newcommand{\reqs}[2]{\begin{center}\colorbox{grey}{\begin{minipage}{\reqpsize}
  \mytextsf{Research questions}\hfill$^{\mbox{\scriptsize see #1 }}$\\[-5.5mm]
  \begin{btritemize}
  \item #2
  \end{btritemize}
\end{minipage}}\end{center}}

\newcommand{\rep}[2]{\begin{center}\colorbox{grey}{\begin{minipage}{\reqpsize}
  \mytextsf{Research pathway}\hfill$^{\mbox{\scriptsize see #1 }}$\\[-5.5mm]
  \begin{btritemize}
  \item #2
  \end{btritemize}
\end{minipage}}\end{center}}

\newcommand{\Set}{{\boldsymbol{\mathscr S}}}
\newcommand{\scat}[1]{\mathbb{#1}}
\newcommand{\cat}[1]{\mathscr{#1}}
\newcommand{\op}{\circ}
\newcommand{\Id}{\mathrm{Id}}
\newcommand{\Di}{\mathrm{Di}}

\newcommand{\qed}{\Box}
\newcommand{\logo}
{\mbox{\tiny$\xymatrix@C=3pt@R=4.5pt{
    & \qed \ar@{-}[d] 
    \ar@/_.5em/@{-}[ddl]<-.25em>
    \ar@/^.5em/@{-}[ddr]<.25em>
    & 
    \\
    & \ar@{-}[dl] \qed \ar@{-}[dr] & 
    \\
    \ar@/_.5em/@{-}[rr]<.125em>
    \qed & & \qed
  }$}}
%%% END macros

\usepackage{fancyhdr}
\pagestyle{fancy}
\lhead{\emph{Marcelo Fiore}}
\chead{Part~B1}
\rhead{\erc}
\lfoot{}
\cfoot{\thepage}
\rfoot{}

\usepackage[compact]{titlesec}

\usepackage{compactbib}

\usepackage{setspace}
\setstretch{.9}

\usepackage{times}

\makeatletter
\renewcommand\@biblabel[1]{#1}
\makeatother

\renewcommand{\thesection}{\arabic{section}}
\renewcommand{\thesubsection}{\arabic{subsection}}

\def\contentsname{\large Contents\\[-7.5mm]}
\usepackage{minitoc}
\tightmtctrue

\usepackage[draft]{hyperref}
\hypersetup{pdftitle={\ERC},pdfauthor={Marcelo Fiore}}

\newcommand{\obullet}{\mbox{$\raisebox{-.85mm}{\huge$\circ$}\hspace*{-2.6mm}\bullet$}}

\renewcommand{\ostar}{\mbox{$\raisebox{-.85mm}{\huge$\circ$}\hspace*{-2.75mm}\star$}}

\setlength{\intextsep}{5pt}

\begin{document}
\setlength{\abovedisplayskip}{5pt}
\setlength{\belowdisplayskip}{5pt}

\setcounter{page}{0}
\thispagestyle{plain}
\twocolumn[\begin{@twocolumnfalse}
{\large European Research Council}

\vspace*{7.5mm}
\begin{center}\Large
ERC Advanced Grant\\
Research Proposal\\
(Part B1)
\end{center}

\vspace*{2.5mm}

\begin{center}\bfseries\LARGE
Mathematical Structures for Type Theories,\\
Logical Systems, and Programming Languages\\
\vspace*{10mm}
\Large\bfseries\sf 
MaStrPLan\\
  \logo
\end{center}

\vspace*{2.5mm}

\begin{center}\large
Principal Investigator: Marcelo Fiore\\
\vspace*{1mm}
Host Institution: University of Cambridge\\
\vspace*{1mm}
Proposal duration: 60 months
\end{center}

\vspace*{5mm}

\renewcommand{\abstractname}{\bfseries\large Summary}
\begin{abstract}\normalsize
Headed by Principal Investigator Marcelo Fiore, {\erc} assembles an
international team of world-leading experts in theoretical and applied
computer science.  The project builds a group of researchers
dedicated to combined research in Category Theory, Mathematical Logic,
Programming Theory, and Type Theory.  Specifically, to understand and develop
formal languages and mathematical models for computer programming and
mathematical proof. 

{\color{red}FINISH}

We aim to produce several artifacts: a framework to analyse and synthesise
type theories; mathematical theories and models for computational phenomena;
formal calculi for proof and/or computation; and experimental high-level
programming languages.
\end{abstract}

\end{@twocolumnfalse}]

\clearpage
\chead{Part\,B1(a)}
\twocolumn[\begin{@twocolumnfalse}
	\hfill{\bfseries\Large 
      Extended Synopsis of the scientific proposal
  }\hfill\null\\
\end{@twocolumnfalse}]

\subsection{Proposal}
\label{Proposal}

\paragraph*{General aim.}

Headed by Principal Investigator Marcelo Fiore, {\erc} assembles an
international team of world-leading experts in theoretical and applied
computer science.  The project builds a group of researchers
dedicated to combined research in Category Theory, Mathematical Logic,
Programming Theory, and Type Theory.  Specifically, to understand and develop
formal languages and mathematical models for computer programming and
mathematical proof. 

\paragraph*{Working thesis.}

The research is driven by the thesis that
\begin{myquote}
\item
Languages 
for computer programming
and languages 
for mathematical proof 
are of the same character.
\end{myquote}
This view %conception 
%%is of course not original to us.  It 
was first developed in the 1960s %, with proponents from computer science
%(\eg~McCarthy, Dijkstra, Hoare, Strachey) and from mathematics (\eg~de Bruijn,
%Scott, Martin-L\"of); 
and is now central to research in programming languages and constructive
mathematics.

An aspect of the thesis that is fundamental to us here is usually referred
to as the Proofs-as-Programs or Propositions-as-Types correspondence.
%There is an aspect of the thesis that is rooted in the constructive
%approach to mathematical practice and is fundamental to us here.  This 
It can be impressionistically %roughly 
presented %as a correspondence 
as follows
\[%\begin{equation}\label{PATproportion}%\xymatrix@C5pt{
%  \frac{\txt{program}}{\txt{Type}} 
%  & \mbox{\Large$\approx$} & 
%  \frac{\txt{proof}}{\txt{Proposition}}
%}
%
  \mbox{proof : Proposition} 
  \enspace \approx \enspace 
  \mbox{program : Type} 
%\end{equation}
\]
One %and one
intuitively explains the correspondence along these lines: %the lines below:
%\begin{myitemize} 
%\item 
  to give a constructive proof of a proposition 
  %(\eg~that every number is smaller than a prime number) 
  is to construct a program that certifies the statement, 
  %(\eg~one that on inputing a number, outputs some prime number bigger
  %than it);
  while
%\item 
  to type a program is to establish a property of that program.
  %(\eg~that certain input/output behaviour or invariant is maintained). 
%\end{myitemize} 

Research stemming from the Propositions-as-Types correspondence has been
very fruitful.  Many of its mathematical theories have been implemented as
computer systems, and used academically and industrially.  Cases in point
are: programming languages for high-assurance code and proof assistants
for computer-aided verification; examples of which are
Haskell, %~\cite{Haskell} 
an industrial-strength functional programming language, 
and Coq, %~\cite{Coq} 
a proof assistant in which bodies of mathematics are being verified.

\paragraph*{The future.}

Transferring mathematical theories to computer
systems has required both technical/theoretical and technological/applied
progress.  This has necessarily demanded increasingly specialised, and to
some extent fragmentary, research.  
Importantly, %though, 
the field has reached a first plateau; encompassing a substantial body of
work and expertise.  
How %But, how
will it continue to advance?  
By considering new directions and challenges, of course.  
But, 
it is our firm view that %, as in the early days of the subject, 
it will be essential to do so generating questions and tackling problems
from the perspectives of a variety of research areas.  
It is both of these aspects that we put forward here.

\paragraph*{Our philosophy.}

Two disciplines are clearly involved in investigating the
Proofs-as-Programs 
correspondence: %~(\ref{PATproportion}) above:
Mathematical Logic and Programming Theory.  
%Much research, as the one proposed here, is geared towards finding a middle
%ground that can accommodate the concerns of both, and expand their 
%realms. %domains.
%
This %However, this 
is only half of the picture.  The full 
picture also involves the disciplines of Type Theory and Category
Theory.  
%Their role is to enrich~(\ref{PATproportion}) as follows 
%\[\xymatrix@C5pt@R2.5pt{
%  & \frac{\txt{\small construction}}{\txt{\small Type}} 
%    \ar@{}[rd]|-{\rotatebox[origin=c]{-45}{\Large$\approx$}}
%  & 
%    \\
%    \frac{\txt{\small program}}{\txt{\small Type}} 
%    \ar@{}[ru]|-{\rotatebox[origin=c]{-135}{\Large$\approx$}}
%    & & 
%    \frac{\txt{\small proof}}{\txt{\small Proposition}}
%  \\
%  & \frac{\txt{\small morphism}}{\txt{\small Object}} 
%  \ar@{}[ru]|-{\rotatebox[origin=c]{-135}{\Large$\approx$}}
%  & 
%}\]
Figure\,\ref{ResearchAreas} 
\begin{figure}[h]
\caption{Research areas and interactions.}
\vspace*{2mm}
\begin{center}
\hspace*{.5mm}
\xymatrix@R=25pt@C=15pt{
& 
\raisebox{7mm}{\fbox{\txt{\small Mathematical\\\small Logic}}}
\ar@/_1em/@{<->}[ddl]<-1em>|-
  {\txt{\scriptsize Propositions\\ \raisebox{1mm}{\scriptsize as Types}}}
\ar@/^1em/@{<->}[ddr]<1em>|-
  {\txt{\scriptsize Proofs as\\\raisebox{1mm}{\scriptsize Programs}}} 
& 
\\
& 
\ar@{<->}[dl]|-
  {\txt{\scriptsize Internal\\\raisebox{1mm}{\scriptsize Languages}}}
\ar@{<->}[dr]|-
  {\txt{\scriptsize Denotational\\\raisebox{1mm}{\scriptsize Semantics}}} 
\ar@{<->}[u]|-
{\txt{\scriptsize Mathematical\\\raisebox{1mm}{\scriptsize Models}}}
\fbox{\txt{\small Category\\\small Theory}}
& 
\\
  \fbox{\txt{\small Type\\\quad\small Theory\quad\null}}
\ar@/_1em/@{<->}[rr]|-
  {\txt{\scriptsize Formal\\\raisebox{1mm}{\scriptsize Languages}}}
& & 
\fbox{\txt{\small Programming\\\small Theory}}
}
\end{center}
\vspace*{-2mm}
\label{ResearchAreas}
\end{figure}
gives a schematic view of the interactions between these four research areas
in this context.  Each of them is complementary to the others.  Together, as a
unified whole, they have shaped fields of computer science and mathematics.  
\hide{
Indeed, consider for instance that: 
\begin{myitemize}
\item
  the categorical interpretation of quantifiers as
  adjoints~\cite{LawvereAinF} informed the development of the
  type-theoretic dependent sums and dependent products~\cite{ScottCV};
\item
  type theories are the foundation of programming-language typing
  systems~\cite{Pierce};
\item
  the control operators of programming languages are key to the
  constructive interpretation of classical proofs~\cite{Griffin}; and 
\item
  model-theoretic studies of the polymorphic lambda
  calculus~\cite{GirardSystemF,Reynolds} led to remarkable small complete
  categories~\cite{Hyland}.
\end{myitemize}
}
%
It is within this framework that we propose cross-cutting research in Category
Theory, Mathematical Logic, Programming Theory, and Type Theory.  
We %Furthermore, we 
contend that an approach neglecting any one of them is to the detriment of
the others; missing the depth and richness of the subject and, crucially,
missing opportunities for research and development.  

\paragraph*{Main goals.}

The motivations for our research proposal are specifically as follows. 
\begin{myitemize}
\item[\raisebox{.75mm}{\tiny$\bigstar$}]\hspace*{-2mm}
  To search for %elucidate 
  %what we hope to be 
  the next-generation framework of Type Theory
  relevant to computer science and mathematics.
\item[\raisebox{.75mm}{\tiny$\bigstar$}]\hspace*{-2mm}
  To exercise and test Category Theory both as a unifying and as a
  foundational mathematical language.
\item[\raisebox{.75mm}{\tiny$\bigstar$}]\hspace*{-2mm}
  To broaden the current role of Mathematical Logic in computation.
\item[\raisebox{.75mm}{\tiny$\bigstar$}]\hspace*{-2mm}
  %To develop Programming Theory, designing, implementing, and experimenting
  %with languages of the future.
  To design, implement, and experiment within Programming Theory, looking
  into the languages of the future.
\end{myitemize}

We particularly aim to contribute with: a framework to analyse and synthesise
type theories; mathematical theories and models for computational phenomena;
formal calculi for proof and/or computation; and experimental high-level
programming languages.

\paragraph*{Research programme.}

The research programme has been precisely conceived to target the goals
above.  It is organised in four strands as outlined below.
\begin{myitemize}
\item[{\bfseries 1\enspace Foundations:}]\mbox{}\enspace\thinspace 
  \emph{A comprehensive research programme on the metamathematics of type
    theories.}
  %--- addressing the problem of what type theories are.

  \vspace*{1mm}
  We will address the fundamental question of what type theories are,
  which will also lead us to rethink them.  
  Our approach 
  %is novel and ambitious.  It 
  aims at an algebraic framework that will generalise to type theories all
  aspects of our current understanding of algebraic
  theories. %, from the perspectives of categorical algebra,
  %equational logic, and universal algebra. 
  %%(see the Algebraic Trinity of Figure~\ref{} on page~\pageref{}).  
  The scale at which we will be attempting this is unprecedented,
  systematically exploring a wide spectrum of key type-theoretic features.
  \vspace*{1mm}

\item[{\bfseries 2\enspace Models:}]\mbox{}\enspace\thinspace
  \emph{Study of mathematical models for type theories and logical systems.}
  %--- targeting semantic problems at the forefront of current understanding. 

  \vspace*{1mm}
  We will tackle semantic problems at the forefront of current understanding.
  %
  In the context of type theories, we will conduct investigations
  concerning a main problem in the area: to reconcile extensional equality
  and intensional identity (which, roughly, respectively correspond to the
  mathematical and computational notions of sameness).  
  %
  In the context of logical systems, we will develop both %novel 
  calculi suggested by mathematical models and %new 
  models for so far intractable calculi, encompassing aspects of resource
  management and computational effects.  
  %A distinguishing novelty of our approach is that it will be informed by the
  %logical notion of polarisation.
  \vspace*{-3mm}%%%HACK!!!
  
\item[{\bfseries 3\enspace Calculi:}]\mbox{}\enspace\thinspace
  \emph{Development of formalisms of deduction as internal languages of
    mathematical models.}
  %--- researching type theories that go beyond current logical frameworks. 
  
  \vspace*{1mm}
  We will aim at evolving the term and type structure of current type
  theories, which in essence has not changed since the late 1960s.  Pursuing
  the view that further evolution will come from mathematical input, we will
  research type theories as formal languages of (higher-dimensional)
  categorical structures.  
  \vspace*{1mm}

\item[{\bfseries 4\enspace Programming:}]\mbox{}\enspace\thinspace
  \emph{Design and implementation of novel computational languages.}
  %--- %looking into the 
  %engineering %of first 
  %principles and concepts from mathematical theories.

  \vspace*{1mm}
  Guided by mathematical theories and pragmatics, we will look into the
  engineering of principles and concepts for programming, exporting them to
  language designs and implementations.  We will specifically consider
  experimental languages, supporting indexed data structures, computational
  effects, and metaprogramming.  
\end{myitemize}

\paragraph*{The team.}

The team to pursue the research programme will be led by 
%\begin{myitemize}
%  \item 
    Principal Investigator~(PI) \emph{Marcelo Fiore} 
    (Computer Laboratory, University of Cambridge): 
    a %world-
    leading expert in applications of category theory to computer science,
    including abstract algebra, concurrency theory, programming-language
    semantics, sheaf theory, and type theory.  
%\end{myitemize}
%
He will be 
backed up %supported 
by an international group of three Senior Visiting Researchers~(SVRs):
%(that are to visit the PI for one month per year): 
%\begin{myitemize}
%\item 
  \emph{Pierre-Louis Curien} 
  (Laboratoire PPS, Univerist\'e Paris Diderot - Paris~7): 
  a %world-
  leading expert in programming-language semantics, with contributions
  into category theory, logic, %rewriting theory, 
  and type theory; 
%\item
  \emph{Peter Dybjer} 
  (Department of Computer Science, Chalmers University of Technology): 
  a %world-
  leading expert in type theory, including its connections with category
  theory and logic, and its implementation in proof assistants; and 
%\item
  \emph{Tim Sheard} 
  (Department of Computer Science, Portland State University): 
  a %world-
  leading expert in pro\-gram\-ming-language design and implementation,
  encompassing functional and metaprogramming systems. 
%\end{myitemize}
%
The PI is to collaborate with the SVRs by mutual research visits, email, and
conference calls.
%
The team is completed by three outstanding Research Associates~(RAs) to be
based at the Computer Laboratory, University of Cambridge, under the PI.  They
are
%\begin{myitemize}
%\item
  \emph{Ki Yung Ahn}
  (Department of Computer Science, Portland State University): 
  a PhD student of %Tim 
  Sheard currently writing up his dissertation on the design and
  implementation of a language for indexed programming;
%\item
  \emph{Nicola Gambino}
  (Dipartimento di Matematica e Informatica, Universit\`a degli Studi di
  Palermo):  
  an established researcher in the areas of type theory and category theory,
  with whom the PI has already collaborated; and 
%\item
  \emph{Guillaume Munch-Maccagnoni}
  (Laboratoire PPS, Univerist\'e Paris Diderot - Paris~7): 
  a PhD student of %Pierre-Louis 
  Curien writing up his dissertation on the Propositions-as-Types
  correspondence in the context of classical logic.
%\end{myitemize}
Currently, the two junior RAs lack future funding.  All RAs would be ready
to join the project from the start.

\hide{
Figure\,\ref{ercTeam} gives a schematic presentation of the team that matches
the research areas as rendered in Figure\,\ref{ResearchAreas}.  
\begin{figure}[h]
\caption{{\erc} team}
\vspace*{2mm}
\begin{center}
\hspace*{.125mm}
\xymatrix@R=10pt@C=32.5pt{
& 
\raisebox{5mm}
  {\fbox{\txt{\small Pierre-Louis\\ \small Curien}}}
\ar@/_1em/@{<->}[dddl]<-1em>
\ar@/^1em/@{<->}[dddr]<1em> 
& 
\\
\\
& 
  {\fbox{\txt{\small Marcelo\\ \small Fiore}}}
\ar@{<->}[dl]|-
  {\txt{\scriptsize Nicola\\ \raisebox{1mm}{\scriptsize Gambino}}}
\ar@{<->}[dr]|-
  {\txt{\scriptsize Ki Yung\\ \raisebox{1mm}{\scriptsize Ahn}}}
\ar@{<->}[uu]|(.475)
  {\txt{\scriptsize Guillaume\\ \raisebox{1mm}{\scriptsize
        Munch-Maccagnoni}}}
& 
\\
\raisebox{0mm}{\fbox{\txt{\small Peter\\ \small\ Dybjer\ \ }}}
\ar@/_1em/@{<->}[rr]
& & 
\raisebox{0mm}{\fbox{\txt{\small Tim\\ \small\ Sheard\ \ }}}
}
\end{center}
\vspace*{-2mm}
\label{ercTeam}
\end{figure}
%Details %Further details 
%are deferred to Section\pref{ResourcesSection}. 
}

\subsection{Origins and influences}
\label{Origins}

We sketch %This section sketches 
the origins and influences that led to the holistic conception of
Figure\,\ref{ResearchAreas}. 
%, roughly expanding through 1935 to 1985.
It is important to appreciate and understand our scientific framework.

\hide{
Section\pref{StateOfTheArtSubsection} describes the current state of the
field and raises questions for research.  %Having set the scene,
Section\pref{ObjectivesSubsection} presents the research objectives we
wish to pursue and the team that we have assembled to reach them.  
%Both of these points are further expounded upon in
%Sections\,(\ref{MethodologySection}\,\&\,\ref{ResourcesSection}).
}

\paragraph*{Logic and computation.}

%It is worth recalling that 
Computer science was born as a branch of mathematics, specifically
mathematical logic. %, even before the first electronic computers were built.
Its inception %The inception of computer science 
was Hilbert's %~\cite{?}
%1928 
Decision Problem, %Entscheidungsproblem (decision problem),
asking whether there is an algorithmic procedure for deciding mathematical
statements.  Negative answers were provided independently by
Church~\cite{Church1936} 
%in 1936 
and 
Turing~\cite{Turing},
%in 1937,
giving birth to the mathematical theory of computation.  Their 
%completely different 
approaches gave rise to different branches of theoretical
computer science. %that still persist today.  
The line of development starting with Church's {\lcalculus} %~\cite{?} 
is concerned with prototypical computational languages that are
used to study high-level programming languages.  
%On the other hand, Turing's machines %~\cite{?} 
%are the most widely 
%used model for analysing computational complexity.
%
This %Church's 
view is central to this proposal.  
\hide{
The {\lcalculus} is a
deceptively simple formal system.  Its syntax consists of three types of
phrases: variables, applications, and abstraction.  Church's seminal
innovation was in introducing the latter one, 
%the other two being already familiar form the realm of
%algebra~\cite{Birkhoff}, 
which in modern terminology is referred to as a binding operator, a notion
that goes beyond the operators of universal algebra. %~\cite{Birkhoff}.
Binding operators are an integral part of all high-level programming
languages.  
%Their mathematical theory is subtle %, see~\ref{}, 
%because they define syntax up to the consistent renaming of bound
%variables (technically referred to as \mbox{$\alpha$-equivalence}).
%This subtlety is made evident by the anecdotal fact that the first definitions
%of substitution were formally flawed.  
%%These important topics are core to our proposed research in
%%Section\pref{AlgebraicTypeTheoryParagraph}.
}

%Incidentally, the notion of variable binding was initially introduced much
%earlier, by Frege~\cite{Frege1879}, in the context of mathematical logic.  
%He used binders to define a formal symbolic
%system axiomatising not only the propositional connectives of Boolean
%logic~\cite{Boole} but also, for the first time, the quantifiers of
%predicate logic.

%The %In modern notation, the 
%{\lcalculus} syntax is given by
%  \[\begin{array}{rcll}
%    s , t & ::= & & \mbox{($\lambda$-terms)}\\
%      & \mid & x & \quad\mbox{(variables)}\\
%      & \mid & (t)s & \quad\mbox{(application)}\\
%      & \mid & \lambda x.\,t & \quad\mbox{(abstraction)}
%  \end{array}\]
%The system has only one rule of computation:
%  \begin{equation}\label{BetaReduction}\begin{array}{rl}
%(\beta) & (\lambda x.\,t)s \,\longrightarrow\, t[s/x]
%  \end{array}\end{equation}
%by which the result of computing a so-called redex $(\lambda x.\,t)s$ is
%the \mbox{$\lambda$-term} $t[s/x]$ denoting the result of substituting the
%free occurrences of $x$ in $t$ by $s$.  Non-terminating behaviour arises from
%self application.

\paragraph*{Type theory and logic.}
\label{SectionATypeTheoryAndLogicParagraph}

\hide{
The concept of `type' was conceived to solve a foundational
problem.  In 1903, %~\cite{Frege1903}, building on~\cite{Frege1879}, 
Frege proposed
a logical system as a foundation for mathematics.
%including arithmetic.  
In his now famous paradox, 
Russell %~\cite{Russell1902}
observed that one of Frege's axioms led to inconsistency.  
%He was led to this contradiction by related contradictions he found while
%developing his account of set theory.  
To overcome such foundational problems, 
he %Russell 
introduced the `doctrine of types'. 
% --- typed systems can avoid these inconsistencies.
}

Type Theory, as we now know it, 
arose from the integration by Church
of types into his \lcalculus, yielding the %so-called 
Simply-Typed Lambda Calculus. %~\cite{Church1940}.  
This was a natural step, particularly if one held the naive interpretation of
\mbox{$\lambda$-abstraction} as defining functions.  
%The type theory known as the Simply-Typed Lambda Calculus has a set of types
%consisting of basic ones
%closed under a function-type constructor. %:
%\[\begin{array}{rcll}
%    \sigma,\tau & ::= & & \mbox{(simple types)}\\
%    & \mid & \theta & \quad\mbox{(basic types)}\\
%    & \mid & \sigma\to\tau & \quad\mbox{(function types)}
%  \end{array}\]
%In the Simply-Typed Lambda
%Calculus, \mbox{$\lambda$-terms}~$t$ are classified by types~$\tau$, in
%contexts $\Gamma$ (assigning types to variables), for which the notation
%$\Gamma\vdash t:\tau$
%is commonly used.  This is done according to syntax-directed rules. 
% that in the mathematical vernacular are presented as follows
%  \[
%  \begin{array}{c}
%    \\ \hline
%    x_1:\tau_1,\ldots,x_n:\tau_n\vdash x_i:\tau_i
%  \end{array}
%  \enspace(1\leq i\leq n)
%\]
%  \[\begin{array}{c}
%    \Gamma\vdash t:\sigma\to\tau
%    \quad 
%    \Gamma\vdash s:\sigma
%    \\ \hline
%    \Gamma\vdash (t)s:\tau
%\end{array}\]
%\vspace*{-2mm}
%  \begin{equation}\label{LambdaAbstraction} 
%  \begin{array}{c}
%    \Gamma,x:\sigma\vdash t:\tau
%    \\ \hline
%    \Gamma\vdash\lambda x.\,t:\sigma\to\tau
%  \end{array}
%\end{equation}
%\vspace*{-2mm}
%
In this context, a fundamental discovery relating Mathematical Logic to Type
Theory was made %by Curry and 
by Howard~\cite{Howard1969}.
%in two slightly different, though related, logical
%contexts (technically, Hilbert-style deduction %~\cite{Hilbert} 
%and Gentzen's natural deduction in sequent form). %~\cite{Gentzen1935}).
This is the Propositions-as-Types correspondence, %; roughly establishing a
%correspondence between terms having types and proofs proving propositions.
referred to in Section\,\ref{Proposal}.
%(\cf~(\ref{PATproportion})).  
%For the Simply-Typed Lambda Calculus, the correspondence
%can be illustrated by noting that the erasure of term information in 
%the typing rules (given above) 
%yields the deduction rules
%\[\begin{array}{c}
%  \begin{array}{c}
%    \\ \hline
%    \tau_1,\ldots,\tau_n\vdash \tau_i
%  \end{array}
%  \enspace(1\leq i\leq n)
%  \\[4mm]
%  \begin{array}{c}
%    \Gamma\vdash \sigma\to\tau
%    \quad 
%    \Gamma\vdash\sigma 
%    \\ \hline
%    \Gamma\vdash\tau
%  \end{array}
%  \enspace\quad
%  \begin{array}{c}
%    \Gamma,\sigma\vdash\tau
%    \\ \hline
%    \Gamma\vdash\sigma\to\tau
%  \end{array}
%\end{array}
%\]
%derivable judgements $x_1:\tau_1,\ldots,x_n:\tau_n\vdash t:\tau$ yields the
%valid judgements $\tau_1,\ldots,\tau_n\vdash\tau$ 
%of Intuitionistic %Minimal 
%Logic. %~\cite{?}.

In another very important direction Church introduced a Simple Theory of
Types~\cite{Church1940} 
as an axiomatization of Higher-Order Logic.
This axiomatization was formalised within a Simply-Typed Lambda Calculus.
%with basic types for both individuals and propositions, further enriched with
%constants for the logical connectives.  
In doing so, he adopted a radically new perspective, shifting the status
of the Simply-Typed Lambda Calculus from that of a `language' to a
`metalanguage'. %;~\ie, a language in which it is possible to represent and
%work with other languages. 
%In this particular case for simple type theories.  
\hide{
  When regarded as a metalanguage, the Simply-Typed
Lambda Calculus is considered as an equational theory. %, with the
%computational \mbox{$\beta$-rule}~(\ref{BetaReduction}) stated as an
%equation together with the extensionality equation
%  \[\begin{array}{rll}
%(\eta) & \lambda x.\,(t)x = t 
%& \mbox{, where $x$ is not free in $t$}
%  \end{array}\]
%\vspace*{-4mm}
}

\hide{
The processes of abstracting from languages to metalanguages has become a
common activity in computer science, and plays an important role in our
proposed investigations.  
}

\paragraph*{Category theory, logic, and type theory.}

\hide{
The theory of categories was introduced by Eilenberg and Mac
Lane~\cite{EilenbergMacLane}.  A category is a mathematical structure
consisting of objects and morphisms.  Morphisms are classified
by pairs of objects. The notation $f:A\to B$ stipulates that $A$ and $B$ are
objects, and that $f$ is a morphism with domain $A$ and codomain~$B$.  
Categories come equipped with an associative law that composes pairs of
morphisms 
%as on the left below
%\begin{equation}\label{Category}
%  \begin{array}{c}
%  f:A\to B\quad g:B\to C
%  \\ \hline
%  g\,f:A\to C
%\end{array}
%\quad\qquad
%1_A:A\to A
%\end{equation}
together with, for every object, an identity morphism 
%as on the right above
that is a neutral element for composition.  
%Categories were defined to introduce functors, a notion of morphism between
%categories, which in turn was defined to introduce natural transformations, a
%notion of morphism between functors.

In understanding categories, it is helpful to have in mind
that they have two kinds of uses: in the
large and in the small.  In the large, categories are seen as mathematical
universes of discourse (within which mathematical constructions take place,
typically by means of universal properties); like the categories of: sets and
functions, algebraic structures and homomorphisms, spaces and continuous
functions.  In the small, categories are seen as mathematical objects
themselves; like a set, a monoid, a preorder, and a graph, all of which can
be suitably regarded as a category.
}

\hide{
Category Theory analyses mathematical structure by isolating the principles
for which mathematical theories work.  
}
%Because of this attention to essentials, it has had considerable success
%in unifying ideas from many areas of mathematics. 
%Today it an indispensable tool in abstract algebra, algebraic geometry,
%mathematical logic, mathematical physics, topology, and theoretical computer
%science, as well as a growing research area in its own right.

The connection between Category Theory, Logic, and Type Theory was
initiated by Lawvere %~\cite{LawvereAinF} 
and Lambek. %~\cite{LambekI}, 
%both
%of whom recognised logical systems and type theories as categories with
%structure.  
%Here we are concerned with the former, postponing the latter to the following
%section.
%
Lawvere's insight was to understand logical connectives and type
constructors as categorical structures arising from 
%universal properties in the form of 
adjoint functors. %, a notion introduced by Kan~\cite{Kan}
%motivated by homology theory.  
%For instance, the categories with structure corresponding to the Simply-Typed
%Lambda Calculus with products are Cartesian Closed Categories.  These have
%categorical products and exponentials 
%, respectively denoted $\times$ and $\Rightarrow$, 
%defined by adjoint situations.
%establishing natural bijective correspondences between
%morphisms as follows:
%  \[
%  \begin{array}{c}
%    C\to A \enspace,\enspace C\to B
%    \\ \hline\hline
%    C\to (A\times B)
%  \end{array}
%  \quad\qquad
%  \begin{array}{c}
%    (C\times A)\to B
%    \\ \hline\hline
%    C\to (A\Rightarrow B)
%  \end{array}
%\]
%The required naturality condition amounts to the computational $\beta$
%laws and the extensionality $\eta$ laws.
%
It
%The similarity between the bijective correspondence on the right above and the
%typing rule for \mbox{$\lambda$-abstraction}~(\ref{LambdaAbstraction}) is not
%casual, and it 
is now well-understood that the Simply-Typed Lambda Calculus with products
provides an internal language for Cartesian Closed Categories: %; namely, 
it is the calculus of all such models.  This view further enriches the
Propositions-as-Types correspondence as 
%in the trinity below %in Figure\,\ref{PAT}.
%\begin{figure}[h]
%\caption{Propositions-as-Types Trinity}
%\vspace*{2mm}
%\[
%\begin{equation}\label{PATtrinity}
%\hspace*{-3.75mm}
%\begin{array}{c}
%\xymatrix@C=0pt@R=17.5pt{
%    & \ar@{<->}[dl]_(.5){\null\hspace*{-15mm}\txt{\scriptsize Internal\\
%        \raisebox{1mm}{\scriptsize Language}}} 
%    \ar@{<->}[dr]^(.55){\null\hspace*{1mm}\txt{\scriptsize Mathematical\\
%        \raisebox{1mm}{\scriptsize Models}}}
%    \txt{\small Cartesian Closed\\ \small Categories} & \\
%    \txt{\small Simply-Typed\\ \small Lambda Calculus\\ \small with products
%    }
%    \ar@{<->}[rr]_-{\txt{\scriptsize
%        Propositions\\\raisebox{1mm}{\scriptsize as Types}}} & & 
%    \txt{Propositional\\ 
%      Intuitionistic\\ Logic}
%  }
%\end{array}
%\vspace*{-1mm}
%\end{equation}
%\]
%\vspace*{-2mm}
%\label{PAT}
%\end{figure}
%and is the standard with respect to which analogous developments in the
%area are measured.  
into a trinity consisting of Simply-Typed Lambda Calculus with products,
Cartesian Closed Categories, and Propositional Intuitionistic Logic.

\paragraph*{Type theory and programming.}

The Simply-Typed Lambda Calculus 
%, regarded as a language rather than as a metalanguage, 
is a prototypical functional programming language.  
%The change of perspective from Type Theory to Programming Theory is not
%straightforward and comes with considerations that enrich both subjects.  
%
In %the context of programming languages, 
this light, %Milner %~\cite{Milner1978}
%understood early on that while a programming language should come with a
%type discipline, %to classify programs according to their type invariants,
%programmers would be better served if type annotations were inferred.
%automatically.  
%
the %The 
problem of type inference 
%(by which given a program one wishes to compute the
%best possible type for it) is %became 
of central practical importance.  
%For
%Combinatory Logic %\footnote{Combinatory Logic is an important symbolic
%  formalism introduced by Sch\"onfinkel~\cite{Schonfinkel} closely related
%  to the {\lcalculus}, but based on algebraic combinators, that enjoys the
%  Propositions-as-Types correspondence with respect to the Hilbert-style
%  deduction system for Intuitionistic %Minimal 
%  Logic.} 
%this problem was solved by Hindley. %~\cite{Hindley1969}.  
%Independently, 
Milner~\cite{Milner1978} 
solved it %in a context more relevant to programming; namely, 
for the Simply-Typed Lambda Calculus with parametric polymorphism.  
%The algorithm is known as the Hindley-Milner type inference method.  
%The approach is very robust. 
%, extending broadly to many related systems.
 
Polymorphism in programming %refers to languages that support abstraction
%mechanisms by which a program %(function or procedure) 
%can be used with a variety of types.  The concept 
was introduced by Strachey~\cite{Strachey1967}.
%who further classified the phenomena into ad-hoc polymorphism and parametric
%polymorphism.  The former is also referred to as overloading and accounts for
%uses of a program with different types (like integers and reals) by means of
%different algorithms.  The latter indicates the use of a uniform program for
%all types (like a duplicator program making copies of its input).
%
Reynolds~\cite{Reynolds} 
formalised 
it %the programming intuition 
by introducing the Polymorphic Typed Lambda Calculus. 
%; an extension of the Simply-Typed Lambda
%Calculus with polymorphic types.  
%Roughly, these are abstract parametric types
%whose programs can be used for all instances of the parameter.  
Strikingly,
this system 
had already been proposed %several years earlier 
by
Girard~\cite{GirardSystemF} 
%under the name of {\SystemF}, 
as the
type-theoretic counterpart of Second-Order Propositional Logic. 
%in the context of the Propositions-as-Types correspondence.  
%Further, 
Girard had
also introduced {\SystemFomega}, the type-theoretic counterpart of
Higher-Order Propositional Logic.  
%These logical systems widely extend the
%expressiveness of the Simply-Typed Lambda Calculus, notably by the
%possibility of encoding inductive data types. %~\cite{BoehmBerarducci}.
{\SystemFomega} lies at the core of the Haskell programming
language. %~\cite{EqProofICFP12}.

These %The above 
type systems aim at %providing 
logical foundations and %, when viewed
%as programming languages, %they can 
only introduce terminating computations.  
As for
%There are various ways in which one can extend them to
%be
Turing
complete %. %computational languages.  
%In this direction, 
%and motivated by model-theoretic studies of the \lcalculus, 
languages,
Scott~\cite{ScottTCS}
introduced an extension of Typed Combinatory Logic with a 
fixpoint combinator. %for general recursion.  
Plotkin~\cite{PlotkinLCF} 
studied the calculus as a programming language, 
%%in the form of a %, shifting from Typed Combinatory Logic to
%%Simply-Typed Lambda Calculus, and 
%In doing so, he 
%opened up further possible distinctions in the study of type theories for
%computation; namely, the consideration of equational theories for
%different function call mechanisms: by value or by name,
%%~\cite{PlotkinCBVCBN}, 
%as in ML or Haskell.  
and enriched type theory with concerns from programming: the distinction
between call-by-value and call-by-name~\cite{PlotkinCBVCBN}.

%The influence of mathematical models on type theories, logical systems, and
%programming languages %as advocated by Scott~\cite{?} 
%plays a central role throughout the proposal.

\subsection{Questions and pathways for research}

We %now %Having introduced the scientific framework, we 
turn attention to topics of active research, identifying questions and
pathways for research.  They are the basis for the research methodology and
plan outlined in the next section.

\setcounter{paragraph}{0}
\paragraph*{Foundations.}

Type theories are ubiquitous in computer science and mathematics, but 
%We have already mentioned several type theories, and we will mention a few
%more in the sequel.  However, 
the following fundamental question remains open:
\req{(\ref{AlgebraicTypeTheoryParagraph})}
{What is a type theory?}

\hide{
Section\pref{AlgebraicTypeTheoryParagraph} aims at a comprehensive
mathematical answer, that will also serve as a framework for our other
type-theoretic developments.  
\hide{
  We regard this as a step towards the related open question:
\req{}{What is a programming language?}
that will be kept in the background of our investigations.
}
}

\hide{%%% BEGIN hide
\paragraph*{Logical wiring.}
\label{LogicalWiringSubsection}

Lambek~\cite{LambekI} recognised the similarity between the basic
structure of a category, given by identities and
composition~(see\pref{Category}), and the wiring of logical deduction
systems, specifically Gentzen's sequent calculi~\cite{Gentzen1935}, given
by the axiom and cut rules.  Intuitionistic sequents led
Lambek~\cite{LambekII} to axiomatize their algebra under the concept of
multicategory.  This correspondence is roughly as outlined below (where the
vector notation $\vec{\,\cdot\,}$ stands for finite sequences of objects): 
\begin{center}\begin{tabular}{|c|c|}\hline
    \begin{tabular}{c}
      \small Intuitionistic\\[-1mm] \small Sequent Calculus
  \end{tabular}
  &
  \begin{tabular}{c}
    \small
    Multicategories
  \end{tabular}
  \\ \hline\hline
  \begin{tabular}{l}
    \hspace*{-15mm}
    \small 
    (Axiom)
    \\[-3mm]
    $\begin{array}{c}\\ \hline P \vdash P\end{array}$
  \end{tabular}
  & 
  \begin{tabular}{l}
    \small
    \hspace*{-7mm}
    (Identity)
  \\[1mm]
  $(A)\stackrel{1_A}\longrightarrow A$
  \end{tabular}
  \\ \hline
  \begin{tabular}{l} 
    \hspace*{-2mm}(Cut) 
    \\[1mm]
    $\begin{array}{c}
    \Gamma\vdash P\quad\Gamma_1,P,\Gamma_2\vdash Q
    \\ \hline 
    \Gamma_1,\Gamma,\Gamma_2\vdash Q
    \end{array}$
  \end{tabular}
  &
  \begin{tabular}{l}
    \small
    \hspace*{-3mm}
  (Multicomposition)
  \\%[1mm]
    \hspace*{-2mm}
    $\begin{array}{c}
    \vec Y \stackrel f\longrightarrow A
    \\
    \vec X,A,\vec Z \stackrel g\longrightarrow B
    \\ \hline\\[-4mm]
    \vec X, \vec Y,\vec Z
    \xymatrix@C=25pt{\ar[r]^-{g\, _{\vec
          X}\hspace*{-.0125mm}\circ\hspace*{-.25mm}_{\vec Z}f}&} 
    B
    \end{array}$
    \hspace*{-2mm}
  \end{tabular}
  \\ \hline
\end{tabular}\end{center}
The analogous for classical sequents was done by Szabo~\cite{Szabo} with the
introduction of polycategories.

The structure of multicategory appeared independently in a very different
mathematical context, the work of May~\cite{May} on homotopical algebra, under
the name of Operad.  Operads are now central to studies in higher-dimensional
algebra (see~\eg~\cite{Leinster,LodayVallette}).  

\rep{(\ref{WiringStructureParagraph})}
{Investigate interactions between logical sequent calculi and algebraic operad
  theory.}
}%%% END hide

\paragraph*{Dependent types.}
\label{DependentTypesParagraph}

Dependent type theory 
was %is a formalism 
introduced by
de~Bruijn. %~\cite{deBruijn} 
It %that 
is an extension of simple type theory that allows types to be indexed by
%(or parameterised by) 
other types.  Such objects abound in computer science and mathematics.  
%For instance, in
%combinatorics one is interested in the type of permutations $\mathfrak
%S(n)$ on $n$ elements, as the index (or parameter) $n$ ranges over the
%natural numbers $\mathbb N$.  In modern notation, this is presented by a
%judgement of the form
%  \[
%  n:\mathbb N \vdash \mathfrak S(n) \mbox{ \mysf type}
%\]
Two %There are two 
essential %fundamental 
constructions on %such 
dependent judgements are: %as follows
%\[
%  \begin{array}{c}
%  i: I \vdash T(i) \mbox{ \mysf type}
%  \\ \hline
%  \vdash \Sigma\,{i:\!I}.\, T(i) \mbox{ \mysf type}
%  \end{array}
%  \qquad
%  \begin{array}{c}
%  i: I \vdash T(i) \mbox{ \mysf type}
%  \\ \hline
%  \vdash \Pi\,{i:\!I}.\, T(i) \mbox{ \mysf type}
%  \end{array}
%\]
%respectively known as 
dependent sums and dependent product types.  %(see~\eg~\cite{Jacobs}).  
Under %They generalise the product and function types of Simply-Typed Lambda
%Calculus and, under 
the Propositions-as-Types correspondence, 
they 
amount to existential and universal quantification.  
%We omit discussing the syntax of terms for these types.  As for their
%equational theory, we only mention that dependent sums may be weak or strong
%and that dependent products may be intensional or extensional.
%%---see \eg~\cite{?}.
%
A crucial %fundamental 
problem %in the area 
is to:
\rep{(\ref{IntensionalTypeTheoryParagraph})}
  {Investigate equality and identity in dependent type theory.}
%This lays at the core of our proposed investigations in
%Section\pref{IntensionalTypeTheoryParagraph}.

Passing %The passage 
from sum and product types to their dependent versions
required new type theories.  Categorical models suggest %further
generalisations 
and the following: %These are the subject of
%Sections\,(\ref{GeneralisedTypeTheoryParagraph}\,\&\,\ref{DirectedTypeTheoryParagraph})
%Section\pref{CalculiSubsection}
%under the following. 
\rep{(\ref{CalculiSubsection})}
  {Develop type theories from mathematical models.}

\hideMathematicalUniverses{
\paragraph*{Mathematical universes.}
\label{MathematicalUniversesParagraph}

The 
construction of %ability to construct 
new mathematical universes of discourse from old ones is 
basic %a fundamental part of the technical toolkit of researchers 
in semantics. 
%, be it either in logic or computation.
%
%One technique to do so 
%is by means of the %to enrich the semantic universe with a `mode of
%variation' by means of a 
%so-called presheaf construction.  
%In its basic form, starting from the universe of sets and functions~$\Set$ one
%considers the universe~$\Set^{\scat C}$ of so-called presheaves consisting of
%the \mbox{$\scat C$-variable} sets for a small category~$\scat C$.  
%The importance of this passage is that the kind of variation embodied in the
%parameter small category translates to new, often surprising, internal
%structure in the universe of presheaves.  
%
%The presheaf construction was introduced by Grothendieck together with a
%more sophisticated and important refinement of it known as the sheaf
%construction~\cite{SGA4} (in the context of the Weil conjectures in
%cohomology theory).  Sheaves are a central object of study in the area of
%mathematics known as Topos Theory~\cite{Elephant}; a topos being a
%universe of discourse for Higher-Order Intuitionistic
%Logic~\cite{LambekScott}.  
%
%Its %The 
%use %of presheaf models 
%in computer science %applications 
%has been prominent. %;~\eg~in 
%concurrency theory, %~\cite{?}, 
%domain theory, %~\cite{?}, 
%lambda calculus, %~\cite{?}, 
%programming-language theory, %~\cite{?}, 
%and type theory. %~\cite{?}.  
%
%In this view, 
%In view of the many possible applications, 
It 
is %thus %it is 
natural to ask:
\reqs{(\ref{MethodologyMathematicalUniversesParagraph})}
  {What mechanisms are there for changing from a type theory to another one
    as universes of discourse? 
%
%    Can this be done while maintaining the relevant computational properties
%    and then incorporated into mechanical proof assistants?
}

One technique to do so 
is %by means of the %to enrich the semantic universe with a `mode of
%variation' by means of a 
the %so-called 
presheaf construction.  

%This question should not only be considered from the topos-theoretic viewpoint
%mentioned above; but also from other approaches, notably that of the related
%forcing technique of Cohen~\cite{Cohen} (introduced by him to prove the
%independence of both the axiom of choice and the continuum hypothesis from
%Zermelo-Fraenkel Set Theory) and its elaboration by Scott and Solovay as
%Boolean-valued models~\cite{ScottSolovay}.

%Recent work in this direction has been done by
%Coquand~\cite{CoquandForcingInTT} and Jaber, Tabareau, and
%Sozeau~\cite{TypeTheoryWithForcing}.\footnote{Note
%  that~\cite{TypeTheoryWithForcing}, despite its title, is about internalising
%  the presheaf construction on partial orders.}
}

\paragraph*{Indexed programming.}
\label{IndexedProgrammingIntro}

%\hideMathematicalUniverses{
In %their most elementary 
discrete form, 
pre\-sheaves can be found in programming
languages as indexed data\-types; programming with which will be %generically
referred to as indexed programming.  
%}

Indexed programming developed from %two main influences:
%(none to do with presheaves though): 
the practical needs of supporting data structures 
with %able to maintain 
strong %data 
invariants 
%, like nested %~\cite{?} 
%and generalised algebraic %~\cite{?}
%datatypes (GADTs) in functional programming~\cite{Omega,Haskell}; 
and the experimentation with dependently-typed programming
languages. %~\cite{Cayenne,Epigram}
%as a by-product of dependent type theory.  
These two views %somehow 
pull in opposite directions and %, as such, 
lead to conceptually different languages.  One is lead to investigate
the following.
\rep{(\ref{IndexedProgrammingParagraph})} 
  {Develop foundational type theories for indexed datatypes.  
   % 
   Design and %subsequently 
   implement indexed programming languages from these and
   %programming-language 
   pragmatics.}

\paragraph*{Resources, effects, modalities.}
\label{ResourcesEffectsModalitiesParagraph}

Girard~\cite{GirardLinearLogic} and Moggi~\cite{MoggiLambdaC} propose two
important analyses of computation, respectively for resources and effects.
The former in the contexts of logic and proof theory; the latter in that
of denotational semantics and category theory.  
%Both %, however, have 
%had tremendous impact in programming-language theory.  
%%The resource analysis in the form of Linear Logic calculi; the effect
%%analysis in the form of Computational \mbox{$\lambda$-calculi}.

\hide{
From the viewpoint of categorical models, the resource and effect management
structures are respectively seen as comonadic and monadic structures.
Comonads and monads being dual categorical concepts that arose in the context
of cohomology theory in the 1960s~\cite{BeckThesis}.

A classical basic result of category theory establishes a strong
correspondence between (co)monads and adjoint functors
(see~\eg~\cite{MacLane}).  One aspect of this is that every adjunction
  \begin{equation}\label{ResourceEffectAdjunction}
  \xymatrix@C=40pt{
    \cat D \ar@/^.75em/[r]|(.625){\mbox{$\,G\,$}}\ar@{}[r]|-\swvdash&
    \ar@/^.75em/[l]|(.625){\mbox{$\,F\,$}} \cat C }
\end{equation}
with $F$ and $G$ respectively left and right adjoint to each other, gives
rise (by composition) to a comonad on $\cat D$ and a monad on $\cat C$.  This
viewpoint gave impetus to further analyses based on the more primitive notion
of adjunction.  

Models of Linear Logic are founded on the theory of
monoidal categories~\cite[Chapter~VII.1]{MacLane}.  For them, one requests
that the adjunction be monoidal with respect to linear structure on $\cat D$
and cartesian (or multiplicative) structure on $\cat C$
(see~\eg~\cite{MelliesCMLL}).  On the other hand, models of Computational
\mbox{$\lambda$-calculi} rely on enriched category theory~\cite{KellyBook}.
Here, the structure is roughly given by an enriched adjunction with $\cat C$
cartesian and $\cat D$ with powers (or exponentials) relative to $\cat C$.
}

\hide{
In the context of Linear Logic, examples are the mixed linear/cartesian models
and calculi of Benton and Wadler, %~\cite{BentonWadler} 
and of Barber and Plotkin. %~\cite{BarberPlotkin}.  
In the context of effect calculi, examples are the Call-By-Push-Value of
Levy. %~\cite{LevyCBPV} 
and the Enriched Effect Calculus of 
Egger, M{\o}gelberg, and Simpson. %~\cite{EEC}.  
}
The predominant model-theoretic view of resources and effects is as
categorical structures arising at opposite sides of an adjunction.  This
is not so in all models and the following %question 
remains unanswered.
%
\req{\pref{PolarisationParagraph}}
  {How can resources and effects be reconciled and unified?}
%
The answer seems to involve
%To answer it, %a broader view of the subject involving the orthogonal
the logical notion of polarisation~\cite{Andreoli}. %seems to be needed.  
This %From the programming-language viewpoint, this 
%further 
enriches the computational scenario %picture 
with %the ability of distinguishing between 
eager \vs~lazy modes of computation and data
structures.  We incorporate this into the following.
%
\rep{(\ref{PolarisationParagraph}\,\&\,\ref{ProgrammingEffectsParagraph})}
  {Study and develop the theory of resource management, computational
    effects, and polarisation.  Percolate this down into the design of
    programming languages.}

\hide{
From the logical point of view, resource and effect structure are so-called
modal operators, and some of the literature has indeed considered them as
such.  %(see \eg~\cite{Kobayashi}).  %BiermanDePaiva
The field of modal logics is broad, with many subfields of specialised logics
motivated by computation, linguistics, and philosophy.  While a class of modal
logics known as temporal logics have played a prominent role, and been very
successful, in the area of computer aided verification; the impact of modal
logics on programming languages has been peripheral.  It is thus natural and
important to reconsider them in this context. 
%
\rep{(\ref{ModalLogicsParagraph}\,\&\,\ref{MetaprogrammingParagraph})}
  {Investigate the Propositions-as-Types 
    correspondence %Trinity 
    for modal logics, and apply it to programming languages.}
We stress that we are specially interested in modalities for computation with
reflection.
}

\paragraph*{Sequent calculi.}
\label{SequentCalculiParagraph}

A theme %that runs 
orthogonal to the logics under consideration is whether
they are specified in natural deduction or sequent calculus
style. %(see \eg~\cite{vonPlato}).

Work %Most of the work 
on the Propositions-as-Types correspondence has 
focussed on %been done for
natural deduction systems, %; especially in relation to programming
%language theory, where the logical syntax matches that of functional
%languages.  
%
mainly because 
%On the other hand, 
there is as yet no established syntax for sequent %-style 
calculi.  
%A question at the core of this situation is:
Two questions at the core of this situation are:
%
\reqs{(\ref{ProgrammingEffectsParagraph})}
  {What is the categorical algebra of classical sequent calculi? %}
%
\hide{
Nevertheless, a syntactic formalism that is proving robust in applications
seems to be emerging.  This will be referred to as the calculus,
formalism, or system~\SysL; and underlies the ones developed
in~\eg~\cite{CurienHerbelin,Wadler,Munch,CurienMunch}.

The main novel features of the formalism~\SysL, and its philosophy, are an
intrinsic symmetry (corresponding to the computation roles of program and
environment) with syntax that reflects an adjoint situation and is closely
connected to abstract machines (which are in fact internal to the
calculus).  From this perspective we ask:
}
%
%\req{(\ref{ProgrammingEffectsParagraph})}
  %{
%\item 
  What can the proof theory of sequent calculi do for programming?}

\subsection{Research methodology and plan}

This section expands on the research programme outlined in
Section\,\ref{Proposal}, giving just a taste of the full scientific proposal.

\subsubsection{Foundations}
\label{Foundations}

\setcounter{paragraph}{0}
\paragraph{Algebraic Type Theory.}
\label{AlgebraicTypeTheoryParagraph}

\hide{
To understand our approach it is best to start by stripping type theories
down to their essential bare structure.  To do so, let us eliminate type
constructors and binding operators from them.  What one is left with is a
notion of type theory that reduces to that of many-sorted algebraic
theory~\cite{Birkhoff}, a thoroughly studied area of abstract algebra.
}

Many-sorted algebra provides the most basic notion of type theory, with no
binding operators and no type constructors.  Its modern understanding is
through an Algebraic Trinity of three interrelated perspectives:
%as in the Algebraic Trinity of Figure\,\ref{AlgebraicTrinity}.
%\begin{figure}[h]
%  \caption{Algebraic Trinity}
%  \vspace*{-2.5mm}
%    \[\xymatrix@R=7.5pt@C=35pt{
%      & \txt{\small Categorical\\ \small Algebra} 
%      \ar@{<->}[dr]^(.55){\null\hspace*{5mm}\txt{\scriptsize Lawvere~\cite{LawvereThesis}}}
%      \ar@{<->}[dl]_(.55){\null\hspace*{-10mm}\txt{\scriptsize Linton~\cite{Linton}}}
%      & 
%      \\
%      \txt{\small Equational\\ \small Logic}
%      \ar@{<->}[rr]_-{\txt{\scriptsize Birkhoff~\cite{Birkhoff}}}
%      &
%      & 
%      \txt{\small Universal\\ \small Algebra} 
%    }\]
%  \vspace*{-2.5mm}
%\label{AlgebraicTrinity}
%\end{figure}
Equational Logic, Categorical Algebra, and Universal Algebra
(see~\cite{Birkhoff,LawvereThesis,Linton}).
%Equational Logic is the metalanguage of algebraic theories, while
%Universal Algebra is its model theory.  They were both introduced by
%Birkhoff~\cite{Birkhoff}, with the former as a by-product of the latter
%related by a soundness and completeness theorem.  The categorical
%viewpoint of algebraic theories came later, with the work of
%Lawvere~\cite{LawvereThesis} and Linton~\cite{Linton}.  Two crucial
%categorical structures that play a pivotal role here are: Lawvere
%theories~(closely related to Hall's abstract clones) and finitary
%monads~(a notion 
%%originally introduced in cohomology theory by Beck~\cite{Check???}, 
%arising in algebra from free constructions).
%
%Each of the approaches to algebraic theories in the Algebraic Trinity gives a
%different viewpoint of the subject.  Thus, it is important to consider them
%all.  
Each of these approaches %to algebraic theories 
gives a different viewpoint of the subject, and is thus important to consider
them all.  
%
\hide{For instance, Equational Logic provides the deductive system for formal
reasoning about equations from axioms; Universal Algebra provides a general
notion of model from which one derives an abstract notion of syntactic
structure by means of freely generated models; Categorical Algebra provides an
invariant notion of theory together with a notion of translation between them
that lifts to adjoint functors between categories of models.}

\hide{
Having deconstructed type theories down to many-sorted algebraic theories:
Can one reconstruct them back, and in the process enrich them, as a unified
mathematical theory encompassing all aspects of the Algebraic Trinity?  This
is the main general goal of this research track.
}

We will pursue an analogous development for varieties of Type Theory.
%
To make substantial progress, %in the area, 
we will systematically explore a broad spectrum of key features present in
type theories; to wit, term and type variable binding with simple,
polymorphic, and dependent typing, in combination with the orthogonal notion
of linearity.  
\hide{%%% BEGIN hide !!!
The scale at which we will be attempting this is graphically represented by
the research space of investigation in
Figure\,\ref{TypeTheoryResearchSpace}, 
\begin{figure}[h]
\caption{Algebraic Type Theory research space}
\vspace*{1mm}
\begin{center}
$\xymatrix@C=2.5pt@R=25pt{
*+[]{\fbox{\scriptsize\txt{Many-Sorted\\ Univ.\,Alg.
%\\ Eq.\,Logic
}}}
\ar[rrr]|-{\mbox{\scriptsize\txt{\emph{type}\\\emph{dependency}}}}
& 
& 
& 
*+[]{\fbox{\scriptsize\txt{Dependently-Sorted\\ Algebra}}}
\ar[dd]|-{\mbox{\scriptsize\emph{binding}}}
\\
*+[]{\fbox{\scriptsize\txt{Univ.\,Alg.%\\ Eq.\,Logic
}}}
\ar[u]|-{\mbox{\scriptsize\emph{sorting}}}
\ar[rd]|-{\mbox{\scriptsize\emph{binding}}} 
\ar[dd]|-{\mbox{\scriptsize\emph{linearity}}}
& &
*+[]{\fbox{\scriptsize\txt{Simple\\Type Theory}}}
\ar[rd]|-{\mbox{\scriptsize\txt{\emph{type}\\\emph{dependency}}}}
\ar[dd]|-{\mbox{\scriptsize\txt{\emph{explicit}\\\emph{polymorphism}}}}
& 
\\ 
& 
*+[]{\fbox{\scriptsize\txt{Binding\\Algebra}}}
\ar[ur]|-{\mbox{\scriptsize\txt{\emph{simple}\\\emph{types}}}}
&
&
*+[]{\fbox{\scriptsize\txt{Dependent\\Type Theory}}}
%\ar@{.>}[d]
\\
*+[]{\fbox{\scriptsize\shortstack{Operads}}}
& & 
*+[]{\fbox{\scriptsize\txt{Polymorphic\\Type Theory}}}
%\ar@{.>}[r]
& 
%*+[]{\fbox{\scriptsize\txt{?}}}
}$
\end{center}
\vspace*{-3mm}
\label{TypeTheoryResearchSpace}
\end{figure}
where the \emph{linearity} dimension could be transported along or mixed
with all the other ones.
}%%% END hide !!!

This research rests on fundamental work, old and recent, of
Fiore~\etal~\cite{FiorePlotkinTuri,FioreLICS08,FioreHurLMCS,FioreSecOrdEqLog,FioreMahmoud,FioreICALP2012}.
\hide{
that started with the first algebraic model of abstract syntax with variable
binding and progressively led to an extension of the Algebraic Trinity to the
realm of Binding Algebra (\viz~algebraic languages with binding operators).
%
The further extension to simple type theories should follow along similar
lines further taking into account the algebraic structure of types.  As for
the other points of the research space in
Figure\,\ref{TypeTheoryResearchSpace}, they are yet to be investigated.
}
Two salient \emph{goals} are:
%For these we have the following specific goals.
%
%\begin{resenumerate}\setcounter{CC}{0}
%\item
\emgoal{
  To develop a mathematical algebraic framework for the semantics of language
  phrases, whereby free models universally characterise the abstract syntax of
  the language;
  and 
%
%\item
  to %To 
  synthesise metalanguages for type theories in the form of formal systems for
  equational deduction, that are sound and complete for the model theory.}
%
%\item
%  To extract syntactic notions of translations between type theories as
%  suggested by the mathematical models, and to use these to provide
%  constructions for the modular combination of type theories.
%
%\item
%  To test and apply the above mathematical theories by formalising them in
%  Coq or Agda, while at the same time exercising these systems to the
%  limit to identify shortcomings triggering new research in the context of
%  proof assistants.
%\end{resenumerate}
%
The outcome of this work will be new unified metamathematical foundations for
Type Theory.

\hide{
The outcome of this work will be mathematical foundations for the
aforementioned aspects of type theories that are currently treated in an
ad~hoc fashion.  
}
\hide{
Furthermore, it will provide a principled approach by which
to reconsider the problem of integrating polymorphism and dependent types. 
%%%to tackle the problem of integrating polymorphism and dependent types in 
%%% a satisfactory 
%%%manner---the unknown in Figure\,\ref{TypeTheoryResearchSpace}. 
}

\hide{ polynomial functors (from discrete to groupoidal) --- the meaning of
  typing rules}

\hidewiring{%%% BEGIN hide
\paragraph{Wiring structure.}
\label{WiringStructureParagraph}

A basic concern in the investigation of
Section\pref{AlgebraicTypeTheoryParagraph} is the
understanding of the mathematical structure of sequents and judgements.
This subject has already received attention, but new perspectives are emerging
that call for its reconsideration.

Remitting ourselves back to Section\pref{LogicalWiringSubsection}, the
first thing to notice is that there are two possible axiomatizations for
multicategory composition: the one exhibited there, known as partial
composition, and a total (or simultaneous) one of the form
{\small\begin{equation}\label{TotalComposition}\begin{array}{c}
  g: A_1,\ldots,A_n \to B
  \\[1mm]
  f_i:\vec{X_i}\to A_i\enspace(1\leq i\leq n)
  \\[.75mm] \hline\\[-2.5mm]
    g\circ(f_1,\ldots,f_n):\vec{X_1},\ldots,\vec{X_n}\to B
\end{array}\end{equation}}%
In the presence of identities, both notions are equivalent.  An interesting
observation from the theory of operads is that this is not so
otherwise~\cite{Markl}.

The %Algebraically, the 
total composition structure~(\ref{TotalComposition}) is well understood
(see~\eg~\cite{FioreFossacs}).
%as given by semigroup structure (technically in the category of Joyal Linear
%Species~\cite{JoyalAdvMath} with respect to the substitution tensor
%product~\cite{KellyOperads,JoyalLNM1234}).  
But, what about the concept of partial composition?  That, after all, is the
predominant in logic and type theory.  For it, Fiore has recently uncovered an
unexpected connection with structure stemming from the mathematical theory of
\emph{Lie algebras}. %~\cite{LieRef}. 
%; specifically, the pre-Lie algebra structure of
%Gerstenhaber~\cite{Gerstenhaber}.  Furthermore, and analogous, though more
%interesting situation occurs with polycategories~\cite{Szabo} and the
%Lie-admissible algebraic structure of Albert~\cite{Albert}.
%
%We regard these observations as the starting point for interactions between
%operad theory and logic and type theory for pursuing research as follows.
%
This is the starting point for research on
\begin{resenumerate}\setcounter{CC}{0}
\item
  New interactions between abstract algebra, and proof and type theory.
\end{resenumerate}
Specifically, 
\begin{resenumerate}\setcounter{CC}{1}
\item 
  There is a rich variety of algebraic structures with notions of composition,
  \eg~motivated by studies in combinatorics, geometry, physics.  What do they,
  and their mathematical theory, transfer to in proof and type theory?  
  
\item 
  Conversely, the binding operators of logic and type theory introduce
  algebraic structure of a kind so far unseen in mathematics.  This enriches
  the theory and is worth investigating.
\end{resenumerate}

\hide{%%% BEGIN hide
\noindent
From operad theory to logic and type theory:
\begin{resenumerate}\setcounter{CC}{0}
  \item
    Kontsevitch introduced the notion of a meager PROP (renamed
    as \raisebox{.75mm}{\scriptsize$\frac 1 2$}PROP by
    Markl~\cite{Markl}) sitting in between that of operad and
    polycategory.  Is there any use for these in logic and type theory?

  \item
    What about other generalisations, like the preshuffle algebras of
    Ronco~\cite{RoncoShuffleBialgebras} and the permutads of Ronco and
    Loday~\cite{Permutads}?  In particular, the former embodies the
    algebra needed in computational contexts with effects, where the
    notion of total composition does not make sense, and the notion of
    partial composition is subject to weaker axioms to those of
    multicategory.  This is relevant to our proposed investigations in
Sections\,(\ref{PolarisationParagraph}\,\&\,\ref{ProgrammingEffectsParagraph}).

  \item
    There is a rich variety of notions of operads, and notions of
    composition between them, many of which are motivated by
    combinatorics.  What can they offer to logic and type theory?  This
    question is not as vacuous as it may seem at first sight.  For instance,
    the shuffle and quasi-shuffle composition (see~\eg~\cite{AguiarMahajan})
    seem to be related to a type theory for non-commutative linear logic of
    Polakow and Pfenning~\cite{NatDedForIntNonCommLinLog}.

\hide{
  \item
    Foundations for a TTyacc (Type Theory yacc) \ldots this is related to what
    is a Type Theory? \ldots The feasibility of building a TTyacc will be
    investigated \ldots relevant to reflection in type theory \ldots extract a
    language from the mathematics
  }
\end{resenumerate}

\noindent
From logic and type theory to operad theory:
  \begin{resenumerate}\setcounter{CC}{3}
  \item
    The new Lie-algebraic view of multicategory composition mentioned
    above gives a rational reconstruction of an explicit characterisation of
    the pre-Lie operad by Chapoton and Livernet~\cite{ChapotonLivernet}.  The
    case of polycategory composition leads to a conjectured characterisation
    for the Lie-admissible operad.  A possible approach to proving it leads to
    the introduction of new interesting algebraic structures.

  \item\label{BellantoniCook}
    A combination of the \emph{linearity} dimension of
    Figure\,\ref{TypeTheoryResearchSpace} with the other dimensions (that take
    place in the traditional \emph{cartesian} setting, where all structural
    rules on contexts are allowed) leads to a notion of mixed
    operad~\cite{FioreMFPS}, whose associated composition operation is new
    to the mathematical theory.

    Mixed operads are the wiring of linear/cartesian languages, which since
    the work of Bellantoni and Cook~\cite{BellantoniCook} play a role in
    implicit complexity theory.  It would be interesting to develop this
    connection.  A main aim would be to introduce new mathematical structure
    in implicit complexity theory that would help with its problems.  This is
    relevant to our proposed investigations in
    Section\pref{SubstructuralTypeTheoryParagraph}.

  \item
    The algebraic structure on operads considered in mathematics is in
    the tradition of linear algebra (\ie~first-order).  However, the binding
    operators of logic and type theory transport to operads as a new kind of
    algebraic structure which thereby enriches the theory and is worth
    investigating.
  \end{resenumerate}
}%%% END hide 
}%%% END hide

\subsubsection{Models}
\label{Models}

\hidenote{\ldots}

\setcounter{paragraph}{0}
\paragraph{Equality and identity in dependent type theory.}
\label{IntensionalTypeTheoryParagraph}

\hide{
We present 
%We have briefly mentioned  intensional and extensional type theory in
%Section\,\ref{DependentTypesParagraph}.  This section examines these two
%flavours of type theory and presents 
%a 
research %programme 
geared to investigate a main problem in 
type theory: %the area: 
%\begin{resenumerate}\setcounter{CC}{0}
%\item
  To reconcile Intensional and Extensional Type Theory in a computational
  framework.
%\end{resenumerate}
%Various aspects of what follows arose and are under discussion with Warren
%(IAS School of Mathematics, Princeton).
}

Central to a type theory is its associated equational theory, %.  Equational
%theories of dependent type theories may be specified in different (though
%typically equivalent) forms (see~\eg~\cite{Adams}).  For the purpose of our
%informal discussion here it will be enough to assume that they determine a
%congruence relation
%%\[
%%  \Gamma \vdash M_1 = M_2: A
%%\]
%%for valid judgements $\Gamma\vdash M_i:A$ ($i=1,2$), 
or %henceforth referred to as 
judgemental equality.
%
In Pure Intensional Type Theory, judgemental equality is generated by
%so-called $\beta$ 
rules of computation. 
%(recall Section\pref{SectionATypeTheoryAndLogicParagraph}).
%
Extensional Type Theory refers to extensions of Pure Intensional Type Theory
with extensionality features.  These may be given by weak rules of
extensionality or by a strong rule of reflection (of an intensional notion of
equality internal to the type theory as a judgemental equality).
\hide{
In their simplest form,
these consist of incorporating %so-called $\eta$ 
rules of extensionality 
%(recall Section\pref{SectionATypeTheoryAndLogicParagraph}) 
into judgemental equality.  Such extensions will be referred to here as Weak
Extensional Type Theory.
%
%Already for these basic kind of systems, 
%%Pure Intensional and Weak Extensional Type Theory, 
%our knowledge of the subject is limited. %poor.
%%
%For instance, %In particular, 
%the concrete mathematical models (as opposed to syntactic ones) of Pure
%Intensional Type Theory 
%%with dependent sums and products 
%that researchers work with (like groupoids, strict $\omega$-categories, and
%simplicial sets) interpret dependent sums and dependent products as adjoints
%and hence are necessarily models of Weak Extensional Type Theory.  A first
%question arises:
%\begin{resenumerate}\setcounter{CC}{1}
%\item 
%  Are there natural mathematical models of Pure Intensional Type Theory?
%\end{resenumerate}
%In examining this one should also consider:
%\begin{resenumerate}\setcounter{CC}{2}
%\item 
%  What is a model of Pure Intensional Type Theory?
%\end{resenumerate}
%In this respect, we note that the current categorical formalisms
%(see~\eg~\cite{Jacobs}) are not flexible enough (as they are either based on
%adjoint interpretations or on ad hoc variations thereof).  On the other hand,
%the approach of Dybjer on Internal Type Theory~\cite{DybjerITT} will provide
%models within Cartmell's Generalised Algebraic Theories~\cite{Cartmell}.
%Their relationship to developments in the Algebraic Type Theory of
%Section\pref{AlgebraicTypeTheoryParagraph} will be investigated.
%}
%%
%\hide{CwFs \vs~substitution algebras}
%
The discussion of further notions of extensionality requires the consideration
of Identity Types,
%  \[
%  x:T,y:T\vdash\Id_T(x,y)\ \mbox{\small$\mathsf{type}$}
%\]
that %are meant to 
provide a notion of equality internal to the type theory.
%
In Strong Extensional Type Theory, %one requires that 
Identity Types completely internalise judgemental equality
(see~\cite{MartinLofETT}).  %Since for rich
%type theories this leads to undecidable type checking, Pure Intensional Type
%Theory has become the foundational core of proof assistants.  %for constructive
%%mathematics. %; like
%%Automath, %~\cite{deBruijn}, 
%%Coq, %~\cite{?} 
%%and 
%%Agda. %~\cite{?}.  
%However, as extensionality lays at the core of mathematics and its practice,
%the quest for a computational framework in between Pure Intensional Type
%Theory and Strong Extensional Type Theory continues.
}

Conceptually, one may regard a type theory as a basic deduction system
extended with type constructors. 
%For instance, Martin-L\"of Type Theory embodies dependent sums and dependent
%products, as well as Identity Types, W~Types, and Universes~\cite{ProgMLTT}.
As such, a type theory is to be thought not just as a single universe of
discourse, but rather as a variety of universes of discourse. 
%, corresponding to the various possible restrictions.  
From this perspective, it is important to understand the interaction of
modularly extending type theories.  A key notion %in this respect 
is that of conservative extension.
Informally, an extension %$T'$ of $T$ is conservative if the restriction of the
%judgemental equality of $T'$ to $T$ coincides with the judgemental equality of
%$T$.
of a type theory is conservative when it does not modify the judgemental
equality of the extended type theory.
%
%Conservativity has been studied for some simple type 
%theories (see~\eg~\cite%[Section\,3]
%{FioreRemarks}), 
%but it has been overlooked for dependent type theories.

Research on the 
so far unexplored 
problem of conservativity in dependent type theory 
%, which has so far been overlooked, 
will be pursued.  We will aim at being comprehensive, targeting Intensional
and Weak/Strong Extensional type theories with
dependent sums, dependent products, inductive types, and universes.  Our
\emph{goal} is to deepen the current understanding of equality/extensionality
and identity/intensionality in type theory via a model-theoretic framework for
establishing conservative extension results.  

The full scientific proposal further expands on the above in connection to an
interesting recent approach to extensionality in Martin-L\"of Type Theory,
\viz~Voevodsky's Univalence Axiom~\cite{UnivalentProgramme}. 
%, motivated by Homotopy Theory.  
Those considerations 
\emph{aim} %are geared 
to investigate a main problem in type theory: To reconcile Intensional and
Extensional Type Theory in a computational framework.

\hide{
%This will lead to 
%sophisticated model constructions that will deepen our understanding of the
%subject.  
%%In particular because it was already the case in proving the conservative
%%extension of the Simply-Typed Lambda Calculus extended with (non-dependent)
%%extensional strong sums~\cite{FioreRemarks}.

%In this direction, 
Martin-L\"of proposed a notion of Identity Type
(see~\eg~\cite{ProgMLTT}) %which is given as an inductively generated
%family~\cite{DybjerIF} with a reflexivity constructor.  These Identity Types
that is %are 
currently receiving renewed special attention, stemming from connections
with homotopy theory and higher-dimensional category
theory~\cite{UnivalentProgramme}.  For them some fundamental matters are still
to be understood.  For instance, 
\hide{
whether the introduced intensional identity
%(an internal notion of equality) 
affects 
%(the external) 
judgemental equality.
Precisely,
}
\begin{resenumerate}\setcounter{CC}{4}
\item
  Does extending (Pure Intensional or Weak Extensional) Type Theory for
  dependent sums and dependent products with Identity Types produce a
  conservative extension?
\end{resenumerate}
\hide{
Also, when seen from the viewpoint of Algebraic Type Theory:
\begin{resenumerate}\setcounter{CC}{5}
\item 
  What is the relationship between the algebraic theory of Identity Types and
  the algebraic theory of Weak Higher-Dimensional Categories?
\end{resenumerate}
This is to be investigated specially noting that the former crucially relies
on binding operators while this is not so for the latter.
}

A recent interesting approach to extensionality being intensively
investigated is the extension by Voevodsky of Pure Intensional Type Theory
with a so-called Univalence Axiom %~\cite{UnivalenceAxiom} 
motivated by homotopical foundations.  This axiom has been shown consistent
from the outset.  We will furthermore tackle the but:
\begin{resenumerate}\setcounter{CC}{6}
\item 
  Is the extension of (Pure Intensional or Weak Extensional) Martin-L\"of Type
  Theory with the Univalence Axiom conservative?
\end{resenumerate}
}

\hideLId{
It is an open problem, referred to as Voevodsky's Main Computational
Conjecture, as to how to give a constructive interpretation of the Univalence
Axiom.  The question arises as to whether one could circumvent this problem by
concentrating instead on a notable consequence of the axiom; namely, that it
somehow endows Identity Types with logical character, very roughly in that the
Identity Type constructor logically commutes with all the other constructors
up to equivalence.  Therefore we will rather aim to:
\begin{resenumerate}\setcounter{CC}{7}
\item 
  Design a constructive type theory with built-in Logical Identity Types.
\end{resenumerate}
In this context, we will:
\begin{resenumerate}\setcounter{CC}{8}
\item
  Investigate possible connections to Strachey's notion of parametric
  polymorphism~\cite{Strachey1967} in computer science as formalised by
  Reynolds~\cite{Reynolds} using logical relations.
\end{resenumerate}
Were these connections to materialise, it will open up a new flow of ideas
between hitherto disconnected fields.
}

{\color{red} REVISION STOPPED HERE}
\hideMathematicalUniverses{
\paragraph{Mathematical universes.}
\label{MethodologyMathematicalUniversesParagraph}

We speculate here on directions in relation to
Section\pref{MathematicalUniversesParagraph}, where we are interested in
building new models from old ones.  A main source of inspiration and
guidance for the development comes from Topos Theory~\cite{Elephant}.

Our proposal is to:
\begin{resenumerate}\setcounter{CC}{0}
\item\label{ConstructionsOnTypeTheories}
  Investigate presheaf, orthogonality, sheaf, glueing, and forcing
  constructions on type theories.
\end{resenumerate}
with the general goal to:
\begin{resenumerate}\setcounter{CC}{1}
\item
  Build a mathematical theory of constructions on type theories that
  produce new type theories from old ones, together with an interpretation
  (or compilation) of the latter into the former.
  
\item \label{MMUPAppItem}
  Implement the mathematics in proof assistants, and exercise it in
  applications.  
\end{resenumerate}

Initial work along this line has been done by Jaber, Tabareau, and
Sozeau~\cite{TypeTheoryWithForcing}, who formalised the presheaf
construction on a partial order in the Calculus of Constructions, and then
re-interpreted back the Calculus of Constructions in the internal presheaf
model.  However, not only a wide spectrum of other possible constructions
(as detailed in
item~\textbf{\ref{MethodologyMathematicalUniversesParagraph}}\thinspace\itemref{ConstructionsOnTypeTheories}
above) remains to be explored; but, even in this basic setting, work remains
to be done: notably to incorporate Inductive Types.

What we envisaged is also more general, in that we take the view that the
new type theory need not be of the same character as the old one.  Thus
there are two aspects to our proposed investigations: the mathematical one
of studying model constructions and the type-theoretic one of designing
internal languages.  To fix ideas, consider as an example that the
presheaf construction on a monoidal category endows the new universe of
discourse with monoidal closed structure (formally via Day's convolution
monoidal structure~\cite{Day}), and thereby introduces new linear structure.
This is important in many applications, for which the reader may
consult~\cite{FioreFossacs}, %, and complementary to our proposed
%investigations in Section\pref{SubstructuralTypeTheoryParagraph}.
that will be at the core of
item~\mbox{\textbf{\ref{MethodologyMathematicalUniversesParagraph}}\thinspace\textbf{\itemref{MMUPAppItem}}}
above.

\hide{
The study of Cohen Forcing for type theory was stated as an open problem
by Beeson~\cite{BeesonBook}, who wrote: ``Forcing has yet to be worked out
directly for Martin-L\"of's system.''.  Work in this direction has only
recently begun.  Specifically, by Coquand and Jaber~\cite{CoquandNote} who
presented an interesting example.  Much remains to be explored both with
respect to the vast literature on set-theoretic forcing and, more
relevantly, with respect to Krivine's Classical
Realizability~\cite{KrivineRA}, an extension of forcing.
}
}

\paragraph{Polarised logic.}
\label{PolarisationParagraph}

We propose here a model-theoretic study of a rich variety of logical
systems encompassing aspects of resource management and computational
effects (recall Section\pref{ResourcesEffectsModalitiesParagraph}).  This
research is work under discussion with Curien and
Munch-Maccagnoni (Laboratoire~PPS, Universit\'e Paris Diderot - Paris~7).
A distinguishing novelty of our approach is that it will be pursued as
informed by the logical notion of polarisation.

The notion of explicit polarisation in logical systems was introduced by
Andreoli~\cite{Andreoli} in his study of proof search in Linear Logic, in
particular by focalisation.  According to it, logical connectives are
classified as either being positive or negative; with these two worlds
being dual to each other (categorically by adjunction).  Following
Andreoli's work, Girard understood the relevance of focalisation via
polarisation as a way to tame down the inherent non-determinism in
computation (by cut elimination) in classical logic; and, in this
direction, introduced the first explicitly polarised logic:
\LC~\cite{GirardLC}.  A crucial aspect of this system, is to make eager
(\viz~call-by-value) and lazy~(\viz,~call-by-name) modes of computation
explicit, allowing for their combination in a framework with eager and
lazy data structures.

Polarisation, though not recognised as such, is also present in the
Computational \mbox{$\lambda$-calculi} based on adjunction models as mentioned
in Section\pref{ResourcesEffectsModalitiesParagraph}.  This key
observation leads to a refinement of the
model~(\ref{ResourceEffectAdjunction}) as follows
  \begin{equation}\label{LCBPV}
  \begin{minipage}{\columnwidth}
  \xymatrix@R=5pt{
    & & 
    \\
    \ar@{}[u]|-{\text{\scriptsize(negative)}}
    \cat N \ar@/^.75em/[dr] & \ar@{}[rd]|-{\sevdash} \ar@{}[dl]|-{\swvdash} &
    \ar@/^.75em/[dl] \cat M \ar@{}[u]|-{\text{\scriptsize(multiplicative)}}\\
    & \ar@/^.75em/[lu] \cat P \ar@/^.75em/[ur]
    \ar@{}[d]|-{\txt{\scriptsize(positive/linear)}}& 
    \\ & & 
  }
\end{minipage}
\end{equation}
with comonadic resource structure given by the adjunction on the right and
monadic effect structure given by the adjunction on the left.

When the effect structure is trivial, the above restricts to models of Linear
Logic~\cite{MelliesCMLL}; while, when the resource structure is trivial, one
recovers the models of Call-By-Push-Value~\cite{LevyCBPV}.  One is thus led to
the following programme.
\begin{resenumerate}\setcounter{CC}{0}
\item\label{ItemOne}
  Devise a sound and complete logical system for the model
  theory~(\ref{LCBPV}), and relate it to existing polarised
  systems~\cite{Munch,CurienMunch}.
\item
  A first model-theoretic result shows that the effect structure on $\cat
  P$ lifts to one on $\cat M$.  Show that this corresponds to an encoding
  of Call-By-Push-Value in the devised logical system.
\item
  Develop a generic syntactic theory able to incorporate concrete
  computational effects into the devised calculus.  Two research possibilities
  are: (i)~considering algebraic theories of effects following the work of
  Plotkin and Power~\cite{PlotkinPowerAlgOpsAndGenEffs}, and
  (ii)~revisiting and refining Filinski's result on representing
  monads~\cite{Filinski} that reduces general monadic effects to the
  storage and escape effects.  
  \hide{Either of these approaches will necessarily
  have to overcome serious shortcomings.  For instance, on the one hand,
  there is as yet no general operational theory of algebraic effects; and,
  on the other, a logical system for storage is not yet in place.  These
  problems will be investigated.}
\item\label{ItemFour}
  The previous development will provide foundational metalanguages.  The
  next step in the programme is to promote them to programming languages,
  where the effects are implicit, by putting them in so-called
  direct-style.  The kind of model theory involved in this development
  will be discussed below.
\end{resenumerate}

The model theory of~(\ref{LCBPV}) can be further specialised to
models with enough internal structure so as to extend the picture as
follows
\begin{equation}\label{LEEC}
  \begin{minipage}{\columnwidth}
  \xymatrix@R=5pt{
    & \ar@/^.75em/[dl] \ar@{}[dl]|-\sevdash \cat N \ar@/^.75em/[dr] &
    \ar@{}[rd]|-{\sevdash} \ar@{}[dl]|-{\swvdash} & \ar@/^.75em/[dl] \cat
    M \\
    \ar@/^.75em/[ur] \cat P^\op & & \ar@/^.75em/[lu] \cat P \ar@/^.75em/[ur] & 
  }\end{minipage}
\end{equation}
where the new adjunction between the negatives and the opposite dual of
the positives enriches the models with control structure of the
linear %~\cite{ThieleckeEtAl} 
and/or delimited continuation %~\cite{PeytonJones?}
kind, though this is to be investigated.

The model theory~(\ref{LEEC}) refines that of two recent developments: the
aforementioned Enriched Effect Calculus, which amounts to the case in
which the linear structure collapses to cartesian one; and the Tensor
Logic of Melli\`es and Tabareau~\cite{TensorLogic}, where the effect
structure is collapsed.  Our programme for~(\ref{LCBPV}),
\ie~items~\mbox{\textbf{\ref{PolarisationParagraph}}\thinspace\textbf{[\ref{ItemOne}--\ref{ItemFour}]}},
will be then also pursued for~(\ref{LEEC}) drawing connections with these
works.

Let us now return to the distinction between metalanguages and programming
languages briefly mentioned in
item~\textbf{\ref{PolarisationParagraph}}\thinspace\itemref{ItemFour} above.
Model theoretically, this can be roughly seen as follows: whereas the
metalanguage is an internal language for co/monadic structure; the programming
language is the internal language for the derived structure of co/free
algebras for the co/monad---typically defined by the categorical co/Kleisli
construction~\cite{MacLane}.  In particular, the Kleisli category of a
computational monad provides a model of call-by-value~\cite{MoggiLambdaC};
while the coKleisli category of a linear exponential comonad provides one
of call-by-name~(\ie,~a cartesian closed category)~\cite{Seely}.

The programming languages corresponding to the metalanguages of the
polarised models~(\ref{LCBPV}) and~(\ref{LEEC}) should be the internal
languages for an analogous, but more intricate, construction relative to
an adjunction rather than a co/monad.  In this context, the following will
be investigated.
\begin{resenumerate}\setcounter{CC}{4}
\item
  What is the mathematical structure of such construction?  
  %
  The question is challenging.  For instance, we envisaged models of {\LC} to
  arise in this manner and, because of the interaction with call-by-value and
  call-by-name modes of computation, the structure should be more general than
  that of a category.

\item
  What would then be the mathematical theory of universal constructions for
  these structures giving rise to logical connectives?  
  %
  Note again from considering {\LC} that this will be subtle, requiring
  constructions of objects with mixed polarities.

\item
  What can the resulting direct-style calculi contribute to programming?
  For discussion on this see Section\pref{ProgrammingEffectsParagraph}.
\end{resenumerate}

\hidenote{polarised dialectica for program extraction motivated by
  polarised dialectica category construction}

\hidenote{Investigate relationship to Chuck Liang \& Dale Miller's work}

\hidenote{Investigate relationship to Oleg Kiselyov and Chung-chieh Shan's
  substructural type system for delimited control}

\paragraph{Modal logics.}
\label{ModalLogicsParagraph}

Recall from Section\pref{ResourcesEffectsModalitiesParagraph} that resource
comonads and effect monads are modalities.  Polarisation in the context of
modal logics has not been considered yet.  We will thus complement the
previous section with the 
\begin{resenumerate}\setcounter{CC}{0}
\item
  Investigation of Polarised Modal Logics.
\end{resenumerate}

Our overall goal here is to
\begin{resenumerate}\setcounter{CC}{1}
\item
  Develop the Propositions-as-Types correspondence for modal logics.
\end{resenumerate}
Here, we are specially interested in grounding it through categorical models
(see Figure\,\ref{ResearchAreas}), that have been scarcely investigated.

Looking ahead in connection to the proposed research in
Section\pref{MetaprogrammingParagraph}, we will be particularly interested
in studying: 
%
Borghuis' Modal Pure Type Systems~\cite{ModalPTS};
%
Artemov's Logic of Proofs~\cite{ArtemovLP} and related systems;
%(\eg~\cite{AltArtemov,ArtemovIemhoff}); 
%Daniyar S Shamkanov: Strong Normalization and Confluence for Reflexive
%  Combinatory Logic 
%
Mendler's Multimodal CK~\cite{MendlerMMCK};
and
%
Park and Im's Calculus $S_\Delta$~\cite{ParkIm}.

\subsubsection{Calculi}
\label{CalculiSubsection}

\hidenote{\ldots}

\setcounter{paragraph}{0}
\paragraph*{Categorical Type Theory.}

The evolution of term and type structure in type theories from its origin
to the current state can be succinctly represented as follows
\begin{center}\begin{tabular}{|c||c|c|c|c|}\hline
  & \small Term Structure & \small Type Structure
  \\[.25mm] \hline\hline
  \small Equational Logic & \small Algebraic & \small ---
  \\[.25mm] \hline
  \txt{\\ \raisebox{1mm}{\small Simple Type}\\\raisebox{1mm}{\small Theory}} &
  \raisebox{-.75mm}{\small Binding} & \raisebox{-.75mm}{\small Algebraic}
  \\[.25mm] \hline
  \txt{\\ \raisebox{.5mm}{\small Polymorphic}\\ \raisebox{1mm}{\small Type
      Theory}} & 
  \raisebox{-.75mm}{\small Binding} &
  \raisebox{-.75mm}{\small Binding} 
  \\[1.5mm] \hline
  \txt{\\ \raisebox{.5mm}{\small Dependent}\\\raisebox{1mm}{\small Type Theory}}
  & 
  \multicolumn{2}{c|}{\raisebox{-1mm}{\small Binding}}
  \\[.25mm] \hline
\end{tabular}\end{center}
Surprisingly, this has not changed since de~Bruijn's Automath in the late
1960s.  We firmly believe that further evolution will come from mathematical
input and, in this direction, propose two lines of research for type theories
as formal languages of categorical structures.  
%These are presented in the two sections below.

\paragraph{Generalised Type Theory.}
\label{GeneralisedTypeTheoryParagraph}

We present work discussed with Gambino (DMI, Universit\`a degli Studi di
Palermo) and Hyland (DPMMS, University of Cambridge) with whom the development
will be undertaken.  

Section\pref{MethodologyMathematicalUniversesParagraph} considers presheaf
categories in the large as mathematical universes of discourse (or gros
toposes).  There is however an alternative view of them in the small as
spaces (or petit toposes) with morphisms between them given by linear (or
cocontinuous) functors, resulting in the bicategory of profunctors (or
distributors)~\cite{Benabou}.  Lawvere~\cite{LawvereMetric} considered
these structures in the general context of enriched category theory and
synthesised a Generalised Logical Calculus out of them.  The enrichments
give rise to different interpretations, \eg~in preorders, categories, and
metric spaces.

The general goal of our research here is to:
\begin{resenumerate}\setcounter{CC}{0}
\item\label{GeneralisedTypeTheoryItem}
   Develop a type theory providing a Propositions-as-Types interpretation of
   the Generalised Logical Calculus.
\end{resenumerate}
This is to be investigated in two stages, respectively corresponding to
enrichment over cartesian closed categories and over symmetric monoidal
closed categories.  
\hide{
In the first case, one roughly has the following
translation table between the Generalised Logical Calculus and Predicate
Logic.
\begin{center}\small\begin{tabular}{|c|c|}\hline
    sum & disjunction\\ 
    coend & existential quantification\\ 
  product & conjunction\\ 
  end & universal quantification\\ 
  exponential & implication\\ 
  hom & equality\\ 
  presheaf application & predicate membership\\
  \hline
\end{tabular}\end{center}
The second stage introduces linearity. 
}

Success 
in this endeavour %here 
will yield a system where the mathematical
calculations done with profunctors (\eg~in the context of the coherence of
profunctor composition~\cite{Benabou}, the mathematical theory of
substitution~\cite{FioreFossacs}, generalised species of
structures~\cite{Species}, and generalised polynomial
functors~\cite{FioreICALP}) can be established formally.

Intuitively, the coend and end constructions mentioned in the table above are
quotients of dependent sums under a compatibility condition and restrictions
of dependent products under a parametricity condition.  Taking this view
seriously, we will aim to:
\begin{resenumerate}\setcounter{CC}{1}
\item
  Extend
  item~\textbf{\ref{GeneralisedTypeTheoryParagraph}}\thinspace\itemref{GeneralisedTypeTheoryItem}
  to a fully-fledged, possibly higher-dimensional, dependent type theory.  
\end{resenumerate}

\hide{
Finally, in an intriguing complementary direction,
Fiore~\cite{FioreFossacs} has suggested a graphical reading of the
Generalised Logical Calculus that is surprisingly close to formalisms for
categorical graphical languages~\cite{Selinger}, and
proof~\cite{GirardLinearLogic}/interaction~\cite{Lafont} nets.  This
aspect deserves investigation.
}

\paragraph{Directed Type Theory.}
\label{DirectedTypeTheoryParagraph}

Homotopy Type Theory (under the acronym HoTT, and the slogan `types are
spaces') is the name being used for the body of work at the boundary between
Homotopy Theory and Dependent Type Theory through Identity Types.

The question arises:
\begin{resenumerate}\setcounter{CC}{0}
\item
  Is there a notion of Directed Type connecting Higher-Dimensional Category
  Theory and Dependent Type Theory on which to establish a body of work on
  Directed Type Theory (DiTT)?
\end{resenumerate}
In this context, while an Identity Type
  \[
  \Id_T(x,y)
\]
is intended to establish the intensional equality of elements of a type $T$ so
that $\Id_T(x,y)$ and $\Id_T(y,x)$ are equivalent; the intuitive idea behind a
Directed Type 
  \[
  \Di_T(x,y)
\]
would be to classify the possible ways in which an element of a type $T$ may
evolve to another one in a possibly irreversible manner; as it happens, for
instance, in computation.  Thereby leading to not necessarily equivalent types
$\Di_T(x,y)$ and $\Di_T(y,x)$, and to the DiTT slogan `types are directed
spaces'.

An analysis of the elimination rule for Identity Types reveals that there
are at least two sources leading to their inherent reversible character.
We mention them below as possible directions of research for investigating
Directed Types.
\begin{resenumerate}\setcounter{CC}{1}
\item
  The use of contexts that allow the commutativity of two consecutive
  variables of the same type, which suggests moving on to a non-commutative
  setting (as \eg~in non-commutative linear logic).
  %~\cite{NatDedForIntNonCommLinLog}).  
  %(This is related to the considerations in
  %Section\pref{SubstructuralTypeTheoryParagraph} below.)

\item\label{TypeDualityItem}
  The lack of a type duality, a property that seems to be inherent to
  directionality, whereby every type $T$ has an associated dual type $T^\circ$
  for which $\Di_T(x,y)$ and $\Di_{T^\circ}(y,x)$ are equivalent.  
\end{resenumerate}

The previous item is of course related to the considerations of the
previous section.  Pursuing this further, it would be natural to
generalise from presheaves, now regarded as discrete fibrations, to the
general notion of fibration~\cite{SGA1} and aim to:
\begin{resenumerate}\setcounter{CC}{3}
\item
  Develop a type theory modelled on Grothendieck fibrations.
\end{resenumerate}

\hide{
As a final point, we note that yet another approach would be to:
\begin{resenumerate}\setcounter{CC}{4}
\item
  Investigate a directed-cylinder construction, intuitively adding a new base
  point to a type that canonically evolves to every element.  
\end{resenumerate}
This would generalise the lifting construction of Domain
Theory~\cite{LiftingKZ} and possibly establish connections with the
Complete Cuboidal Sets model of Axiomatic Domain
Theory~\cite{CompleteCuboidalSets} opening new directions.  
}

\hide{%%% BEGIN hide
\paragraph{Substructural Type Theory.}
\label{SubstructuralTypeTheoryParagraph}

The work of Bellantoni and Cook~\cite{BellantoniCook}
%, after Leivant~\cite{Leivant}, 
on implicit computational complexity gave a characterisation of the
polynomial-time computable functions on finite strings by means of a
safe-recursion scheme that restricted primitive recursion.  A crucial
ingredient of their approach is to divide function arguments into normal and
safe ones.  Recursion is allowed on normal arguments but not on safe ones.  In
the terminology that we have been using, normal arguments are linear while
safe ones are cartesian.  In fact, the notion of safe composition
coincides with that for mixed linear/cartesian operads (see
item~\textbf{\ref{WiringStructureParagraph}}\thinspace\itemref{BellantoniCook}).
%on page~\pageref{BellantoniCook}).  
Since then, the role of resource management in taming complexity has been
prominently recognised, and a big body of work on implicit computational
complexity (\eg~on soft, bounded, light, and linear logics and types) has
followed.
% Baillot, Dal Lago, Hofmann, Lafont

Recently, Beckmann, Buss, and Friedman~\cite{BeckmannBussFriedman} 
%and later on Arai~\cite{Arai} 
have extended the safe-recursion scheme to arbitrary set functions.
Changing viewpoint form the mathematical universe of sets to that of
types, one is lead to:
\begin{resenumerate}\setcounter{CC}{0}
\item
  Extend dependent type theory to the mixed linear/cartesian setting.
\end{resenumerate}
In particular, in what concerns to taming computational complexity:
\begin{resenumerate}\setcounter{CC}{1}
\item
  Investigate a safe version of the Induction-Recursion scheme of
  Dybjer~\cite{DybjerIR}.  
\end{resenumerate}
This development should be also accompanied by a model theory, possibly based
on 
\begin{resenumerate}\setcounter{CC}{2}
\item 
  Realisability models relative to the Linear Combinatory Algebras of
  Abramsky, Haghverdi, and Scott~\cite{AHS}.
\end{resenumerate}
}%%% END hide

\subsubsection{Programming}
\label{Programming}

This track of the proposal is concerned with the design and implementation
of programming languages from first principles and pragmatics, followed up
by their subsequent test, use, and distribution.

Three directions for this research are presented.  These will be
considered as units in their own right, and also in relation to each other.

\setcounter{paragraph}{0}
\paragraph{Indexed programming.}
\label{IndexedProgrammingParagraph}

This section outlines work under discussion with Ahn and Sheard
(Department of Computer Science, Portland State University) on indexed
programming (recall Section\pref{IndexedProgrammingIntro}).

The main problem to be addressed here is to:
\begin{resenumerate}\setcounter{CC}{0}
\item
  Design a language for programming indexed data structures that will
  inform the design of the next generation of programming languages.
\end{resenumerate}
The task can be approached from two different viewpoints.  In a top-down
fashion, one may turn dependent type theories into programming languages;
conversely, in a bottom-up fashion, one may extend functional programming
languages with indexing structure.  These two approaches pull the
programming language design into two opposite directions. 
\hide{; as pictorially
presented in Figure\,\ref{DesignSpaceFigure}, 
\begin{figure}[h]
\caption{Indexed-programming design space}
\vspace*{-2mm}
\[
\null\hspace*{-2mm}
\txt{\small Functional\\ \small Programming\\ \small Language}
\hspace*{3mm}\xy
0;/r.75pc/:+(0,.25)
,{\hloop-\hcross\hcross\hcross\hcross\hcross\hcross\hcross\hloop}
\endxy\null\hspace*{3mm}
\txt{\small Constructive\\ \small Proof\\ \small Assistant}
\vspace*{-2mm}
\]
\hdashrule[1ex]{\columnwidth}{1pt}{2.5pt}
\begin{tabular}{lcl}
%\scriptsize Programming && \scritpsize Proof\\ 
\small
ML, %~\cite{SML}, 
Haskell %~\cite{Haskell} 
&\hspace*{25mm}&
\small
Coq, %~\cite{Coq}, 
Agda %~\cite{Agda}
\\[-.5mm]
\small
\SystemFomega %~\cite{GirardSystemF} 
&& 
\small
MLTT, %~\cite{MLTT}, 
ICC %~\cite{ICC}
\\[-.5mm]
\small
Polymorphism %~\cite{Strachey1967} 
&& 
\small
Type dependency %~\cite{deBruijn} 
\\[-.5mm]
\small
Impredicativity && 
\small
Universes %~\cite{Universes}
\\[-.5mm]
\small
Type inference %~\cite{Milner1978,Wells} 
&&
\small
Type checking
\\[-.5mm]
\small
Recursive types %~\cite{FPC}
&& 
\small
Inductive types~ %\cite{InductiveFamilies}
\\[-.5mm]
\small
Equality types %~\cite{GHC-IFP2012} 
&& 
\small
Identity types %~\cite{IdTypes}
\end{tabular}
\label{DesignSpaceFigure}
\end{figure}
where the main features inherent to each approach are also listed.}

Most of the work in this area has concentrated on exploring the top-down
approach (see~\eg~\cite{Cayenne} and \cite{Epigram}).  Here we will pursue
research on the bottom-up approach, which has not been explored
systematically.  A strong pragmatic reason for this is that our uppermost
interest is in a language targeted to programmers.  This does not mean that we
are not interested in the activity of proving as in constructive proof
assistants.  However, our main goal, under the Propositions-as-Types paradigm,
is: 
\begin{resenumerate}\setcounter{CC}{1}
\item
  To prove properties as a by-product of programming; rather than to extract
  programs from proving properties.
\end{resenumerate}
For this to be meaningful, both in theory and practice, the language will
need to guarantee logical consistency in a setting that naturally allows
programming strong invariants for rich indexing type structure.  For
instance, so that the code of a compiler guarantees its correctness.

In the above direction, we are investigating:
\begin{resenumerate}\setcounter{CC}{2}
\item
  {\SystemFi}: an extension of {\SystemFomega} with static type-indexing
  structure.
\end{resenumerate}
{\SystemFi} is to serve as the mathematical foundation for designing, and
giving operational semantics to, our programming language.  We stress here
that indices are static, \ie~determined at compile time.  This is the main
feature pulling us away from traditional dependently-typed formalisms.

The logical consistency of {\SystemFi} amounts to establishing its strong
normalisation; while as an extension of {\SystemFomega} it will embody
rich type structure.  In this setting, following Ahn and
Sheard~\cite{AhnSheard11}, we will:
\begin{resenumerate}\setcounter{CC}{3}
\item
  Study programming primitives corresponding to a variety of induction
  proof principles.
\end{resenumerate}
These are to be incorporated in the programming language design.  Here, it
is interesting to note that the scheme provided by Mendler's
iterator~\cite{MendlerIter}, that crucially takes advantage of
parametric polymorphism, allows for terminating iteration over recursively
defined datatypes of possibly mixed variance.  Thus, going beyond the
typical inductive types of type theory stemming from Dybjer's Inductive
Families~\cite{DybjerIF}.  It would be interesting to establish a formal
connection between these approaches so that they can inform each other;
in particular in the context of termination checking. %~\cite{AbelFoetus}. 

\hide{
Dually, we will also:
\begin{resenumerate}\setcounter{CC}{4}
\item
  Develop programming primitives corresponding to coinduction proof
  principles.
\end{resenumerate}
These have received less attention.
}

The above design considerations are to be shaped by the following maxim:
\begin{resenumerate}\setcounter{CC}{4}%{5}
\item
  Design language constructs with minimal type annotation supporting
  maximal type inference.
\end{resenumerate}
Indeed, this is the main open problem in dependently-typed programming language
theory.

\hide{
Finally, let us mention an intriguing possibility suggested by the
model-theoretic considerations of
Section\pref{MethodologyMathematicalUniversesParagraph}.  The indexing
structure in dependent type theory is discrete.  In many applications,
however, one is interested in indexed types that furthermore relate
structure on indices to structure on the associated indexed family of types.
The question arises as to how to:
\begin{resenumerate}\setcounter{CC}{5}%{6}
\item
  Build idioms or language abstractions for defining and then programming
  with indexed types equipped with internal varying structure.
\end{resenumerate}
A specific motivation for this comes from the possibility of directly
programming, as opposed to having to code, presheaf structure (\eg~as is
needed in Normalisation by Evaluation~\cite{FiorePPDP}).
}

\paragraph{Effects.}
\label{ProgrammingEffectsParagraph}

Our main aim here is to consider the model-theoretic investigations of
Section\pref{PolarisationParagraph} from a proof-theoretic viewpoint and
percolate this down to programming language theory.

A first main novelty in our approach is that the proof theory to be
considered is based on sequent calculus, rather than the traditional line
followed so far in programming language foundations based on sequent-style
natural deduction.

The proof-theoretic formalism that we will be adhering to is the
calculus~{\SysL} mentioned in Section\pref{SequentCalculiParagraph}.  A first
main question that arises is:
\begin{resenumerate}\setcounter{CC}{0}
\item 
  What is the programming paradigm stemming from sequent calculi in
  general, and the calculus~{\SysL} in particular?
\end{resenumerate}
For instance, 
\begin{resenumerate}\setcounter{CC}{1}
\item
  Does the inherent symmetry of the {\SysL}~formalism lead to a new programming
  style?

\item
  How can the close connection between the system~{\SysL} and abstract
  machines be exploited?  For instance, in formal compiler correctness.
\end{resenumerate}
These investigations will also require mathematical principles to be
developed.  Specifically, we will 
\begin{resenumerate}\setcounter{CC}{3}
\item
  Establish a Propositions-as-Types 
  %correspondence for sequent calculi through Category Theory (see
  %Figure\,\ref{ResearchAreas}) relating the system~{\SysL} to adjoints.
  trinity (see (\ref{PATtrinity})) between classical sequent calculi and
  computational languages with effects and rich type structure through
  adjoint interpretations.
\end{resenumerate}

As for polarisation, we believe that programmers will be able to intuitively
assimilate the eager {\vs}~lazy modes of computation and data structures
underlying it.  But, pragmatically,
\begin{resenumerate}\setcounter{CC}{4}
\item
  How will a programmer be able to easily code polarisation in a
  programming language?
\end{resenumerate}

Once the above is sorted, one can consider experimenting with programming
languages based on the various systems of
Section\pref{PolarisationParagraph} further enriched with computational
effects.  To this end, there is a large body of theoretical work on
effects that needs to be examined, evaluated, and reconsidered.  An
interesting stepping stone here is the recent work of Bauer and
Pretnar~\cite{BauerPretnar} on the experimental programming language Eff.  

Eff is based on the algebraic theory of effects and handlers of Plotkin
\etal~\cite{PlotkinPowerAlgOpsAndGenEffs,PlotkinPretnar};
but, for the pragmatics of supporting seemingly non-algebraic control
operators, it goes beyond algebraic models.  This is indeed so with
respect to (first-order) algebraic theories.  Interestingly, as it has
transpired in conversation between Fiore and Staton, the consideration of
the richer Second-Order Algebraic Theories of Fiore
\etal~\cite{FioreHurLMCS,FioreMahmoud} seems to enrich the class of
specifiable effects to incorporate aspects of control.  Thus, our research
into this topic will also
\begin{resenumerate}\setcounter{CC}{5}
\item
  Investigate Second-Order (and, when they are developed, Polymorphic)
  Algebraic Theories as a mathematical basis for an algebraic theory of
  effects encompassing control.
\end{resenumerate}
Targeted goals here will be program logics for effects and control, and formal
correctness proofs of (continuation and/or abstract machine) implementations
of programming languages with effects.

\hidenote{Dependent CBPV}

\paragraph{Metaprogramming.}
\label{MetaprogrammingParagraph}

{\color{red}

  generic programming

  reflection

  TDPE with sums: reflection with delimited control operators

  System $F_\omega^*$: Tillmann Rendel, Klaus Ostermann, Christian
  Hofer (2009),Typed Self-Representation

{\bf feel free to change any of this -- Tim}

In a metaprogramming system there are two languages of interest. The
object-language which is the object of study and the meta-language which
manipulates object programs as data. The purpose of the meta-language is
to define algorithms for the purpose of constructing or analyzing
object-programs. Metaprograms play a role in many kinds of systems

{\bf Generic programming.}
One style of generic programming defines a datastucture that reflects how
users define new types. For example, a value of type \verb+Def Tree+
describes how a new \verb+Tree+ datatype is defined. A generic program then
analyzes this structure of \verb+Def+ types to write code parametric over
any user defined type. For example a generic equality might have type
\verb+Def t -> t -> t -> Bool+ and a generic marshalling function
might have type \verb+Def t -> t -> List Bit+. 

{\bf Reflection.} One kind of reflection is where the object-language and
the meta-language coincide
}

\hidenote{System~{\SysL} internalisation of stacks?}

\hidenote{ssreflect?}

\clearpage
\setcounter{page}{1}
\chead{Part\,B1(b)}
\twocolumn[\begin{@twocolumnfalse}
	\hfill{\bfseries\Large 
    Curriculum Vitae
  }\hfill\null\\
\end{@twocolumnfalse}]

\subsection*{\S\enspace\thinspace Personal data}

\begin{myitemize}
\item
\textbf{\em Name}: 
\begin{tabular}[t]{l}
  Marcelo Pablo Fiore.
\end{tabular}

\item
\textbf{\em Nationality}: 
\begin{tabular}[t]{l}
  Argentinian/Italian.
\end{tabular}

\item
\textbf{\em Date of birth}: 
\begin{tabular}[t]{l}
  June 29, 1966.
\end{tabular}

\item
\textbf{\em Marital status}: 
\begin{tabular}[t]{l}
  Married, with two children (8\,\&\,3).
\end{tabular}

\item
\textbf{\em Address}: 
\begin{tabular}[t]{l}
  University of Cambridge, 
  Computer\\ Laboratory, 15 JJ Thomson
  Avenue,\\ Cambridge CB3 0FD, UK.
\end{tabular}

\item
\textbf{\em Web}: 
\begin{tabular}[t]{l}\small
  \url{www.cl.cam.ac.uk/~mpf23}
\end{tabular}

\item
\textbf{\em Email}: 
\begin{tabular}[t]{l}\small
  \texttt{Marcelo.Fiore@cl.cam.ac.uk}
\end{tabular}

\item
\textbf{\em Phone}: 
\begin{tabular}[t]{l}
  +44 (0)1223 334622.
\end{tabular}
\end{myitemize}

\subsection*{\S\enspace\thinspace University appointments}

\begin{myitemize}
\item
\textbf{\em Professor} in \emph{Mathematical Foundations of Computer
Science},  Computer Laboratory, University of Cambridge.  Since October
2011.

\item
\textbf{\em Reader} in \emph{Mathematical Foundations of Computer
Science}, Computer Laboratory, University of Cambridge.  October 2005 to
September 2011.

\item
\textbf{\em University Lecturer}, Computer Laboratory, University of
Cambridge.  October 2000 to September 2005.

\item
\textbf{\em Lecturer} in Computer Science, School of Cognitive and
Computing Sciences, University of Sussex.  April 1998 to September 2000. 

\item
\textbf{\em Research Fellow}, Laboratory for Foundations of Computer
Science, Department of Computer Science, University of Edinburgh.  January
1994 to March 1998.
\end{myitemize}

\subsection*{\S\enspace\thinspace University education}

\begin{myitemize}
\item
{\em Postgraduate}: \textbf{\em PhD in Computer Science}, University of
Edinburgh, June 1994.  \emph{Thesis}: Axiomatic Domain Theory in
Categories of Partial Maps.  \emph{Supervisor}: Prof Gordon Plotkin
(University of Edinburgh).  \emph{Examiner}: Prof Dana Scott (Carnegie
Mellon University)

\item
{\em Graduate}: \textbf{\em Licentiate in Informatics}, Escuela Superior
Latino Americana de Inform\'atica, Universidad Nacional de Luj\'an, Buenos
Aires (Argentina), December 1989.  \emph{Thesis}: On-line Algorithms for
Traversal of Dynamic Directed Hypergraphs with Application to
Satisfiability of Horn Formulae.  \emph{Supervisor}: Prof Giorgio
Ausiello, Uniersit\`a di Roma ``La Sapienze'', Italy.
\end{myitemize}

\subsection*{\S\enspace\thinspace Awards}

\begin{myitemize}
\item
\textbf{\em 10-Year Most Influential PPDP Paper Award} for the article
\emph{Semantic Analysis of Normalisation by Evaluation for Typed Lambda
Calculus} in the 2002 International Symposium on Principles and Practice
of Declarative Programming.

\item
\textbf{\em Distinguished Dissertation in Computer Science} for 1995.
\emph{Axiomatic Domain Theory in Categories of Partial Maps}.  Doctoral
thesis selected by the Conference of Professors and Heads of Computing in
conjunction with the British Computer Society.
\end{myitemize}

\subsection*{\S\enspace\thinspace Postgraduate supervision}
\vspace*{-1.9mm}
\begin{myitemize}
\item[]\small\ at the Computer Laboratory, University of Cambridge
\end{myitemize}

\begin{myitemize}
\item
  Marco Ferreira Devesas Campos.  Second year of PhD in Computer Science.

\item
  Ola Mahmoud, 2011 PhD in Computer Science.  \emph{Thesis}: Second-Order
  Algebraic Theories.

\item
  Marco Ferreira Devesas Campos, 2011 MPhil in Advanced Computer Science.
  \emph{Research essay}: Generalizing Bigraphs to DAG-like Place Graphs.

\item
  Eirik Tsarpalis, 2010 Certificate of Postgraduate Studies.

\item
  Maciej Wos, 2010 MPhil in Advanced Computer Science. \emph{Research essay}:
  Generic Programming with Fixed-Points for Nested Datatypes.

\item
  Chung-Kil Hur, 2010 PhD in Computer Science.  \emph{Thesis}: Categorical
  Equational Systems: Algebraic Models and Equational Reasoning.

\item
  Samuel Staton, 2007 PhD in Computer Science.  \emph{Thesis}:
  Name-Passing Process Calculi: Operational Models and Structural
  Operational Semantics.
\end{myitemize}

\subsection*{\S\enspace\thinspace Mentoring}
\vspace*{-2mm}
\begin{myitemize}
\item[]\small\ at the Computer Laboratory, University of Cambridge
\end{myitemize}

\begin{myitemize}
\item
Bartosz Klin.  EPSRC Postdoctoral Research Fellow in Theoretical Computer
Science.  October 2008 to September 2011.  

\item
Johan Glimming.  Visiting Research Fellow funded by the Swedish Research
Council VR.  October 2008 to September 2010.  

\item
Samuel Staton.  EPSRC Postdoctoral Research Fellow in Theoretical Computer
Science.  June 2007 to May 2010.  
\end{myitemize}

\subsection*{\S\enspace\thinspace Examining}

\begin{myitemize}
\item
\textbf{\em External} committee member for the PhD theses of: 
%
Wei Chen (University of Nottingham, 2012);
%
Julianna Zsido (Universit\'e de Nice Sophia Antipolis, 2010); 
%
Stephane Gimenez (Universit\'e Paris Diderot - Paris~7, 2009); 
%
Vincenzo Ciancia (Dipartamento di Informatica, Universit\`a di Pisa,
2008); 
%
Joachim de Lataillade (Universit\'e Paris Diderot - Paris~7, 2007); 
%
Emmanuel Beffara (Universit\'e Paris Diderot - Paris~7, 2005); 
%
Pierre Hyvernat (Institut Math\'ematique de Luminy, 2005);
%
Bartosz Klin (Aarhus University, 2004);
%
Vincent Balat~(Universit\'e Paris Diderot - Paris~7, 2002);
%
Krzysztof Worytkiewicz~(\'Ecole Polytechnique F\'ed\'erale de Lausanne,
2000).

\item
\textbf{\em Internal} committee member for the PhD theses of: 
%
Bjarki Holm (2011); 
%
Ranald Clouston (2009); 
%
Paul Hunter (2007); 
%
Lucy Brace-Evans (2007).

\item
Committee member for the Professorship habilitation of
%
Davide Sangiorgi~(INRIA Sophia Antipolis, 2002).

\item
Committee member for the MPhil thesis of
%
Henrik Enstr{\o}m (Aarhus University, 1998).

\item
Computer Science Tripos Examiner for Part IB, Part II, Part II (General),
and Diploma in 2006/07, 2007/08, and 2008/09 (Chair).  
\pagebreak
Computer Laboratory, University of Cambridge.  
\end{myitemize}

\vspace*{-1.25mm}
\subsection*{\S\enspace\thinspace Funding}
\vspace*{-1mm}
\begin{myitemize}
\item
  EPSRC grant submission (under consideration).
  \emph{Efficient Extraction of Feasible Programs}.  \textbf{\em Principal
    Investigator}, with co-investigators Timothy Griffin and Glynn Winskel
  (University of Cambridge).  Joint submission with Dr Mart\'{\i}n
  Escard\'o (University of Birmingham), Paulo Oliva (Queen Mary University
  of London), Ulrich Berger and Monika Seisenberger (Swansea University).
  Travel funding for 24 months.

\item
  \textbf{\em Visiting grant} from the Gunma University Foundation for
  Science and Technology.  Department of Computer Science, Gunma
  University.  April 3--12, 2012.

\item
  \textbf{\em Research in Paris grant} from the \emph{programme d'accueil
    des chercheurs \'etrangers de la Ville de Paris}, Laboratoire PPS,
  Universit\'e Paris Diderot - Paris~7.  October--December 2011.

\item
  \emph{Domain theory for concurrency---New categorical foundations}.
  \textbf{\em Co-investigator} of Glynn Winskel.  EPSRC Grant
  GR/T22049/01.  July 2005 to June 2008.

\item
  \emph{Mathematical models for functional and concurrent computation}.
  \textbf{\em EPSRC Advanced Research Fellowship}. October 2000 to
  September 2005.  
\end{myitemize}

\vspace*{-1.25mm}
\subsection*{\S\enspace\thinspace Invited research visits}
\vspace*{-1mm}

\begin{myitemize}
\item
  Short-Term Scholar invitation to the Univalent Foundations of Mathematics
  Program at the School of Mathematics, Institute for Advanced Study,
  Princeton. 

\item
  Assistant Prof Makoto Hamana at the Department of Computer Science,
  Gunma University.  In April 2012.

\item
  Prof Peter Dybjer at the Department of Computer Science and Engineering,
  Chalmers University of Technology.  In September 2008. 

\item
  Prof Anders Kock at the Department of Mathematical Sciences, Aarhus
  University.  In August 2007. 

\item
  Prof Mart\'{\i}n Abadi at the Department of Computer Science, University
  of California at Santa Cruz.  November 2003. 

\item
  Dr Vincent Danos at Laboratoire PPS, Universit\'e Paris Diderot -
  Paris~7.  June 2002.

\item
  Dr Paul-Andr\'e Melli\`es at Laboratoire PPS, Universit\'e Paris Diderot
  - Paris~7.  June 2001.

\item
  Mart\'{i}n Abadi at Bell Labs Research, Lucent Technologies, Palo Alto.
  In October--November 1999. 

\item
  Prof Gordon Plotkin at LFCS, Department of Computer Science, University
  of Edinburgh.  In November--December 1998.

\item
  Prof Marta Bunge at Department of Mathematics and Statistics, McGill
  University.  September and October 1998.

\item
  Prof Glynn Winskel at Department of Computer Science, Aarhus University,
  Denmark.  In August--September 1997.
\end{myitemize}

\vspace*{-1mm}
\subsection*{\S\enspace\thinspace Hosted research visitors}
\vspace*{-2mm}
\begin{myitemize}
\item[]\small\ at the Computer Laboratory, University of Cambridge
\end{myitemize}
\pagebreak

\noindent
PhD Student Ki Yung Ahn 
(Portland State University) in September 2012; 
%
Dr Michael Warren (IAS School of Mathematics, Princeton) in May 2012; 
%
Associate Prof Iliano Cervesato (Carnegie Mellon University - Qatar
Campus) in July 2011; 
%
PhD Student Guillaume Munch-Maccagnoni, (Universit\'e Paris Diderot -
Paris~7) MarchMay 2011 and in October 2012;
%
Prof Pierre-Louis Curien (Universit\'e Paris Diderot - Paris~7)
April--June 2009, April--June 2010, and in October 2012; 
%
Assistant Prof Makoto Hamana (Gunma University) in September 2009, March
2010 and October 2012;
%
Ichiro Hasuo (RIMS, Kyoto University) in November 2009;
%
Mat\'{\i}as Menni (Lifia, Universidad Nacional de La Plata) in July 2007
and April 2009;
%
Prof Franck van Breugel (York University, Toronto) sabbatical leave in
2004--05.

\vspace*{-0mm}
\subsection*{\S\enspace\thinspace Peer-reviewing}

\begin{myitemize}
\item
  \textbf{\em Programme committees}: CT 2013, ICALP 2013, Wollic 2012,
  MFPS XXVIII, MFCS 2012, URC 2010, SOS 2009, FOSSACS 2009, CSL 08, ICALP
  2008, MFPS XXIII (Chair), LICS 2007, SOFSEM 07, MSFP06, CT 2006, MFPS
  XXI, ICALP05, TCS 2004, CTCS 2004, CMCIM03, WAIT2003 (Co-chair),
  LICS2003, FOSSACS2002, MFPS XVII, LICS2000, WAIT98.

\item
  Reviewer for national and international conferences and journals.

\item
  Reviewer for the Uruguayan National Agency for Research and Innovation
  (ANII).

\item
  EPSRC Computer Science Panel Member.  February~1, 2005.
\end{myitemize}

\vspace*{-0mm}
\subsection*{\S\enspace\thinspace Memberships}

\begin{myitemize}
\item
  Fellow of \textbf{\em Christ's College}, University of Cambridge.  Since
  October 2001.

\item
  \textbf{\em EPSRC Peer Review College} member. October 2000 to September
  2010.

\item
  Editorial Board member of \textbf{\em Applied Categorical Structures}.
  Since October 2004.

\item
  Editorial Board member of the \textbf{\em Bulletin of Symbolic Logic}
  reviews for the \emph{Association for Symbolic Logic}.  January 2009 to
  December 2011.
\end{myitemize}

\vspace*{-0mm}
\subsection*{\S\enspace\thinspace Teaching}
\begin{myitemize}
\item Sundry undergraduate and postgraduate lecturing at the Computer
  Laboratory, University of Cambridge.  
\item Director of Studies for Computer Science in Christ's College
  Cambridge.  October 2001 to September 2011.  
\item Sundry undergraduate lecturing and tutorials at the School of
  Cognitive and Computing Sciences, University of Sussex. 
\item Sundry postgraduate lecturing at the Laboratory for Foundations of
  Computer Science, University of Edinburgh.
\end{myitemize}

\subsection*{\S\enspace\thinspace Administration}

\begin{myitemize}
\item
Sundry committees at the University of Cambridge and Christ's College
Cambridge.
\end{myitemize}

\clearpage
\setcounter{page}{1}
\chead{Part\,B1(c)}
\twocolumn[\begin{@twocolumnfalse}
	\hfill{\bfseries\Large 
    10-Year Track Record
  }\hfill\null\\
\end{@twocolumnfalse}]

%\subsection*{\S\enspace\thinspace Personal statement}

My research is in theoretical computer science, and by its own nature develops
in the context of close specialised collaboration.  My work stems from the
interaction between two approaches: abstract and concrete. At the abstract
level, I build mathematical theories. At the concrete level, theories are
applied to specific problems. This duality has yielded deep results and
provided research breadth, giving fundamental contributions to a variety of
fields and finding connections among them; \eg~domain theory, category theory,
concurrency theory, sheaf theory, type theory, logic, algebra.  Indeed,
note that I publish in computer science and mathematics venues, and I have
been invited speaker at both computer science and mathematics meetings.
My \mbox{h-index} is 21, according to Google scholar.  

\subsection*{\S\enspace\thinspace Award}

\begin{mybigitemize}
\item%[$\star$]
\textbf{\em 10-Year Most Influential PPDP Paper Award} for the article
\emph{Semantic Analysis of Normalisation by Evaluation for Typed Lambda
Calculus} in the 2002 International Symposium on Principles and Practice
of Declarative Programming.
\end{mybigitemize}

\subsection*{\S\enspace\thinspace Selected %peer-reviewed 
  publications}
\vspace*{-1.125mm}
\begin{myitemize}
\item[]\small\ 
  Number of citations provided according to Google
  \\\mbox{} 
  scholar.  Conference publications further selected for
  \\\mbox{} 
  journal proceedings have been circled.  Representative 
  \\\mbox{} 
  publications have been starred.
\end{myitemize}

\paragraph*{Research monograph}

\begin{mybigitemize}
\item[$\star$] 
  M.\,Fiore (2004). \emph{Axiomatic Domain Theory in Categories of Partial
    Maps}.  Distinguished Dissertations in Computer Science, No.\,14.
  Paperback Edition, Cambridge University
  Press.\mbox{}\hfill{\small[111~citations]}
\end{mybigitemize}

\paragraph*{Journals}

\begin{mybigitemize}
\item[$\obullet$]
  M.\,Fiore and C.-K.\,Hur (2011).  On the mathematical synthesis of
  equational logics.   In \emph{Selected Papers of the Conference
    ``Typed Lambda Calculi and Applications 2009''}.  \emph{Logical
    Methods in Computer Science}, Volume 7, ISSUE 3, PAPER
  12.\\\mbox{}\hfill{\small[19 (conference version) citations]}

%\item
%  M.\,Fiore and T.\,Leinster (2010).  An abstract characterization of
%  Thompsons group F.  \emph{Semigroup Forum}, Volume 80, Number 2,
%  325--340.\\\mbox{}\hfill{\small[1~citation]}

\item[$\star$]
  M.\,Fiore and S.\,Staton (2009).  A congruence rule format for
  name-passing process calculi.  In \emph{Special Issue on Structural
    Operational Semantics (SOS)}. \emph{Information and Computation},
  207(2):209-236.\\\mbox{}\hfill{\small[26 (conference version) + 5~citations]}

\item[$\obullet$]
  M.\,Fiore and C.-K.\,Hur (2008).  On the construction of free
  algebras for equational systems.  In \emph{Special Issue for the
    Thirty-fourth International Colloquium on Automata, Languages and
    Programming (ICALP'07).  Theoretical Computer Science},
  410:1704--1729.\\\mbox{}\hfill{\small[15 (conference version) + 11 citations]}

\item[$\star$]
  M.\,Fiore, N.\,Gambino, M.\,Hyland, and G.\,Winskel (2008).  The
  cartesian closed bicategory of generalised species of structures. 
  \emph{Journal of the London Mathematical Society},
  77:203-220.\mbox{}\hfill{\small[19~citations]}

%\item
%  G.L.\,Cattani and M.\,Fiore (2007).  The bicategory-theoretic
%  solution of recursive domain equations.  In \emph{Computation,
%    Meaning, and Logic: Articles dedicated to Gordon Plotkin, Electronic Notes
%    in Theoretical Computer Science}, Volume 172,
%  pp\,203--222.\\\mbox{}\hfill{\small[0~citations]}

\item[$\obullet$]
  M.\,Fiore and S.\,Staton (2006).  
  Comparing operational models of
  name-passing process calculi.  \emph{Information and Computation},
  Vol\,204, Issue 4, pp\,524--560.\\\mbox{}\hfill{\small[15 (conference version)
    + 29~citations]}

\item[$\obullet$]
  M.\,Fiore, R.\,Di Cosmo, and V.\,Balat (2006).  Remarks on
  isomorphisms in typed lambda calculi with empty and sum types. 
  \emph{Annals of Pure and Applied Logic}, Vol\,141, Issues 1--2,
  pp\,35--50.\\\mbox{}\hfill{\small[19 (conference version) + 11 citations]}

%\item
%  M.\,Fiore and M.\,Menni (2005).  Reflective Kleisli subcategories
%  of the category of Eilenberg-Moore algebras for factorization monads.
%   In \emph{Proceedings of the International Category Theory
%    Conference (CT2004), Theory and Applications of Categories}, Vol\,15,
%  No\,2, pp\,40--65.\\\mbox{}\hfill{\small[3~citations]}

\item
  M.\,Fiore and T.\,Leinster (2005).  Objects of categories as
  complex numbers.  \emph{Advances in Mathematics},
  190(2):264--277.\mbox{}\hfill{\small[11~citations]} 

\item
  M.\,Fiore and T.\,Leinster (2004).  An Objective Representation of
  the Gaussian Integers.  \emph{Journal of Symbolic Computation},
  37(6): 707--716.\mbox{}\hfill{\small[5~citations]} 

\item[$\ostar$]
  M.\,Fiore, E.\,Moggi, and D.\,Sangiorgi (2002).  A fully-abstract
  model for the pi-calculus.  Information and Computation,
  179:76--117.\\\mbox{}\hfill{\small[95 (conference version) + 30 citations]}
\end{mybigitemize}

\paragraph*{Conferences}

\begin{mybigitemize}
\item[$\star$]
  M.\,Fiore (2012).  Discrete Generalised Polynomial Functors.  In
  \emph{Automata, Languages, and Programming Conference (ICALP'12)}.
  Volume 7392 of Lecture Notes in Computer Science,
  pp.\,214--226.\\\mbox{}\hfill{\small[0~citations]}

%\item
%  M.\,Hamana and M.\,Fiore (2011).  A foundation for GADTs and
%  Inductive Families: Dependent polynomial functor approach.  In
%  \emph{ACM SIGPLAN Seventh Workshop on Generic Programming (WGP'11)},
%  pp.\,59--70.  ACM Press.\\\mbox{}\hfill{\small[0~citations]}

\item[$\obullet$]
  M.\,Fiore and O.\,Mahmoud (2010).  Second-order algebraic theories.
   In \emph{Proceedings of the Thirty-fifth International Symposium
    on Mathematical Foundations of Computer Science (MFCS'10)}.  
  Volume 6281 of Lecture Notes in Computer Science,
  pp\,368--380.\mbox{}\hfill{\small[7~citations]}

\item[$\ostar$]
  M.\,Fiore and C.-K.\,Hur (2010).  Second-order equational logic. In
  \emph{Proceedings of the Nineteenth EACSL Annual Conference on Computer
    Science Logic (CSL'10)}.  Volume 6247 of Lecture Notes in
  Computer Science, pp\,320--335.\mbox{}\hfill{\small[11~citations]}

%\item
%  M.\,Fiore and S.\,Staton (2010).  Positive structural operational
%  semantics and monotone distributive laws.  \emph{In Short
%    Contributions for the Tenth International Workshop on Coalgebraic Methods
%    in Computer Science (CMCS 2010)}.  CWI Technical report SEN-1004,
%  pp\,8--9.\\\mbox{}\hfill{\small[0~citations]}

%%\item
%%  M.\,Fiore and C.-K.\,Hur (2008).   Term equational systems and
%%  logics.  In \emph{Proceedings of the Twenty-fourth Conference on
%%    the Mathematical Foundations of Programming Semantics (MFPS XXIV)}.
%%   Electronic Notes in Theoretical Computer Science, volume 218,
%%  pp\,171--192.\\\mbox{}\hfill{\small[19~citations]}

\item[$\star$]
  M.\,Fiore (2008).  Second-order and dependently-sorted abstract
  syntax.  In \emph{Logic in Computer Science Conference (LICS'08)},
  pp.\,57--68.  IEEE, Computer Society
  Press.\mbox{}\hfill{\small[22~citations]}

%%\item
%%  M.\,Fiore and C.-K.\,Hur (2007).  Equational systems and free
%%  constructions.  In \emph{International Colloquium on Automata,
%%    Language and Programming (ICALP 2007)}. Volume 4596 of Lecture Notes in
%%  Computer Science, pp\,607--619.\\\mbox{}\hfill{\small[15~citations]}

\item[$\obullet$]
  M.\,Fiore (2007).  Differential structure in models of
  multiplicative biadditive intuitionistic linear logic.  In
  \emph{Typed Lambda Calculi and Applications (TLCA'07)}.  Volume
  4583 of Lecture Notes in Computer Science,
  pp\,163--177.\mbox{}\hfill{\small[14~citations]}

%%\item
%%  M.\,Fiore and S.\,Staton (2006).  A congruence rule format for
%%  name-passing process calculi from mathematical structural operational
%%  semantics.  In \emph{Twenty-first Logic in Computer Science
%%    Conference (LICS'06)}, pp\,49--58.  IEEE, Computer Society
%%  Press.\\\mbox{}\hfill{\small[26~citations]}

%%\item
%%  M.\,Fiore and S.\,Staton (2004).  Comparing operational models of
%%  name-passing process calculi.  In \emph{Proceedings of the 7th
%%  Coalgebraic Methods in Computer Science Workshop (CMCS'04)}. 
%%  Volume 106 of Electronic Notes in Theoretical Computer Science, pages
%%  91--104.  Elsevier.\mbox{}\hfill{\small[29 citations]}

\item
  M.\,Fiore (2004).  Isomorphisms of generic recursive polynomial
  types.   In \emph{Thirty-first Symposium on Principles of
    Programming Languages (POPL'04)}, pp\,64--76.  ACM
  Press.\mbox{}\hfill{\small[22~citations]}

\item[$\star$]
  V.\,Balat, R.\,Di Cosmo, and M.\,Fiore (2004).  Extensional
  normalisation and type-directed partial evaluation for typed lambda
  calculus with sums.   
  In \emph{Thirty-first Symposium on Principles
    of Programming Languages (POPL'04)}, pp\,64--76.  ACM
  Press.\mbox{}\hfill{\small[47~citations]}

\item[$\star$]
  M.\,Fiore (2002).  Semantic analysis of normalisation by evaluation
  for typed lambda calculus.  In \emph{Fourth Principles and Practice
    of Declarative Programming Conference (PPDP'02)}, pages 26--37. 
  ACM Press.\mbox{}\hfill{\small[45~citations]}

%%\item
%%  M.\,Fiore, R.\,Di Cosmo, and V.\,Balat (2002).  Remarks on
%%  isomorphisms in typed lambda calculi with empty and sum types.  In
%%  \emph{Seventeenth Logic in Computer Science Conference (LICS'02)},
%%  pp\,147--156.   IEEE, Computer Society
%%  Press.\\\mbox{}\hfill{\small[19~citations]}
\end{mybigitemize}

\paragraph*{Invited}

\begin{mybigitemize}
\item[$\star$]
  M.\,Fiore (2005).  Mathematical models of computational and combinatorial
  structures.  {\em Foundations of Software Science and Computation Structures
    (FOSSACS'05)}.  Volume 3441 of Lecture Notes in Computer Science,
  pp\,25--46.\mbox{}\hfill{\small[18~citations]}
\end{mybigitemize}

\subsection*{\S\enspace\thinspace Invited addresses}
\vspace*{-1.125mm}
\begin{myitemize}
\item[]\small\ 
  Salient invitations have been starred.
\end{myitemize}


\begin{mybigitemize}
\item[$\star$]
  Indexing Structures in Programming Language Semantics and Design.
  \emph{Fourteenth International Symposium on Principles and Practice of
    Declarative Programming~(PPDP'12)}, 
  %10-Year Most Influential PPDP Paper Award talk, 
  Leuven (Belgium), September~2012.

\item
  Lie in Logic.  \emph{S\'{e}minaire CHoCoLa}, \'Ecole Normale Sup\'erieure de
  Lyon (France), May~2012.

\item[$\star$]
  Second-order algebra and generalised polynomial functors.  \emph{Logic and
    Interactions 2012}, Centre International de Rencontres Math\'{e}matiques,
  Luminy (France), February~2012.

\item
  Estructuras matem\'{a}ticas en lenguajes de programaci\'{o}n.
  \emph{French-Argentinean Laboratoire Internationale Associ\'e LIA INFINIS},
  %inaugural meeting, 
  Departamento de Computaci\'{o}n, Facultad de Ciencias Exactas y Naturales -
  Universidad de Buenos Aires (Argentina), December~2011.

\item
  On Higher-Order Algebra.   \emph{Third Scottish Category Theory Seminar},
  University of Strathclyde, Glasgow (Scotland), December~2010.

\item
  Algebraic simple type theory.  \emph{Curry-Howard and Concurrency Theory
    (CHoCo) S\'eminaire}, l'Ecole Normale Sup\'erieure de Lyon (France),
  December 2009.

\item[$\star$]
  Mathematical synthesis of equational deduction systems.  \emph{Ninth
    International Conference on Typed Lambda Calculi and Applications
    (TLCA'09)}, Brasilia (Brazil), June 2009.  

\item
  Algebraic Type Theory.  \emph{Eighty-eighth Peripatetic Seminar on Sheaves
    and Logic (88th PSSL), Celebrating the 60th birthdays of Martin Hyland and
    Peter Johnstone}, University of Cambridge (UK), April~2009.

\item[$\star$]
  Algebraic theories and equational logics.  \emph{Twenty-fourth Conference on
    the Mathematical Foundations of Programming Semantics (MFPS~XXIV)},
  University of Pennsylvania~(USA), May~2008.  

\item
  Second-order and dependently-sorted algebraic theories.  \emph{Workshop on
    Categorical and Homotopical Structures in Proof Theory}, Centre de Recerca
  Matem\`atica, %~(CRM), 
  Barcelona~(Spain), February 2008.

\item
  An axiomatics and a combinatorial model of creation/annihilation operators
  and differential structure.  \emph{Categorical Quantum Logic
    Workshop (CQL)}, %Computing Laboratory, 
  University of Oxford (UK), August 2007.

\item
  A mathematical theory of substitution and its applications to syntax and
  semantics.  \emph{Coalgebraic Logic Workshop (CoL)}, %Computing Laboratory,
  University of Oxford (UK), August 2007.

\item[$\star$]
  Towards a mathematical theory of substitution.  \emph{Annual International
    Conference on Category Theory (CT'07)}, Carvoeiro, Algarve (Portugal),
  June 2007.
 
\item[$\star$]
  A mathematical theory of substitution and its applications to syntax and
  semantics.  \emph{Workshop on Mathematical Theories of Abstraction,
    Substitution and Naming in Computer Science}, International Centre for
  Mathematical Sciences, %(ICMS), 
  Edinburgh (Scotland), May~2007.
 
\item[$\star$]
  Analytic functors and domain theory.  \emph{Symposium for Gordon Plotkin},
  Laboratory for Foundations of Computer Science, %~(LFCS), 
  University of Edinburgh (Scotland), September~2006.
 
\item[$\star$]
  On the structure of substitution. \emph{Twenty-second Mathematical
    Foundations of Programming Semantics Conference (MFPS XXII)}, %DISI,
  University of Genova (Italy), May~2006.
 
\item
  Isomorphisms on generic mutually recursive polynomial types.  \emph{Second
    International Workshop on Isomorphisms of Types}, Universit\'e de Toulouse
  (France), October 2005.

\item[$\star$]
  Mathematical models of computational and combinatorial structures.  
  \emph{Foundations of Software Science and Computation Structures
  (FOSSACS'05)} for European Joint Conferences on Theory and Practice of
  Software (ETAPS'05), Edinburgh~(UK), April 2005.  

\item[$\star$]
  Computational semantics.  \emph{Winter School on Semantics and
    Applications}, Instituto de Computaci\'on, Facultad de Ingenier\'{\i}a,
  Universidad de la Rep\'ublica, Montevideo~(Uruguay), July 2003.

\item
  A semantic framework for name and value passing process calculi.
  \emph{Workshop on Concurrency and Mobility, Logic and Foundations of
    Computation}, Fields Institute Summer School, University of Ottawa
  (Canada), June 2003.

\item[$\star$]
  Imaginary types.  \emph{Nineteenth Conference on the Mathematical
    Foundations of Programming Semantics (MFPS~XIX)}, Montr\'eal~(Canada),
  March~2003.

\item
  Foundational theories of combinatorial type structures.  \emph{Birmingham,
    Leicester, and Nottingham Christmas Theory Day}, Department of Computer
  Science, University of Birmingham~(UK), December~2002.
\end{mybigitemize}
\vspace*{-6mm}
\end{document}
