\documentclass[11pt,twocolumn]{article}

\usepackage{amsfonts,euler}
\usepackage{MnSymbol}

\usepackage[all]{xy}
\UseComputerModernTips
\xyoption{knot}

\usepackage[a4paper,includehead,includefoot,headheight=13.6pt,
  headsep=4mm,top=1cm,bottom=1cm,left=2cm,right=2cm]{geometry}

\usepackage{float}
\floatstyle{ruled}
\restylefloat{figure}

\usepackage{dashrule}

\usepackage{multirow}

\usepackage{color}
\definecolor{grey}{gray}{.85}

\setcounter{secnumdepth}{4}
\newcommand{\myparagraph}[1]{\paragraph*{\em #1}}
\newcommand{\mynumparagraph}[1]{\paragraph{\em #1}}

%%% BEGIN macros
\newenvironment{myitemize}
  {\begin{list}{$\bullet$}
  {\setlength{\topsep}{2pt}
   \setlength{\partopsep}{2pt}
   \setlength{\itemsep}{2.5pt}
   \setlength{\parsep}{2.5pt}
   \setlength{\leftmargin}{1em}
   \setlength{\labelwidth}{.5em}}}
  {\end{list}}
\newenvironment{btritemize}
  {\begin{list}{\btr}
  {\setlength{\topsep}{2pt}
   \setlength{\partopsep}{2pt}
   \setlength{\itemsep}{2.5pt}
   \setlength{\parsep}{2.5pt}
   \setlength{\leftmargin}{1em}
   \setlength{\labelwidth}{.5em}}}
  {\end{list}}
\newcounter{CC}
\newenvironment{resenumerate}
  {\begin{list}{[\textbf{\arabic{CC}]}}
  {\usecounter{CC}
   \setlength{\topsep}{2pt}
   \setlength{\partopsep}{2pt}
   \setlength{\itemsep}{2.5pt}
   \setlength{\parsep}{2.5pt}
   \setlength{\leftmargin}{1.65em}
   \setlength{\labelwidth}{1.15em}
 }}
  {\end{list}}

\newcommand{\mysf}{\small\sf}
\newcommand{\mytextsf}[1]{\textsf{\small #1}}
\newcommand{\erc}{{\small\sf MaStrPLan}}
\newcommand{\ERC}{Mathematical Structures for Type Theories,\\[-.5mm] Logical
  Systems, and Programming Languages}

\newcommand{\hide}[1]{}
\newcommand{\hidefootnote}[2]{}
\newcommand{\note}[1]{\begin{quote}{\color{blue}$\leadsto$ \bf\em
      #1}\end{quote}}
\newcommand{\hidenote}{\hide}

\newcommand{\pref}[1]{\,(\ref{#1})}
\newcommand{\itemref}[1]{\textbf{[\ref{#1}]}}

\newcommand{\eg}{\emph{eg.}}
\newcommand{\Eg}{\emph{Eg.}}
\newcommand{\vs}{\emph{vs.}}
\newcommand{\ie}{\emph{ie.}}
\newcommand{\viz}{\emph{viz.}}
\newcommand{\etal}{\emph{et al.}}

\newcommand{\lcalculus}{\mbox{$\lambda$-calculus}}
\newcommand{\SystemL}{\mbox{System~$L$}}
\newcommand{\SystemF}{\mbox{System~$F$}}
\newcommand{\SystemFi}{\mbox{System~$F_i$}}
\newcommand{\SystemFomega}{\mbox{System~$F_\omega$}}
\newcommand{\LC}{\mbox{$LC$}}

\newcommand{\btr}{$\blacktriangleright$}

\newcommand{\reqpsize}{8.113395cm}%{\columnwidth}

\newcommand{\req}[2]{\begin{center}\colorbox{grey}{\begin{minipage}{\reqpsize} 
  \mytextsf{Research question}\hfill$^{\mbox{\scriptsize #1 }}$\\[-5.5mm]
  \begin{btritemize}
  \item #2
  \end{btritemize}
\end{minipage}}\end{center}}

\newcommand{\reqs}[2]{\begin{center}\colorbox{grey}{\begin{minipage}{\reqpsize}
  \mytextsf{Research questions}\hfill$^{\mbox{\scriptsize #1 }}$\\[-5.5mm]
  \begin{btritemize}
  \item #2
  \end{btritemize}
\end{minipage}}\end{center}}
\newcommand{\rep}[2]{\begin{center}\colorbox{grey}{\begin{minipage}{\reqpsize}
  \mytextsf{Research pathway}\hfill$^{\mbox{\scriptsize #1 }}$\\[-5.5mm]
  \begin{btritemize}
  \item #2
  \end{btritemize}
\end{minipage}}\end{center}}

\newcommand{\Set}{{\boldsymbol{\mathscr S}}}
\newcommand{\scat}[1]{\mathbb{#1}}
\newcommand{\cat}[1]{\mathscr{#1}}
\newcommand{\op}{\circ}
\newcommand{\Id}{\mathrm{Id}}
\newcommand{\Di}{\mathrm{Di}}
%%% END macros

\usepackage{fancyhdr}
\pagestyle{fancy}
\lhead{Marcelo Fiore}
\chead{Part\,B2}
\rhead{\erc}
\lfoot{}
\cfoot{\thepage}
\rfoot{}

\usepackage[compact]{titlesec}

\usepackage{compactbib}

\usepackage{setspace}
\setstretch{.9}

\usepackage{times}

\makeatletter
\renewcommand\@biblabel[1]{#1}
\makeatother

\renewcommand{\thesection}{\alph{section}}
\renewcommand{\thesubsection}{\alph{section}-\arabic{subsection}}
\renewcommand{\theparagraph}{\alph{section}-\arabic{subsection}.\roman{paragraph}}

\def\contentsname{\large Contents\\[-7.5mm]}
\usepackage{minitoc}
\tightmtctrue

\usepackage[draft]{hyperref}
\hypersetup{pdftitle={\ERC},pdfauthor={Marcelo Fiore}}

\begin{document}

\twocolumn[\begin{@twocolumnfalse}
	\hfill{\bfseries\LARGE The Scientific Proposal}\hfill\null\\
\end{@twocolumnfalse}]

%{\small\tableofcontents}

\section{\underline{State of the art and ob}j\underline{ectives}}
\label{StateOfTheArtSection}

We present the general scientific background. %of the research proposal.
This develops in the context of both %theoretical 
computer science and %pure 
mathematics, encompassing four main areas of research: category theory,
mathematical logic, programming theory, and type theory.  Each of these is
regarded as a complimentary part of the others, forming a unified whole
through which the subject is to be investigated.  Figure\,\ref{ResearchAreas}
gives a schematic view of the areas together with their interactions.
\begin{figure}[h]
\caption{Research areas and interactions.}
\vspace*{2mm}
\begin{center}
\hspace*{.5mm}
\xymatrix@R=25pt@C=12.5pt{
& 
\raisebox{7.5mm}{\fbox{\txt{\small Mathematical\\\small Logic}}}
\ar@/_1em/@{<->}[ddl]<-1em>|-
  {\txt{\scriptsize Propositions\\\scriptsize as Types}}
\ar@/^1em/@{<->}[ddr]<1em>|-
  {\txt{\scriptsize Proof\\\scriptsize Theory}} 
& 
\\
& 
\ar@{<->}[dl]|-
  {\txt{\scriptsize Internal\\\scriptsize Languages}}
\ar@{<->}[dr]|-
  {\txt{\scriptsize Denotational\\\scriptsize Semantics}} 
\ar@{<->}[u]|-
  {\txt{\scriptsize Model Theory}}
\fbox{\txt{\small Category\\\small Theory}}
& 
\\
  \fbox{\txt{\small Type\\\quad\small Theory\quad\null}}
\ar@/_1em/@{<->}[rr]|-
  {\txt{\scriptsize Programming\\\scriptsize Languages}}
& & 
\fbox{\txt{\small Programming\\\small Theory}}
}
\end{center}
\vspace*{-2mm}
\label{ResearchAreas}
\end{figure}
We contend that an approach neglecting any one of them is to the detriment of
the others, missing out the depth and richness of the subject and, with this,
opportunities for research and development.  

In Section\pref{Origins} we outline the origins and influences that lead
to this conception. %, roughly expanding through 1935 to 1985.
Section\pref{StateOfTheArtSubsection} follows with the current state of the
field and raising questions for research.  %Having set the scene,
Section\pref{ObjectivesSubsection} gives the research objectives that are to
be pursued for advancing current knowledge. %in the area.  
It also presents the {\erc} team that will achieve this.  Both these points
are respectively expounded upon in
Sections\,(\ref{MethodologySection}\,\&\,\ref{ResourcesSection}).

\subsection{Origins and influences}
\label{Origins}

\paragraph{Logic and computation.}

Computer science was born as a branch of mathematics, specifically
mathematical logic, before electronic computers were built.  Its inception
was the 1928 Entscheidungsproblem (decision problem) of Hilbert %~\cite{?}
asking whether there is an algorithmic procedure for deciding mathematical
statements.  Negative answers were provided independently by
Church~\cite{Church1936} and Turing~\cite{Turing} around 1936--1937,
giving birth to the mathematical theory of computation.  Their completely
different approaches gave rise to different branches of theoretical
computer science that still permeate the field.  On the one hand, the line
of development starting with Church's {\lcalculus} %~\cite{?} 
is concerned with prototypical computational languages that are
instrumental in studying high-level programming languages.  On the other
hand, Turing's machines %~\cite{?} 
are the most widely used model for analysing computational complexity.

Church's view is central to our concerns here.  The {\lcalculus} is a
deceptively simple formal system.  Its syntax consists of three types of
phrases: variables, applications, and abstraction.  Church's seminal
innovation was in introducing the latter one, 
%the other two being already familiar form the realm of
%algebra~\cite{Birkhoff}, 
which in modern terminology is referred to as a binding operator, a notion
that goes beyond the operators of universal algebra~\cite{Birkhoff}.
Binding operators are an integral part of all high-level  programming
languages.  Their mathematical theory is subtle %, see~\ref{}, 
because they define syntax up to the consistent renaming of bound
variables (technically referred to as \mbox{$\alpha$-equivalence}).
Consider, for instance, the anecdotal fact that the first definitions of
substitution were formally flawed.  
%These important topics are core to our proposed research in
%Section\pref{AlgebraicTypeTheoryParagraph}.

Incidentally, the notion of variable binding was initially introduced much
earlier, again in the context of mathematical logic.  This was done by
Frege~\cite{Frege1879} for the purpose of defining a formal symbolic
system axiomatising not only the propositional connectives of Boolean
logic~\cite{Boole} but, for the first time, also the quantifiers of
predicate logic.

In modern notation, the \lcalculus\ syntax is given by
  \[\begin{array}{rcll}
    s , t & ::= & & \mbox{($\lambda$-terms)}\\
      & \mid & x & \quad\mbox{(variables)}\\
      & \mid & (t)s & \quad\mbox{(application)}\\
      & \mid & \lambda x.\,t & \quad\mbox{(abstraction)}
  \end{array}\]
The system has only one rule of computation:
  \begin{equation}\label{BetaReduction}\begin{array}{rl}
(\beta) & (\lambda x.\,t)s \,\longrightarrow\, t[s/x]
  \end{array}\end{equation}
by which the result of computing a so-called redex $(\lambda x.\,t)s$ is
the \mbox{$\lambda$-term} $t[s/x]$ denoting the result of substituting the
free occurrences of $x$ in $t$ by $s$.  Non-terminating behaviour arises from
self application.

\paragraph{Type theory and logic.}

The concept of `type' was also born as a result of solving a foundational
problem.  In~\cite{Frege1903}, building on~\cite{Frege1879}, Frege put
forward a logical system as a foundation for mathematics including
arithmetic.  By his now famous paradox, Russell~\cite{Russell1902}
observed that one of Frege's axioms led to inconsistency.  He was led to
this by related contradictions found while developing his account of set
theory.  Russell's introduction of the `doctrine of
types'~\cite[Appendix~B]{Russell1903} came as a direct result to overcome
such foundational problems.

Type Theory as we now know it arose however from the integration by Church
of types into his \lcalculus~\cite{Church1940}.  This was a natural step,
having in mind the naive interpretation of \mbox{$\lambda$-abstraction} as
defining functions.  The type theory known as the Simply-Typed Lambda
Calculus has set of types consisting of basic ones closed under a
function-type constructor:
  \[\begin{array}{rcll}
    \sigma,\tau & ::= & & \mbox{(simple types)}\\
    & \mid & \theta & \quad\mbox{(basic types)}\\
    & \mid & \sigma\to\tau & \quad\mbox{(function types)}
  \end{array}\]
In it, \mbox{$\lambda$-terms}~$t$ are classified according to types~$\tau$ in
contexts $\Gamma$ assigning types to variables, for which the notation
  \[
  \Gamma\vdash t:\tau
\]
is commonly used.  This is done according to syntax-directed rules that in the
mathematical vernacular are presented as follows
  \[\begin{array}{c}
    1\leq i\leq n
    \\ \hline
    x_1:\tau_1,\ldots,x_n:\tau_n\vdash x_i:\tau_i
  \end{array}\]
  \[\begin{array}{c}
    \Gamma\vdash t:\sigma\to\tau
    \quad 
    \Gamma\vdash s:\sigma
    \\ \hline
    \Gamma\vdash (t)s:\tau
  \end{array}\]
  \begin{equation}\label{LambdaAbstraction} 
  \begin{array}{c}
    \Gamma,x:\sigma\vdash t:\tau
    \\ \hline
    \Gamma\vdash\lambda x.\,t:\sigma\to\tau
  \end{array}
\end{equation}

A fundamental discovery in two slightly different, though related, logical
contexts (technically, Hilbert-style deduction~\cite{Hilbert} and
Gentzen's natural deduction in sequent form~\cite{Gentzen1935}), was made
by Curry~\cite{Curry1934} and Howard~\cite{Howard1969}.  This will be
referred to here as the 
Propositions-as-Types correspondence. %(or Curry-Howard correspondence).  
Roughly, it establishes a correspondence between terms of types and proofs
of propositions.  For the Simply-Typed Lambda Calculus, the correspondence
can be illustrated by noting that the erasure of term information in its
typing rules yields the deduction rules
  \begin{center}
  $\begin{array}{c}
    1\leq i\leq n
    \\ \hline
    \tau_1,\ldots,\tau_n\vdash \tau_i
  \end{array}$
  \\[2mm]
  $\begin{array}{c}
    \Gamma\vdash \sigma\to\tau
    \quad 
    \Gamma\vdash\sigma 
    \\ \hline
    \Gamma\vdash\tau
  \end{array}
  \enspace\quad
  \begin{array}{c}
    \Gamma,\sigma\vdash\tau
    \\ \hline
    \Gamma\vdash\sigma\to\tau
  \end{array}$
\end{center}
of Intuitionistic %Minimal 
Logic. %~\cite{?}.

In another very important direction, Church introduced a Simple Theory of
Types~\cite{Church1940} as an axiomatization of Higher-Order Logic
formalised within a Simply-Typed Lambda Calculus with basic types
for individuals and propositions enriched with constants for the logical
connectives.  In doing so, he adopted a radically new perspective,
shifting the status of the Simply-Typed Lambda Calculus from that of being
a `language' to becoming a `metalanguage';~\ie, a language for languages,
in this particular case for simple type theories.  When regarded as a
metalanguage, the Simply-Typed Lambda Calculus is considered as an
equational theory, with the \mbox{$\beta$-rule}~(\ref{BetaReduction})
stated as an equation together with the extensionality equation
\[\begin{array}{rll}
(\eta) & \lambda x.\,(t)x = t 
& \mbox{, where $x$ is not free in $t$}
  \end{array}\]

Such abstraction processes from languages to metalanguages have become a
common activity in computer science and play an important role in our
proposed investigations.  

\paragraph{Category theory, logic, and type theory.}

The theory of categories was introduced by Eilenberg and Mac
Lane~\cite{EilenbergMacLane}.  A category is a mathematical structure
consisting of objects and morphisms; where the latter are classified
according to pairs of the former: as their domain and codomain.  In a
category, the notation $f:A\to B$ stipulates that $A$ and $B$ are objects,
and that $f$ is a morphism with domain $A$ and codomain~$B$.  Categories
come equipped with an associative law that composes pairs of morphisms as
on the left below
\begin{equation}\label{Category}
  \begin{array}{c}
  f:A\to B\quad g:B\to C
  \\ \hline
  g\,f:A\to C
\end{array}
\quad\qquad
1_A:A\to A
\end{equation}
together with, for every object, an identity morphism as on the right above
that is a neutral element for composition.  Categories were defined to
introduce functors, a notion of morphism between categories, which in turn was
defined to introduce natural transformations, a notion of morphism between
functors.

It is helpful to appreciate that there are two main uses for categories: in the
large and in the small.  In the large, categories are seen as mathematical
universes of discourse (within which mathematical constructions take place,
typically by means of universal properties); like the categories of: sets and
functions, algebraic structures and homomorphisms, spaces and continuous
functions.  In the small, categories are seen as mathematical objects
themselves; like a set, a monoid, a preorder, and a graph, all of which can
be suitably regarded as a category.

The connection between category theory, logic, and type theory was
initiated by Lawvere~\cite{LawvereThesis} and Lambek~\cite{LambekI}, both
of whom recognised logical systems and type theories as categories with
structure.  
%Here we are concerned with the former, postponing the latter to the following
%section.

Lawvere's insight was to understand logical connectives and type
constructors as categorical structures arising from universal properties
in the form of adjoint functors, a notion introduced by Kan~\cite{Kan}
motivated by homology theory.  For instance, the categories with structure
corresponding to the Simply-Typed Lambda Calculus with products are
Cartesian Closed Categories.  These have categorical products and
exponentials, respectively denoted $\times$ and $\Rightarrow$, defined by
adjoint situations establishing natural bijective correspondences between
morphisms as follows:
\[
  \begin{array}{c}
    C\to A \enspace,\enspace C\to B
    \\ \hline\hline
    C\to A\times B
  \end{array}
  \quad\qquad
  \begin{array}{c}
    C\times A\to B
    \\ \hline\hline
    C\to A\Rightarrow B
  \end{array}
\]
The required naturality condition amounts to the $\beta$ and $\eta$ laws.

The similarity between the bijective correspondence on the right above and
the typing rule for \mbox{$\lambda$-abstraction}~(\ref{LambdaAbstraction})
is not casual, and it is now well-understood that the Simply-Typed Lambda
Calculus provides an internal language for Cartesian Closed Categories;
namely, it is the calculus of all such models.  This view further enriches the
Propositions-as-Types correspondence as in Figure\,\ref{PAT},
\begin{figure}[h]
\caption{Propositions-As-Types correspondence}
\vspace*{2mm}
\[\xymatrix@C=10pt@R=12.5pt{
    & \ar@{<->}[dl]_(.5){\null\hspace*{-15mm}\txt{\scriptsize Internal
        Language}} \ar@{<->}[dr]^(.55){\null\hspace*{4mm}\txt{\scriptsize Categorical
        Models}}
    \txt{\small Cartesian Closed\\ \small Categories} & \\
    \txt{\small Simply-Typed\\ \small Lambda Calculus\\ \small with products}
    \ar@{<->}[rr]_-{\txt{\scriptsize Proof Theory}} & & 
    \txt{Minimal\\ Intuitionistic\\ Logic}
  }\]
\vspace*{-2mm}
\label{PAT}
\end{figure}
and it has become the standard with respect to which analogous developments in
the area are measured.  

\paragraph{Type theory and programming.}

The Simply-Typed Lambda Calculus, regarded as a language (rather than as a
metalanguage), is a prototypical functional programming language.  The change
of perspective from Type Theory to Programming Theory is not straightforward
and comes with considerations that enrich both subjects.  

In the context of programming languages, Milner %~\cite{Milner1978}
understood early on that while a programming language should come with a
type discipline to classify programs according to their type invariants,
programmers would be better served if type annotations were inferred
automatically.  

The problem of type inference (by which given a program one wishes to compute
the best possible type for it) became of central practical importance.  For
Combinatory Logic\footnote{Combinatory Logic is an important symbolic
  formalism introduced by Sch\"onfinkel~\cite{Schonfinkel} closely related
  to the {\lcalculus}, but based on algebraic combinators, that enjoys the
  Propositions-as-Types correspondence with respect to the Hilbert-style
  deduction system for Intuitionistic %Minimal 
  Logic.} 
this problem was solved by Hindley~\cite{Hindley1969}.  Independently,
however, Milner~\cite{Milner1978} solved it in a context more relevant to
programming for the Simply-Typed Lambda Calculus with parametric
polymorphism.  The algorithm is now known as the Hindley-Milner type
inference method.  The approach is very robust, having proved to extend
broadly.  
 
Polymorphism in programming refers to languages that support abstraction
mechanisms by which a program (function or procedure) can be used with a
variety of types.  The concept was introduced by
Strachey~\cite{Strachey1967}, who further classify it as ad-hoc
polymorphism and parametric polymorphism.  The former is also referred to
as overloading and accounts for uses of a program with different types
(like integers and reals) by means of different algorithms.  The latter
indicates the use of a uniform program for all types (like a duplicator
program making copies of its input).

Reynolds formalised the programming intuition by introducing the Polymorphic
Typed Lambda Calculus; an extension of the Simply-Typed Lambda Calculus with
polymorphic types.  Roughly, these are abstract parametric types whose
programs can be used for all instances of the parameter.  Strikingly, this
system had already been proposed several years earlier by
Girard~\cite{GirardSystemF} under the name of {\SystemF}, as the
type-theoretic counterpart of Second-Order Propositional Logic in the
context of the Propositions-as-Types correspondence.  Further, Girard had
also introduced {\SystemFomega}; the type-theoretic counterpart of
Higher-Order Propositional Logic.  These logical systems widely extend the
expressiveness of the Symply-Typed Lambda Calculus, notably by the
possibility of encoding inductive data types~\cite{BoehmBerarducci}.
{\SystemFomega} lies at the core of the Haskell programming
language. %~\cite{EqProofICFP12}.

The type systems mentioned above aim at providing logical foundations and, as
such, when viewed as programming languages can only introduce terminating
computations.  There are various ways in which one can extend them to Turing
complete computational languages.  In this direction, and motivated by
model-theoretic studies of the \lcalculus, Scott~\cite{ScottTCS}
introduced an extension of Typed Combinatory Logic with a fixpoint
combinator for general recursion.  Plotkin~\cite{PlotkinLCF} studied it as a
programming language, shifting from Typed Combinatory Logic to Simply-Typed
Lambda Calculus.  In doing so, he opened up further possible distinctions in
the study of type theories for computation; namely, the consideration of
equational theories for different function call mechanisms: by value or by
name~\cite{PlotkinCBVCBN}, as in ML or Haskell.  

The influence of model theories on type theories, logical systems, and
programming languages %as advocated by Scott~\cite{?} 
plays a central role throughout the proposal.

\subsection{State of the art: Questions and pathways for research}
\label{StateOfTheArtSubsection}

Having introduced the scientific framework, we turn attention to topics of
active research in the area.  The intent is not to be comprehensive.
Rather we identify main questions and pathways for research that are to be
the stepping stones for the research plan detailed in
Section\pref{MethodologySection}.

\setcounter{paragraph}{0}
\paragraph{Foundations.}

We have already mentioned several type theories, and we will mention a few
more in the sequel.  However, the following fundamental open question
remains open:
\req{(\ref{AlgebraicTypeTheoryParagraph})}
{What is a type theory?}

Section\pref{AlgebraicTypeTheoryParagraph} aims at a comprehensive
mathematical answer, that will also serve as a framework for our other
type-theoretic developments.  
\hide{
  We regard this as a step towards the related open question:
\req{}{What is a programming language?}
that will be kept in the background of our investigations.
}

\paragraph{Logical wiring.}
\label{LogicalWiringSubsection}

Lambek~\cite{LambekI} recognised the similarity between the basic
structure of a category, given by identities and
composition~(see\pref{Category}), and the wiring of logical deduction
systems, specifically Gentzen's sequent calculi~\cite{Gentzen1935}, given
by the axiom and cut rules.  Intuitionistic sequents led Lambek to
axiomatize their algebra under the concept of multicategory.  This
correspondence is roughly as outlined below (where the vector notation
$\vec{\,\cdot\,}$ stands for finite sequences of objects): 
\begin{center}\begin{tabular}{|c|c|}\hline
    \begin{tabular}{c}
      \small Intuitionistic\\[-1mm] \small Sequent Calculus
  \end{tabular}
  &
  \begin{tabular}{c}
    \small
    Multicategories
  \end{tabular}
  \\ \hline\hline
  \begin{tabular}{l}
    \hspace*{-15mm}
    \small 
    (Axiom)
    \\[-3mm]
    $\begin{array}{c}\\ \hline P \vdash P\end{array}$
  \end{tabular}
  & 
  \begin{tabular}{l}
    \small
    \hspace*{-7mm}
    (Identity)
  \\[1mm]
  $(A)\stackrel{1_A}\longrightarrow A$
  \end{tabular}
  \\ \hline
  \begin{tabular}{l} 
    \hspace*{-2mm}(Cut) 
    \\[1mm]
    $\begin{array}{c}
    \Gamma\vdash A\quad\Gamma_1,A,\Gamma_2\vdash B
    \\ \hline 
    \Gamma_1,\Gamma,\Gamma_2\vdash B
    \end{array}$
  \end{tabular}
  &
  \begin{tabular}{l}
    \small
    \hspace*{-3mm}
  (Multicomposition)
  \\%[1mm]
    \hspace*{-2mm}
    $\begin{array}{c}
    \vec Y \stackrel f\longrightarrow A
    \\
    \vec X,A,\vec Z \stackrel g\longrightarrow B
    \\ \hline\\[-4mm]
    \vec X, \vec Y,\vec Z
    \xymatrix@C=25pt{\ar[r]^-{g\, _{\vec
          X}\hspace*{-.0125mm}\circ\hspace*{-.25mm}_{\vec Z}f}&} 
    B
    \end{array}$
    \hspace*{-2mm}
  \end{tabular}
  \\ \hline
\end{tabular}\end{center}
The analogous for classical sequents was done by Szabo~\cite{Szabo} with the
introduction of polycategories.

The structure of multicategory appeared independently in a very different
mathematical context, the work of May~\cite{May} on homotopical algebra, under
the name of Operad.  Operads are now central to studies in higher-dimensional
algebra, see~\eg~\cite{Leinster,LodayVallette}.  

\rep{(\ref{WiringStructureParagraph})}
{Investigate interactions between logical sequent calculi and algebraic operad
  theory.}

\paragraph{Dependent types.}
\label{DependentTypesParagraph}

Dependent type theory is a formalism introduced by
de~Bruijn~\cite{deBruijn} that extends that of simple type theories by
allowing types to be indexed or parameterised by other types.  Such
objects abound in computer science and mathematics.  For instance, in
combinatorics one is interested in the type of permutations $\mathfrak
S(n)$ on $n$ elements, as the index (or parameter) $n$ ranges over the
natural numbers $\mathbb N$.  In modern notation, this is presented by a
judgement of the form
\[
  n:\mathbb N \vdash \mathfrak S(n) \mbox{ \mysf type}
\]
There are two fundamental constructions on such dependent judgements as
follows
\[
  \begin{array}{c}
  i: I \vdash T(i) \mbox{ \mysf type}
  \\ \hline
  \vdash \Sigma\,{i:\!I}.\, T(i) \mbox{ \mysf type}
  \end{array}
  \qquad
  \begin{array}{c}
  i: I \vdash T(i) \mbox{ \mysf type}
  \\ \hline
  \vdash \Pi\,{i:\!I}.\, T(i) \mbox{ \mysf type}
  \end{array}
\]
respectively known as dependent sums and dependent product types.  These
generalise the product and function types of Simply-Typed Lambda Calculus and,
under the Propositions-as-Types correspondence, amount to existential and
universal quantification.  We omit discussing the syntax of terms for these
types.  As for their equational theory, we only mention that dependent sums
may be weak or strong and that dependent products may be intensional (under
$\beta$ equality) or extensional (under $\beta\eta$ equality). 
%---see \eg~\cite{?}.

A basic question in this area, for which there is as yet no satisfactory
answer is worth noting:
\req{(\ref{IntensionalTypeTheoryParagraph})}
  {What is a model of intensional type theory?}
This is part of the proposed investigations
in Section\pref{IntensionalTypeTheoryParagraph}.

The passage from sum and product types to their dependent versions
required new type theories.  Categorical models suggest further
generalisations.  These are the subject of
Sections\,(\ref{GeneralisedTypeTheoryParagraph}\,\&\,\ref{DirectedTypeTheoryParagraph})
under the following. 
\rep{(\ref{CalculiSubsection})}
  {Develop type theories from mathematical models.}

\paragraph{Mathematical universes.}
\label{MathematicalUniversesParagraph}

The ability to construct new mathematical universes of discourse from old ones
is a fundamental part of the technical toolkit of researchers in semantics, be
it either in logic or computation.

One technique to do so is to enrich the semantic universe with a mode of
variation.  In its basic form, starting from the universe of sets and
functions~$\Set$ one considers the universe~$\Set^{\scat C}$ of so-called
presheaves consisting of the \mbox{$\scat C$-variable} sets for a small
category~$\scat C$.  The importance of this passage is that the kind of
variation embodied in the parameter small category translates to new, often
surprising, internal structure in the universe of presheaves.  

The presheaf construction was introduced by Grothendieck together with a
more sophisticated and important refinement of it known as the sheaf
construction~\cite{SGA4} (in the context of the Weil conjectures in
cohomology theory).  Sheaves are a central object of study in the area of
mathematics known as Topos Theory~\cite{Elephant}; a topos being a
universe of discourse for Higher-Order Intuitionistic
Logic~\cite{LambekScott}.  

In view of the many possible applications, it is natural to ask:
\reqs{(\ref{MethodologyMathematicalUniversesParagraph})}
  {Which mechanisms are there for changing from a type theory to another one
    as universes of discourse? 

    Can this be done while maintaining the relevant computational properties
    and then incorporated into mechanical proof assistants?}

This question should not only be considered from the topos-theoretic viewpoint
mentioned above, but also from other approaches, notably that of the related
forcing technique of Cohen~\cite{Cohen} (introduced by him to prove the
independence of both the axiom of choice and the continuum hypothesis from
Zermelo-Fraenkel Set Theory) and its elaboration by Scott and Solovay as
Boolean-valued models~\cite{ScottSolovay}.

%Recent work in this direction has been done by
%Coquand~\cite{CoquandForcingInTT} and Jaber, Tabareau, and
%Sozeau~\cite{TypeTheoryWithForcing}.\footnote{Note
%  that~\cite{TypeTheoryWithForcing}, despite its title, is about internalising
%  the presheaf construction on partial orders.}

\paragraph{Indexed programming.}
\label{IndexedProgrammingIntro}

The use of presheaf categories in computer science applications has been
prominent;~\eg~in 
programming language theory, %~\cite{?}, 
lambda calculus, %~\cite{?}, 
domain theory, %~\cite{?}, 
concurrency theory, %~\cite{?}, 
and type theory. %~\cite{?}.  
In particular, in their most elementary discrete form, presheaves can be
found in programming languages as indexed datatypes; programming with
which will be generically referred to as indexed programming.  

Indexed programming developed from two main influences (none to do with
presheaves though): the practical needs of supporting data structures able to
maintain strong data invariants, like nested %~\cite{?} 
and generalised %~\cite{?}
datatypes in functional programming~\cite{Omega,Haskell}; and the
experimentation with dependently-typed programming
languages~\cite{Cayenne,Epigram} as a by-product of dependent type theory.
These two views somehow pull in opposite directions and, as such, lead to
conceptually different languages.  One is lead to investigate the following.
\rep{(\ref{IndexedProgrammingParagraph})} 
  {Develop foundational type theories for indexed datatypes.  
    
   Design and %subsequently 
   implement indexed programming languages from these and
   %programming-language 
   pragmatics.}

\paragraph{Resources, effects, modalities.}
\label{ResourcesEffectsModalitiesParagraph}

In the late 1980s, two important analyses of computation, respectively for
resources and effects, were proposed by Girard~\cite{GirardLinearLogic} and by
Moggi~\cite{MoggiLambdaC}.  The former was in the contexts of logic and proof
theory; the latter in that of denotational semantics and category theory.
Both, however, have had tremendous impact in programming language theory.  The
resource analysis in the form of Linear Logic calculi, and the effect analysis
in the form of Computational \mbox{$\lambda$-calculi}.

From the viewpoint of categorical models, the resource and effect management
structures are respectively seen as comonadic and monadic structures.
Comonads and monads being dual categorical concepts that arose in the context
of cohomology theory in the 1960s~\cite{BeckThesis}.

A classical basic result of category theory establishes a strong
correspondence between (co)monads and adjoint functors,
see~\eg~\cite{MacLane}.  One aspect of this is that every adjunction 
\begin{equation}\label{ResourceEffectAdjunction}
  \xymatrix@C=40pt{
    \cat D \ar@/^.75em/[r]|(.625){\mbox{$\,G\,$}}\ar@{}[r]|-\swvdash&
    \ar@/^.75em/[l]|(.625){\mbox{$\,F\,$}} \cat C }
\end{equation}
with $F$ and $G$ respectively left and right adjoint to each other, gives
rise (by composition) to a comonad on $\cat D$ and a monad on $\cat C$.  

This viewpoint gave impetus to further analyses based on the more primitive
notion of adjunction.  Models of Linear Logic are founded on the theory of
monoidal categories~\cite[Chapter~VII.1]{MacLane}.  For them, one requests
that the adjunction be monoidal with respect to linear structure on $\cat D$
and cartesian (or multiplicative) structure on $\cat C$,
see~\eg~\cite{MelliesCMLL}.  On the other hand, models of Computational
\mbox{$\lambda$-calculi} rely on enriched category theory~\cite{KellyBook}.
Here, the structure is roughly given by an enriched adjunction with $\cat C$
cartesian and $\cat D$ with powers (or exponentials) relative to $\cat C$.

Examples in the context of linear logic are the mixed linear/cartesian models
and calculi of Benton and Wadler~\cite{BentonWadler} and of Barber and
Plotkin~\cite{BarberPlotkin}; and, in the context of effect calculi, the
Call-By-Push-Value of Levy~\cite{LevyCBPV} and the Enriched Effect
Calculus of Egger, M{\o}gelberg, and Simpson~\cite{EEC}.  Metaphorically
speaking, both these lines of work regard resources and effects as being
opposite sides of the same coin.  However, this is not so in all models;
and the following question remains unanswered.
%
\req{\pref{PolarisationParagraph}}
  {How can resources and effects be reconciled and unified?}
%
To answer it, a broader view of the subject involving the orthogonal
notion of polarisation seems to be needed.  From the programming-language
viewpoint, this further enriches the computational picture with the
ability of distinguishing between eager \vs~lazy modes of computation and
data structure.  We incorporate this into the following.
%
\rep{(\ref{ProgrammingEffectsParagraph})}
  {Study and develop the theory of resource management, computational
    effects, and polarisation.  Percolate this down into the design of
    programming languages.}

From the logical point of view, resource comonads and effect monads are
so-called modal operators, and some of the literature has indeed
considered them as such (see \eg~\cite{Kobayashi}).  %BiermanDePaiva
The field of modal logics is broad, with many subfields of specialised
logics motivated by computation, linguistics, and philosophy.  While a
class of modal logics known as temporal logics have played a prominent
role and been very successful in the area of computer aided verification;
the impact of modal logics on programming languages has been peripheral.
It is thus natural and important to reconsider them in this context. 
%
\rep{(\ref{ModalLogicsParagraph}\,\&\,\ref{MetaprogrammingParagraph})}
  {Investigate the Propositions-As-Types correspondence for modal logics,
    and apply it to programming languages.}
Let us stress that we are specially interested in modalities for
computation with reflection.

\paragraph{Sequent calculi.}
\label{SequentCalculiParagraph}

A theme that runs orthogonal to the logics under consideration is whether
they are specified in natural deduction or sequent calculus
style~\cite{Gentzen1935}.

Most of the work on the Propositions-As-Types correspondence has been done
for natural deduction systems; especially in relation to programming
language theory, where the logical syntax matches that of functional
languages.  

On the other hand, there is as yet no established syntax for sequent-style
calculi.  A question at the core of this situation is:
%
\req{(\ref{ProgrammingEffectsParagraph})}
  {What is the categorical algebra of classical sequent calculi?}
%
Nevertheless, a syntactic system that is proving robust in applications
seems to be emerging.  This is the {\SystemL} of Curien and
Herbelin~\cite{CurienHerbelin}, applied further
in~\eg~\cite{Wadler,Munch,CurienMunch}.

The main novel features of {\SystemL}, and its philosophy, are an
intrinsic symmetry (reflecting the computation roles of program and
environment) and a close connection with abstract machines (which are
internalised into the calculus).  From this perspective:
%
\req{(\ref{ProgrammingEffectsParagraph})}
  {What can the proof theory of sequent calculi do for programming?}

\subsection{Research objectives}
\label{ObjectivesSubsection}

The general aim of {\erc} is to build a group dedicated to combined
research in category theory, mathematical logic, programming theory, and
type theory to advance current knowledge in their mathematical foundations
and practical applications.  To this end, {\erc} assembles an
international team of world-leading experts in each of the project
research areas.  

The team consists of \emph{Principal Investigator} Marcelo Fiore;
\emph{Senior Researchers} Pierre-Louis Curien, Peter Dybjer, and Timothy
Sheard; and \emph{Junior Researchers} Ki Yung Ahn, Nicola Gambino, and
Guillaume Munch-Maccagnoni.  Figure\,\ref{ercTeam} gives a schematic
presentation of the team that matches the research areas as rendered in
Figure\,\ref{ResearchAreas}.  
\begin{figure}[h]
\caption{{\erc} team}
\vspace*{2mm}
\begin{center}
\hspace*{.125mm}
\xymatrix@R=10pt@C=32.5pt{
& 
\raisebox{5mm}
  {\fbox{\txt{\small Pierre-Louis\\ \small Curien}}}
\ar@/_1em/@{<->}[dddl]<-1em>
\ar@/^1em/@{<->}[dddr]<1em> 
& 
\\
\\
& 
  {\fbox{\txt{\small Marcelo\\ \small Fiore}}}
\ar@{<->}[dl]|-
  {\txt{\scriptsize Nicola\\ \raisebox{1mm}{\scriptsize Gambino}}}
\ar@{<->}[dr]|-
  {\txt{\scriptsize Ki Yung\\ \raisebox{1mm}{\scriptsize Ahn}}}
\ar@{<->}[uu]|(.475)
  {\txt{\scriptsize Guillaume\\ \raisebox{1mm}{\scriptsize
        Munch-Maccagnoni}}}
& 
\\
\raisebox{0mm}{\fbox{\txt{\small Peter\\ \small\ Dybjer\ \ }}}
\ar@/_1em/@{<->}[rr]
& & 
\raisebox{0mm}{\fbox{\txt{\small Timothy\\ \small Sheard}}}
}
\end{center}
\vspace*{-2mm}
\label{ercTeam}
\end{figure}
Further details are deferred to Section\pref{ResourcesSection}. 

The overall research objectives that we are to undertake are classified
under four headings as follows.
\begin{myitemize}
\item[{\bfseries 1\enspace Foundations:}]\mbox{}\enspace\thinspace 
  A comprehensive research programme on the metamathematics of type theories.

  We will provide an algebraic answer to what type theories are, establishing
  connections with areas of abstract algebra where related structures play a
  central role.

\item[{\bfseries 2\enspace Models:}]\mbox{}\enspace\thinspace
  Study %Development and study 
  of mathematical models for type theories and logical systems. 

  We will investigate model theories targeting semantic problems at the
  forefront of current understanding, specifically for intensional type theory
  and polarised logic.
  
\item[{\bfseries 3\enspace Calculi:}]\mbox{}\enspace\thinspace
  Development of formalisms of deduction as internal languages of mathematical
  theories.
  
  We will research type theories that go beyond current logical
  frameworks, especially motivated by higher-dimensional category theory
  and complexity theory.
  
\item[{\bfseries 4\enspace Programming:}]\mbox{}\enspace\thinspace
  Design and implementation of, and experimentation with, novel computational
  languages.  
  
  We will target languages with first-class indexed data structures,
  programming paradigms stemming from sequent calculi embodying
  computational effects, and explore extensions for metaprogramming.
\end{myitemize}

\section{\underline{Methodolo}gy\hspace{-1mm}\underline{\,}}
\label{MethodologySection}

This section expands our research objectives providing a plan together
with the methodology for its completion.

\hidenote{\ldots}

\subsection{Foundations}
\label{Foundations}

\hidenote{\ldots}

\setcounter{paragraph}{0}
\paragraph{Algebraic Type Theory.}
\label{AlgebraicTypeTheoryParagraph}

To understand our approach it is best to start by stripping type theories
down to their essential bare structure.  To do so, let us eliminate type
constructors and binding operators from them.  What one is left with is a
notion of type theory that reduces to that of many-sorted algebraic
theory~\cite{Birkhoff}, a thoroughly studied area of algebra.

The modern understanding of many-sorted algebraic equational theories is
through three interrelated perspectives as in the Algebraic Trinity of
Figure\,\ref{AlgebraicTrinity}.
\begin{figure}[h]
  \caption{Algebraic Trinity}
  \vspace*{2mm}
  \begin{center}$\xymatrix@R=7.5pt@C=35pt{
      & \txt{\small Categorical\\ \small Algebra} 
      \ar@{<->}[dr]^(.55){\null\hspace*{5mm}\txt{\scriptsize Lawvere~\cite{LawvereThesis}}}
      \ar@{<->}[dl]_(.55){\null\hspace*{-10mm}\txt{\scriptsize Linton~\cite{Linton}}}
      & 
      \\
      \txt{\small Equational\\ \small Logic}
      \ar@{<->}[rr]_-{\txt{\scriptsize Birkhoff~\cite{Birkhoff}}}
      &
      & 
      \txt{\small Universal\\ \small Algebra} 
    }$\end{center}
  \vspace*{-2mm}
\label{AlgebraicTrinity}
\end{figure}
Equational Logic is the metalanguage of algebraic theories, while
Universal Algebra is its model theory.  They were both introduced by
Birkhoff~\cite{Birkhoff}, with the former as a by-product of the latter
related by a soundness and completeness theorem.  The categorical
viewpoint of algebraic theories came later, with the work of
Lawvere~\cite{LawvereThesis} and Linton~\cite{Linton}.  Two crucial
categorical structures that play a pivotal role here are: Lawvere
theories~(closely related to Hall's abstract clones) and finitary
monads~(a notion 
%originally introduced in cohomology theory by Beck~\cite{Check???}, 
arising in algebra from free constructions).

Each of the approaches to algebraic theories of the Algebraic Trinity
gives a different viewpoint of the subject.  For instance, Equational
Logic provides the deductive system for formal reasoning about equations
from axioms; Universal Algebra provides a general notion of model from
which one derives an abstract notion of syntactic structure by means of
freely generated models; Categorical Algebra provides a notion of
translation between theories that lifts to adjoint functors between
categories of models.

Having deconstructed type theories down to many-sorted algebraic theories, can
one reconstruct them back (and in the process enrich them) as a unified
mathematical theory encompassing all aspects of the Algebraic Trinity?  This
is the main general goal of this research track.

To make substantial progress in the area, our proposal is to
systematically explore a broad spectrum of key features present in type
theories. The scale at which we will be attempting this is unprecedented,
and is graphically represented by the research space of investigation in
Figure\,\ref{TypeTheoryResearchSpace}, 
\begin{figure}[t]
\caption{Algebraic Type Theory research space}
\vspace*{1mm}
\begin{center}
$\xymatrix@C=2.5pt@R=25pt{
*+[]{\fbox{\scriptsize\txt{Many-Sorted\\ Univ.\,Alg.
%\\ Eq.\,Logic
}}}
\ar[rrr]|-{\mbox{\scriptsize\txt{\emph{type}\\\emph{dependency}}}}
& 
& 
& 
*+[]{\fbox{\scriptsize\txt{Dependently-Sorted\\ Algebra}}}
\ar[dd]|-{\mbox{\scriptsize\emph{binding}}}
\\
*+[]{\fbox{\scriptsize\txt{Univ.\,Alg.%\\ Eq.\,Logic
}}}
\ar[u]|-{\mbox{\scriptsize\emph{sorting}}}
\ar[rd]|-{\mbox{\scriptsize\emph{binding}}} 
\ar[dd]|-{\mbox{\scriptsize\emph{linearity}}}
& &
*+[]{\fbox{\scriptsize\txt{Simple\\Type Theory}}}
\ar[rd]|-{\mbox{\scriptsize\txt{\emph{type}\\\emph{dependency}}}}
\ar[dd]|-{\mbox{\scriptsize\txt{\emph{explicit}\\\emph{polymorphism}}}}
& 
\\ 
& 
*+[]{\fbox{\scriptsize\txt{Binding\\Algebra}}}
\ar[ur]|-{\mbox{\scriptsize\txt{\emph{simple}\\\emph{types}}}}
&
&
*+[]{\fbox{\scriptsize\txt{Dependent\\Type Theory}}}
%\ar@{.>}[d]
\\
*+[]{\fbox{\scriptsize\shortstack{Operads}}}
& & 
*+[]{\fbox{\scriptsize\txt{Polymorphic\\Type Theory}}}
%\ar@{.>}[r]
& 
%*+[]{\fbox{\scriptsize\txt{?}}}
}$
\end{center}
\vspace*{-3mm}
\label{TypeTheoryResearchSpace}
\end{figure}
where the \emph{linearity} dimension could be transported along or mixed
with all the other ones.

Initial investigations in this direction have been carried out by Fiore
and PhD~students~\cite{FioreHur,FioreMahmoud}.  These extend the Algebraic
Trinity to the realm of Binding Algebra,~\viz~algebraic languages with binding
operators.  The extension to simple type theories should follow along
similar lines further taking into account the algebraic structure of
types.  As for the other points of the research space in
Figure\,\ref{TypeTheoryResearchSpace}, they are yet to be investigated.  For
these we have the following goals.
\begin{resenumerate}\setcounter{CC}{0}
\item
  To develop a mathematical algebraic framework for the semantics of
  language phrases, whereby free models universally characterise the
  abstract syntax of the language.

\item
  To synthesise metalanguages for type theories in the form of formal
  systems for equational deduction, that are sound and complete for the
  model theory.

\item
  To extract syntactic notions of translations between type theories as
  suggested by the mathematical models, and to use these to provide
  constructions for the modular combination of type theories.

\item
  To test and apply the above mathematical theories by formalising them in
  Coq or Agda, while at the same time exercising these systems to the
  limit to identify shortcomings to trigger new research in the context of
  proof assistants.
\end{resenumerate}
The outcome of this work will be mathematical foundations for the
aforementioned aspects of type theories that are currently treated in an
ad~hoc fashion.  Furthermore, it will provide a principled approach by which
to reconsider the problem of integrating polymorphism and dependent types. 
%%%to tackle the problem of integrating polymorphism and dependent types in 
%%% a satisfactory 
%%%manner---the unknown in Figure\,\ref{TypeTheoryResearchSpace}. 

\hide{ polynomial functors (from discrete to groupoidal) --- the meaning of
  typing rules}

\hide{CwFs \vs~substitution algebras}

\paragraph{Wiring structure.}
\label{WiringStructureParagraph}

A basic concern in the investigation of
Section\pref{AlgebraicTypeTheoryParagraph} is the
understanding of the mathematical structure of sequents and judgements.
This subject has already received attention, but new perspectives are emerging
that call for its reconsideration.

Remitting ourselves back to Section\pref{LogicalWiringSubsection}, the
first thing to notice is that there are two possible axiomatizations for
multicategory composition: the one exhibited there, known as partial
composition, and a total (or simultaneous) one of the form
{\small\begin{equation}\label{TotalComposition}\begin{array}{c}
  g: A_1,\ldots,A_n \to B
  \\[1mm]
  f_i:\vec{X_i}\to A_i\enspace(1\leq i\leq n)
  \\[.75mm] \hline\\[-2.5mm]
    g\circ(f_1,\ldots,f_n):\vec{X_1},\ldots,\vec{X_n}\to B
\end{array}\end{equation}}%
In the presence of identities, both notions become equivalent.  An interesting
observation from the theory of operads is that this is not so
otherwise~\cite{Markl}.

Algebraically, the total composition structure~(\ref{TotalComposition}) is
well understood as given by semigroup structure (technically in the
category of Joyal Linear Species~\cite{JoyalAdvMath} with respect to the
substitution tensor product~\cite{KellyOperads,JoyalLNM1234}).  But, what
about the concept of partial composition?  That, after all, is the
predominant in logic and type theory.  For it, Fiore has recently
uncovered an unexpected connection with structure stemming from
Lie-algebra theory; specifically, the pre-Lie algebra structure of
Gerstenhaber~\cite{Gerstenhaber}.  Furthermore, and analogous, though more
interesting situation occurs with polycategories~\cite{Szabo} and the
Lie-admissible algebraic structure of Albert~\cite{Albert}.

We regard these observations as the starting point for interactions between
operad theory and logic and type theory for pursuing research as follows.

\noindent
From operad theory to logic and type theory:
\begin{resenumerate}\setcounter{CC}{0}
  \item
    Kontsevitch introduced the notion of a meager PROP (renamed
    as \raisebox{.75mm}{\scriptsize$\frac 1 2$}PROP by
    Markl~\cite{Markl}) sitting in between that of operad and
    polycategory.  Is there any use for these in logic and type theory?

  \item
    What about other generalisations, like the preshuffle algebras of
    Ronco~\cite{RoncoShuffleBialgebras} and the permutads of Ronco and
    Loday~\cite{Permutads}?  In particular, the former embodies the
    algebra needed in computational contexts with effects, where the
    notion of total composition does not make sense, and the notion of
    partial composition is subject to weaker axioms to those of
    multicategory.  This is relevant to our proposed investigations in
Sections\,(\ref{PolarisationParagraph}\,\&\,\ref{ProgrammingEffectsParagraph}).

  \item
    There is a rich variety of notions of operads, and notions of
    composition between them, many of which are motivated by
    combinatorics.  What can they offer to logic and type theory?  This
    question is not as vacuous as it may seem at first sight.  For instance,
    the shuffle and quasi-shuffle composition (see~\eg~\cite{AguiarMahajan})
    seem to be related to a type theory for non-commutative linear logic of
    Polakow and Pfenning~\cite{NatDedForIntNonCommLinLog}.

\hide{
  \item
    Foundations for a TTyacc (Type Theory yacc) \ldots this is related to what
    is a Type Theory? \ldots The feasibility of building a TTyacc will be
    investigated \ldots relevant to reflection in type theory \ldots extract a
    language from the mathematics
  }
\end{resenumerate}

\noindent
From logic and type theory to operad theory:
  \begin{resenumerate}\setcounter{CC}{3}
  \item
    The new Lie-algebraic view of multicategory composition mentioned
    above gives a rational reconstruction of an explicit characterisation of
    the pre-Lie operad by Chapoton and Livernet~\cite{ChapotonLivernet}.  The
    case of polycategory composition leads to a conjectured characterisation
    for the Lie-admissible operad.  A possible approach to proving it leads to
    the introduction of new interesting algebraic structures.

  \item\label{BellantoniCook}
    A combination of the \emph{linearity} dimension of
    Figure\,\ref{TypeTheoryResearchSpace} with the other dimensions (that take
    place in the traditional \emph{cartesian} setting, where all structural
    rules on contexts are allowed) leads to a notion of mixed
    operad~\cite{FioreMFPS}, whose associated composition operation is new
    to the mathematical theory.

    Mixed operads are the wiring of linear/cartesian languages, which since
    the work of Bellantoni and Cook~\cite{BellantoniCook} play a role in
    implicit complexity theory.  It would be interesting to develop this
    connection.  A main aim would be to introduce new mathematical structure
    in implicit complexity theory that would help with its problems.  This is
    relevant to our proposed investigations in
    Section\pref{SubstructuralTypeTheoryParagraph}.

  \item
    The algebraic structure on operads considered in mathematics is in
    the tradition of linear algebra (\ie~first-order).  However, the binding
    operators of logic and type theory transport to operads as a new kind of
    algebraic structure which thereby enriches the theory and is worth
    investigating.
  \end{resenumerate}

\subsection{Models}
\label{Models}

\hidenote{\ldots}

\setcounter{paragraph}{0}
\paragraph{Intensional dependent type theory.}
\label{IntensionalTypeTheoryParagraph}

We have referred to intensional and extensional type theory in
Section\pref{DependentTypesParagraph}.  Recall that the essential
difference between them is that the latter extends the computational (or
$\beta$) rules of the former with extensionality (or $\eta$) rules.  Since
type checking for intensional type theory is decidable, while it is not so
for extensional type theory, the former has become the core of
foundational systems for proof assistants as metalanguages for
constructive mathematics; as 
Automath, %~\cite{deBruijn}, 
Coq, %~\cite{?} 
and 
Agda. %~\cite{?}.  
All these systems are based on very rich type theories.  For instance, the
two current mainstream systems Coq and Agda are respectively based on the
Calculus of Constructions~\cite{CoquandHuet} and Martin-L\"of Type
Theory~\cite{ProgMLTT}.  The proposed investigations below are in this
context.  Some aspects of these arose and are under discussion with
Warren~(IAS School of Mathematics, Princeton).

The overall goal here is to:
\begin{resenumerate}\setcounter{CC}{0}
\item 
  Develop concrete and abstract models of intensional dependent type theories;
  and use them both to establish relationships between type theories and
  between type theory and homotopical mathematical theories.  Research on the
  latter point has only recently gathered momentum in the community.
\end{resenumerate}

Conceptually, one may regard a type theory as a basic deduction system
extended with type constructors.  For instance, Martin-L\"of Type Theory
embodies dependent sums and dependent products, as well as Identity Types,
W~Types, and Universes~\cite{ProgMLTT}.  As such, a type theory is to be
thought not just as a single universe of discourse, but rather as a
variety of universes of discourse corresponding to the various possible
restrictions.  From this perspective, it is important to understand the
interaction of modularly extending type theories, central to which is the
notion of conservative extension.  Informally, one system is a
conservative extension of another if it faithfully embeds it.  

Conservativity has been studied for some simple type 
theories, %, see~\eg~\cite{Lafont,FioreRemarks}, 
but it has been overlooked for dependent type theories.  We thus propose
here to:
\begin{resenumerate}\setcounter{CC}{1}
\item
  Develop a model-theoretic framework for establishing conservative
  extension results for Martin-L\"of Type Theory and Pure Type Systems.
\end{resenumerate}
The programme will lead to sophisticated model constructions that will
deepen our understanding of the subject.  In particular because this was
already the case in proving the conservative extension of the Simply-Typed
Lambda Calculus extended with (non-dependent) extensional strong
sums~\cite{FioreRemarks}.

To pursue the program for intensional type theory, some fundamental
problems will need to be addressed.  For a start:
\begin{resenumerate}\setcounter{CC}{2}
\item 
  What is a model of intensional dependent type theory?
\end{resenumerate}
Despite the fundamental character of the question there is as yet no
established answer.  This is in opposite contrast with the extensional
case, where categorical models are well understood~\cite{Jacobs}.  There
are various reasons for this situation.

\marginpar{\color{red}revise --- \cite{DybjerITT}}

Firstly, current categorical models are not flexible enough to dispense with
extensionality.  In this respect, the work on Algebraic Type Theory of
Section\pref{AlgebraicTypeTheoryParagraph} will provide new directions.
Moreover, concrete models of intensional type theory that researchers work
with (like groupoids, strict $\omega$-groupoids, and simplicial sets)
interpret dependent sums and dependent products as adjoints and are hence
extensional.  A question thus arises:
\begin{resenumerate}\setcounter{CC}{3}
\item 
  Are there natural non-extensional models of dependent type theory?
\end{resenumerate}

Secondly, the study of intensional models with Identity Types~\cite{HS} is
very much under development.  In fact, Identity Types are currently
receiving renewed special attention, steaming from connections with
homotopy theory %~\cite{?} 
(and higher-dimensional category theory) %~\cite{?})
that are feeding back into mathematical foundations, specifically
Voevodsky's Univalent Foundations of Mathematics program. %~\cite{?}.  
A main open question~\cite{???} to be considered is: 
\begin{resenumerate}\setcounter{CC}{4}
\item 
  What is the relationship between the syntactic model of Identity Types and
  the mathematical Homotopy Types? 
  %and Weak Higher-Dimensional Categories?
\end{resenumerate}
Steps into this problem have been made,
see~\eg~\cite{AwodeyHofstraWarren}.  But there are fundamental matters to
be understood; for instance, when seen from the viewpoint of Algebraic
Type Theory: 
\begin{resenumerate}\setcounter{CC}{5}
\item 
  What is the relationship between the algebraic theory of Identity Types
  and the algebraic theory of Weak Higher-Dimensional Categories?
\end{resenumerate}
This is to be investigated specially noting that the former crucially
relies on binding operators while the latter does not.

The Univalent Foundations of Mathematics program mentioned above is based
on Voevodsky's Univalence Axiom. %~\cite{?}.  
This axiom has been shown consistent with Martin-L\"of Type Theory from
the outset, but:
\begin{resenumerate}\setcounter{CC}{6}
\item 
  Is the extension of Martin-L\"of Type Theory with the Univalence Axiom
  conservative?
\end{resenumerate}

Also, it is an intriguing fact that the Univalence Axiom somehow endows
Identity Types with logical character, very roughly in that the Identity
Type constructor logically commutes with all the other type constructors
up to equivalence.  It would be interesting to investigate the converse:
\begin{resenumerate}\setcounter{CC}{7}
\item
  Do Logical Identity Types imply the Univalence Axiom?
\end{resenumerate}

Finally, it is a main open problem as to how to give a constructive
interpretation of the Univalence Axiom.  One could instead aim to:
\begin{resenumerate}\setcounter{CC}{8}
\item
  Design a constructive type theory with built-in Logical Identity Types.  
\end{resenumerate}
Here, we will:
\begin{list}{[\textbf{\arabic{CC}]}}
  {\usecounter{CC}
   \setlength{\topsep}{2pt}
   \setlength{\partopsep}{2pt}
   \setlength{\itemsep}{2.5pt}
   \setlength{\parsep}{2.5pt}
   \setlength{\leftmargin}{2.15em}
   \setlength{\labelwidth}{1.65em}}
  \setcounter{CC}{9}
\item
  Investigate possible connections to Strachey's notion of parametric
  polymorphism~\cite{Strachey1967} in computer science as formalised by
  Reynolds~\cite{Reynolds} using logical relations.
\end{list}
Were these connections to materialise, it will open up a new flow of ideas
between hitherto disconnected fields.

\paragraph{Mathematical universes.}
\label{MethodologyMathematicalUniversesParagraph}

We speculate here on directions in relation to
Section\pref{MathematicalUniversesParagraph}, where we are interested in
building new models from old ones.  A main source of inspiration and
guidance for the development comes from Topos Theory~\cite{Elephant}.

Our proposal is to:
\begin{resenumerate}\setcounter{CC}{0}
\item\label{ConstructionsOnTypeTheories}
  Investigate presheaf, orthogonality, sheaf, glueing, and forcing
  constructions on type theories.
\end{resenumerate}
with the general goal to:
\begin{resenumerate}\setcounter{CC}{1}
\item
  Build a mathematical theory of constructions on type theories that
  produce new type theories from old ones, together with an interpretation
  (or compilation) of the latter into the former.
  
\item
  Implement the mathematics in proof assistants, and exercise it in
  applications.  
\end{resenumerate}

Initial work in this line has been done by Jaber, Tabareau, and
Sozeau~\cite{TypeTheoryWithForcing}, who formalised the presheaf
construction on a partial order in the Calculus of Constructions, and then
re-interpreted back the Calculus of Constructions in the internal presheaf
model.  However, not only a wide spectrum of other possible constructions
(as detailed in
item~\textbf{\ref{MethodologyMathematicalUniversesParagraph}}\thinspace\itemref{ConstructionsOnTypeTheories}
above) remains to be explored; but, even in this basic setting, work remains
to be done: notably to incorporate Inductive Types.

What we envisaged is also more general, in that we take the view that the
new type theory need not be of the same character as the old one.  Thus
there are two aspects to our proposed investigations: the mathematical one
of studying model constructions and the type theoretic one of designing
internal languages.  To fix ideas, consider as an example that the
presheaf construction on a monoidal category endows the new universe of
discourse with monoidal closed structure (formally via Day's convolution
monoidal structure~\cite{Day}), and thereby introduces new linear structure.
This is important in many applications, for which the reader may
consult~\cite{FioreFossacs}, and complementary to our proposed
investigations in Section\pref{SubstructuralTypeTheoryParagraph}.

The study of Cohen Forcing for type theory was stated as an open problem
by Beeson~\cite{BeesonBook}, who wrote: ``Forcing has yet to be worked out
directly for Martin-L\"of's system.''.  Work in this direction has only
recently begun.  Specifically, by Coquand and Jaber~\cite{CoquandNote} who
presented an interesting example.  Much remains to be explored both with
respect to the vast literature on set-theoretic forcing and, more
relevantly, with respect to Krivine's Classical
Realizability~\cite{KrivineRA}, an extension of forcing.

\paragraph{Polarised logic.}
\label{PolarisationParagraph}

We propose here a model-theoretic study of a rich variety of logical
systems encompassing aspects of resource management and computational
effects (recall Section\pref{ResourcesEffectsModalitiesParagraph}).  This
research constitutes work under discussion with Curien and
Munch-Maccagnoni (Laboratoire~PPS, Universit\'e Paris Diderot - Paris~7).
A novelty of our approach is that this will be pursued as informed by the
logical notion of polarisation.

The notion of explicit polarisation in logical systems was introduced by
Andreoli~\cite{Andreoli} in his study of proof search in Linear Logic, in
particular by focalisation.  According to it, logical connectives are
classified as either being positive or negative; with these two worlds
being dual to each other (categorically by adjunction).  Following
Andreoli's work, Girard understood the relevance of focalisation via
polarisation as a way to tame down the inherent non-determinism in
computation (by cut elimination) in classical logic; and, in this
direction, introduced the first explicitly polarised logic:
\LC~\cite{GirardLC}.  A crucial aspect of this system, is to make eager
(\viz~call-by-value) and lazy~(\viz,~call-by-name) modes of computation
explicit, allowing for their combination in a framework with eager and
lazy data structure.

Polarisation, though not recognised as such, is also present in the
Computational \mbox{$\lambda$-calculi} based on adjunctions as mentioned
in Section\pref{ResourcesEffectsModalitiesParagraph}.  This key
observation leads to a refinement of the
model~(\ref{ResourceEffectAdjunction}) as follows
\begin{equation}\label{LCBPV}
  \begin{minipage}{\columnwidth}
  \xymatrix@R=5pt{
    & & 
    \\
    \ar@{}[u]|-{\text{\scriptsize(negative)}}
    \cat N \ar@/^.75em/[dr] & \ar@{}[rd]|-{\sevdash} \ar@{}[dl]|-{\swvdash} &
    \ar@/^.75em/[dl] \cat M \ar@{}[u]|-{\text{\scriptsize(multiplicative)}}\\
    & \ar@/^.75em/[lu] \cat P \ar@/^.75em/[ur]
    \ar@{}[d]|-{\txt{\scriptsize(positive/linear)}}& 
    \\ & & 
  }
\end{minipage}
\end{equation}
with comonadic resource structure given by the adjunction on the right and
monadic effect structure given by the adjunction on the left.

When the resource structure is trivial, the above restricts to models of
Linear Logic~\cite{MelliesCMLL}; while, when the linear structure
collapses to cartesian structure, one recovers the models of
Call-By-Push-Value~\cite{LevyCBPV}.  One is thus led to the following
programme.
\begin{resenumerate}\setcounter{CC}{0}
\item
  Devise a sound and complete logical system for~(\ref{LCBPV}), and relate
  it to existing polarised systems~\cite{Munch,CurienMunch}.
\item
  A first model-theoretic result shows that the effect structure on $\cat
  P$ lifts to one on $\cat M$.  Show that this corresponds to an encoding
  of Call-By-Push-Value in the devised logical system.
\item
  Develop a generic syntactic theory able to incorporate concrete
  computational effects into the devised calculus.  Two research possibilities
  are: (i)~considering algebraic theories of effects following the work of
  Plotkin and Power~\cite{PlotkinPowerAlgOpsAndGenEffs}, and
  (ii)~revisiting and refining Filinski's result on representing
  monads~\cite{Filinski} that reduces general monadic effects to the
  storage and escape effects.  Either of these approaches will necessarily
  have to overcome serious shortcomings.  For instance, on the one hand,
  there is as yet no general operational theory of algebraic effects; and,
  on the other, a logical system for storage is not yet in place.  These
  problems will be investigated.
\item\label{ItemFour}
  The previous development will provide foundational metalanguages.  The
  next step in the programme is to promote them to programming languages,
  where the effects are implicit, by putting them in so-called
  direct-style.  The kind of model theory involved in this development
  will be discussed below.
\end{resenumerate}

The model theory of~(\ref{LCBPV}) above can be further specialised to
models with enough internal structure so as to extend the picture as
follows
\begin{equation}\label{LEEC}
  \begin{minipage}{\columnwidth}
  \xymatrix@R=5pt{
    & \ar@/^.75em/[dl] \ar@{}[dl]|-\sevdash \cat N \ar@/^.75em/[dr] &
    \ar@{}[rd]|-{\sevdash} \ar@{}[dl]|-{\swvdash} & \ar@/^.75em/[dl] \cat
    M \\
    \ar@/^.75em/[ur] \cat P^\op & & \ar@/^.75em/[lu] \cat P \ar@/^.75em/[ur] & 
  }\end{minipage}
\end{equation}
where the new adjunction between the negatives and the opposite dual of
the positives enriches the models with control structure of the
linear %~\cite{ThieleckeEtAl} 
and/or delimited continuation %~\cite{PeytonJones?}
kind, though this is to be investigated.

The model theory~(\ref{LEEC}) refines that of two recent developments: the
aforementioned Enriched Effect Calculus, which amounts to the case in
which the linear structure collapses to cartesian one; and the Tensor
Logic of Melli\`es and Tabareau~\cite{TensorLogic}, where the effect
structure is collapsed.  Our programme for~(\ref{LCBPV}) will be then also
pursued for~(\ref{LEEC}) drawing connections with these works.

Let us now return to the distinction between metalanguages and programming
languages briefly mentioned in
item~\textbf{\ref{PolarisationParagraph}}\thinspace\itemref{ItemFour} above.
Model theoretical, this can be roughly seen as follows: whereas the
metalanguage is an internal language for co/monadic structure; the programming
language is the internal language for the derived structure of co/free
algebras for the co/monad---typically defined by the categorical co/Kleisli
construction~\cite{MacLane}.  In particular, the Kleisli category of a
computational monad provides a model of call-by-value~\cite{MoggiLambdaC};
while the coKleisli category of a linear exponential comonad provides one
of call-by-name~(\ie,~a cartesian closed category)~\cite{Seely}.

The programming languages corresponding to the metalanguages of the
polarised models~(\ref{LCBPV}) and~(\ref{LEEC}) should be the internal
languages for an analogous, but more intricate, construction relative to
an adjunction rather than a co/monad.  In this context, the following will
be investigated.
\begin{resenumerate}\setcounter{CC}{4}
\item
  What is the mathematical structure of such construction?  The question
  is subtle.  For instance, we envisaged models of {\LC} to arise in this
  manner and, because of the interaction with call-by-value and call-by-name
  modes of computation, the structure should be more general that that of a
  category.
\item
  What would then be the mathematical theory of universal constructions
  for these structures giving rise to logical connectives?  Notice again
  from considering {\LC} that also this will be subtle, requiring
  constructions of objects with mixed polarities.
\item
  What can the resulting direct-style calculi contribute to programming?
  For discussion on this see Section\pref{ProgrammingEffectsParagraph}.
\end{resenumerate}

\hidenote{polarised dialectica for program extraction motivated by
  polarised dialectica category construction}

\hidenote{Investigate relationship to Chuck Liang \& Dale Miller's work}

\hidenote{Investigate relationship to Oleg Kiselyov and Chung-chieh Shan's
  substructural type system for delimited control}

\paragraph{Modal logics.}
\label{ModalLogicsParagraph}

As mentioned in Section\pref{ResourcesEffectsModalitiesParagraph},
resource comonads and effect monads are modalities.  Polarisation in the
context of modal logics has not been considered yet.  We will thus
complement the previous section with the 
\begin{resenumerate}\setcounter{CC}{0}
\item
  Investigation of Polarised Modal Logics.
\end{resenumerate}

Our overall goal here is to
\begin{resenumerate}\setcounter{CC}{1}
\item
  Develop the Propositions-As-Types correspondence for modal logics.
\end{resenumerate}
specially in establishing the categorical model theory vertex of it (see
Figure\,\ref{PAT}), that has been scarcely investigated.

Looking ahead in connection to the proposed research in
Section\pref{MetaprogrammingParagraph}, we will be particularly interested
in studying: 
%
Borghuis' Modal Pure Type Systems~\cite{ModalPTS};
%
Artemov's Logic of Proofs~\cite{ArtemovLP} and related systems;
%(\eg~\cite{AltArtemov,ArtemovIemhoff}); 
%Daniyar S Shamkanov: Strong Normalization and Confluence for Reflexive
%  Combinatory Logic 
%
Mendler's Multimodal CK~\cite{MendlerMMCK};
and
%
Park and Im's Calculus $S_\Delta$~\cite{ParkIm}.

\subsection{Calculi}
\label{CalculiSubsection}

\hidenote{\ldots}

\setcounter{paragraph}{0}
\paragraph*{Categorical Type Theory.}

The evolution of term and type structure in type theories from its origin
to the current state can be succinctly represented as follows
\begin{center}\begin{tabular}{|c||c|c|c|c|}\hline
  & \small Term Structure & \small Type Structure
  \\[.25mm] \hline\hline
  \small Equational Logic & \small Algebraic & \small ---
  \\[.25mm] \hline
  \txt{\\ \raisebox{1mm}{\small Simple Type}\\\raisebox{1mm}{\small Theory}} &
  \raisebox{-.75mm}{\small Binding} & \raisebox{-.75mm}{\small Algebraic}
  \\[.25mm] \hline
  \txt{\\ \raisebox{.5mm}{\small Polymorphic}\\ \raisebox{1mm}{\small Type
      Theory}} & 
  \raisebox{-.75mm}{\small Binding} &
  \raisebox{-.75mm}{\small Binding} 
  \\[1.5mm] \hline
  \txt{\\ \raisebox{.5mm}{\small Dependent}\\\raisebox{1mm}{\small Type Theory}}
  & 
  \multicolumn{2}{c|}{\raisebox{-1mm}{\small Binding}}
  \\[.25mm] \hline
\end{tabular}\end{center}
and it has not changed since de~Bruijn's Automath in the late 1960s.  We
believe that further evolution will come from mathematical input, and in
this direction propose two lines of research for type theories as formal
languages of categorical structures.  These are presented in the two
sections below.

\paragraph{Generalised Type Theory.}
\label{GeneralisedTypeTheoryParagraph}

Section\pref{MethodologyMathematicalUniversesParagraph} considers presheaf
categories in the large as mathematical universes of discourse (or gross
toposes).  There is however an alternative view of them in the small as
spaces (or petit toposes) with morphisms between them given by linear (or
cocontinuous) functors, resulting in the bicategory of profunctors (or
distributors)~\cite{Benabou}.  Lawvere~\cite{LawvereMetric} considered
these structures in the general context of enriched category theory and
synthesised a Generalised Logical Calculus out of them.  The enrichments
give rise to different interpretations, \eg~in preorders, categories, and
metric spaces.

The general goal of our research here is to:
\begin{resenumerate}\setcounter{CC}{0}
\item\label{GeneralisedTypeTheoryItem}
   Develop a type theory providing a Propositions-As-Types interpretation
   of the Generalised Logical Calculus.
\end{resenumerate}
This is to be investigated in two stages, respectively corresponding to
enrichment over cartesian closed categories and over symmetric monoidal
closed categories.  In the first case, one roughly has the following
translation table between the Generalised Logical Calculus and Predicate
Logic.
\begin{center}\begin{tabular}{|c|c|}\hline
  sum & disjunction\\ 
  coend & existential quantification\\ 
  product & conjunction\\ 
  end & universal quantification\\ 
  exponential & implication\\ 
  hom & equality\\ 
  presheaf application & predicate membership\\
  \hline
\end{tabular}\end{center}
The second stage introduces linearity. 

Success in this endevour will yield a system where the mathematical
calculations done with profunctors (\eg~in the context of the coherence of
profunctor composition~\cite{Benabou}, the mathematical theory of
substitution~\cite{FioreFossacs}, generalised species of
structures~\cite{Species}, and generalised polynomial
functors~\cite{FioreICALP}) can be established formally.

The coend and end constructions mentioned in the table above are intuitively
quotients of dependent sums under a compatibility condition and restrictions
of dependent products under a parametricity condition.  Taking this view
seriously, we will aim to:
\begin{resenumerate}\setcounter{CC}{1}
\item
  Extend
  item~\textbf{\ref{GeneralisedTypeTheoryParagraph}}\thinspace\itemref{GeneralisedTypeTheoryItem}
  above to a fully-fledged, possibly higher-dimensional, dependent type
  theory.  
\end{resenumerate}

Finally, in an intriguing complementary direction,
Fiore~\cite{FioreFossacs} has suggested a graphical reading of the
Generalised Logical Calculus that is surprisingly close to formalisms for
categorical graphical languages~\cite{Selinger}, and
proof~\cite{GirardLinearLogic}/interaction~\cite{Lafont} nets.  This
aspect deserves detailed investigation.

\paragraph{Directed Type Theory.}
\label{DirectedTypeTheoryParagraph}

Homotopy Type Theory (under the acronym HoTT, and the slogan `types are
spaces') is the name being used for the body of work at the boundary between
Homotopy Theory and Dependent Type Theory through Identity Types.

The question arises:
\begin{resenumerate}\setcounter{CC}{0}
\item
  Is there a notion of Directed Type connecting Higher-Dimensional Category
  Theory and Dependent Type Theory on which to establish a body of work on
  Directed Type Theory (DiTT)?
\end{resenumerate}
In this context, while an Identity Type
\[
  \Id_T(x,y)
\]
is intended to establish the intensional equality of elements of a type $T$ so
that $\Id_T(x,y)$ and $\Id_T(y,x)$ are equivalent; the intuitive idea behind a
Directed Type 
\[
  \Di_T(x,y)
\]
would be to classify the possible ways in which an element of a type $T$ may
evolve to another one in a not possibly reversible manner; as it happens, for
instance, in computation.  Thereby leading to not necessarily equivalent types
$\Di_T(x,y)$ and $\Di_T(y,x)$, and to the DiTT slogan `types are directed
spaces'.

An analysis of the elimination rule for Identity Types reveals that there
are at least two sources leading to their inherent reversible character.
We mention them below as possible directions of research for investigating
Directed Types.
\begin{resenumerate}\setcounter{CC}{1}
\item
  The use of contexts allowing the commutativity of two consecutive variables
  of the same type, which suggests moving on to a non-commutative setting (as
  \eg~in non-commutative linear logic). %~\cite{NatDedForIntNonCommLinLog}).  
  %(This is related to the considerations in
  %Section\pref{SubstructuralTypeTheoryParagraph} below.)

\item\label{TypeDualityItem}
  The lack of a type duality, a property that seems to be inherent to
  directionality, whereby every type $T$ has an associated dual type $T^\circ$
  for which $\Di_T(x,y)$ and $\Di_{T^\circ}(y,x)$ are equivalent.  
\end{resenumerate}

The previous item is of course related to the considerations of the
previous section.  Pursuing this further, it would be natural to
generalise from presheaves, now regarded as discrete fibrations, to the
general notion of fibration and aim to:
\begin{resenumerate}\setcounter{CC}{3}
\item
  Develop a type theory modelled on Grothendieck fibrations~\cite{SGA1}.
\end{resenumerate}

As a final point, we note that yet another approach would be to:
\begin{resenumerate}\setcounter{CC}{4}
\item
  Investigate a directed-cylinder construction, intuitively adding a new base
  point to a type that canonically evolves to every element.  
\end{resenumerate}
This would generalise the lifting construction of Domain
Theory~\cite{LiftingKZ} and possibly establish connections with the
Complete Cuboidal Sets model of Axiomatic Domain
Theory~\cite{CompleteCuboidalSets} opening new directions.  

\paragraph{Substructural Type Theory.}
\label{SubstructuralTypeTheoryParagraph}

The work of Bellantoni and Cook~\cite{BellantoniCook}
%, after Leivant~\cite{Leivant}, 
on implicit computational complexity gave a characterisation of the
polynomial-time computable functions on finite strings by means of a
safe-recursion scheme that restricted primitive recursion.  A crucial
ingredient of their approach is to divide function arguments into normal and
safe ones.  Recursion is allowed on normal arguments but not on safe ones.  In
the terminology that we have been using, normal arguments are linear while
safe ones are cartesian.  In fact, the notion of safe composition
coincides with that for mixed linear/cartesian operads (see
item~\textbf{\ref{WiringStructureParagraph}}\thinspace\itemref{BellantoniCook}).
%on page~\pageref{BellantoniCook}).  
Since then, the role of resource management in taming complexity has been
prominently recognised, and a big body of work on implicit computational
complexity (\eg~on soft, bounded, light, and linear logics and types) has
followed.
% Baillot, Dal Lago, Hofmann, Lafont

Recently, Beckmann, Buss, and Friedman~\cite{BeckmannBussFriedman} 
%and later on Arai~\cite{Arai} 
have extended the safe-recursion scheme to arbitrary set functions.
Changing viewpoint form the mathematical universe of sets to that of
types, one is lead to:
\begin{resenumerate}\setcounter{CC}{0}
\item
  Extend dependent type theory to the mixed linear/cartesian setting.
\end{resenumerate}
In particular, in what concerns to taming computational complexity:
\begin{resenumerate}\setcounter{CC}{1}
\item
  Investigate a safe version of the Induction-Recursion scheme of
  Dybjer~\cite{DybjerIR}.  
\end{resenumerate}
This development should be also accompanied by a model theory, possibly based
on 
\begin{resenumerate}\setcounter{CC}{2}
\item 
  Realisability models relative to the Linear Combinatory Algebras of
  Abramsky, Haghverdi, and Scott~\cite{AHS}.
\end{resenumerate}

\subsection{Programming}
\label{Programming}

This track of the proposal is concerned with the design and implementation
of programming languages from first principles and pragmatics, followed up
by their subsequent test and use.

Three directions for this research are presented.  These will be
considered as research units in their own right, and also in relation to
each other.

\setcounter{paragraph}{0}
\paragraph{Indexed programming.}
\label{IndexedProgrammingParagraph}

This section outlines work under discussion with Ahn and Sheard
(Department of Computer Science, Portland State University) on indexed
programming (recall Section\pref{IndexedProgrammingIntro}).

The main problem to be addressed here is to:
\begin{resenumerate}\setcounter{CC}{0}
\item
  Design a language for programming indexed data structures that will be
  archetypal for the next generation of such languages.
\end{resenumerate}
The task can be approached from two different viewpoints.  In a top-down
fashion, one may turn dependent type theories into programming languages;
conversely, in a bottom-up fashion, one may extend functional programming
languages with indexing structure.  These two approaches pull the
programming language design into two opposite directions.  This is
pictorially presented in Figure\,\ref{DesignSpaceFigure}, 
\begin{figure}[h]
\caption{Indexed-programming design space}
\[
\null\hspace*{-2mm}
\txt{\small Functional\\ \small Programming\\ \small Language}
\hspace*{3mm}\xy
0;/r.75pc/:+(0,.25)
,{\hloop-\hcross\hcross\hcross\hcross\hcross\hcross\hcross\hloop}
\endxy\null\hspace*{3mm}
\txt{\small Constructive\\ \small Proof\\ \small Assistant}
\]
\hdashrule[1ex]{\columnwidth}{1pt}{2.5pt}
\begin{tabular}{lcl}
%\scriptsize Programming && \scritpsize Proof\\ 
\small
ML, %~\cite{SML}, 
Haskell %~\cite{Haskell} 
&\hspace*{25mm}&
\small
Coq, %~\cite{Coq}, 
Agda %~\cite{Agda}
\\[-.5mm]
\small
\SystemFomega %~\cite{GirardSystemF} 
&& 
\small
MLTT, %~\cite{MLTT}, 
ICC %~\cite{ICC}
\\[-.5mm]
\small
Polymorphism %~\cite{Strachey1967} 
&& 
\small
Type dependency %~\cite{MartinLofTT} 
\\[-.5mm]
\small
Impredicativity && 
\small
Universes %~\cite{Universes}
\\[-.5mm]
\small
Type inference %~\cite{Milner1978,Wells} 
&&
\small
Type checking
\\[-.5mm]
\small
Recursive types %~\cite{FPC}
&& 
\small
Inductive types~ %\cite{InductiveFamilies}
\\[-.5mm]
\small
Equality types %~\cite{GHC-IFP2012} 
&& 
\small
Identity types %~\cite{IdTypes}
\end{tabular}
\label{DesignSpaceFigure}
\end{figure}
where the main features inherent to each approach are also listed.

Most of the work in this area has concentrated on exploring the top-down
approach (see~\eg~\cite{Cayenne} and~\cite{Epigram}).  Here we will rather
pursue research on the bottom-up approach, which has not been explored
systematically.  There is a strong pragmatic reason for this: our outmost
interest is in a language targeted to programmers.  However, this should not
be taken to mean that we are not interested in the activity of proving as in
constructive proof assistants; but indeed our main goal, under the
Propositions-As-Types paradigm, is to: 
\begin{resenumerate}\setcounter{CC}{1}
\item
  Prove as a by-product of programming; rather than extract programs from
  proving.
\end{resenumerate}
For this to be meaningful, both in theory and practice, the language will
need to guarantee logical consistency in a setting that naturally allows
programming strong invariants for rich indexing type structure.  For
instance, so that the code of a compiler guarantees its correctness.

In the above direction, we are investigating:
\begin{resenumerate}\setcounter{CC}{2}
\item
  {\SystemFi}: an extension of {\SystemFomega} with first-class static
  type-indexing structure.
\end{resenumerate}
which is to serve as the mathematical foundation for designing, and giving
operational semantics to, our programming language.  We stress here that
indices are static, \ie~determined at compile time.  This is a main
feature pulling us away from traditional dependently-typed formalisms.

The logical consistency of {\SystemFi} amounts to establishing its strong
normalisation; while as an extension of {\SystemFomega} it will embody
rich type structure.  In this setting, following Ahn and
Sheard~\cite{AhnSheard}, we will:
\begin{resenumerate}\setcounter{CC}{3}
\item
  Study programming primitives corresponding to a variety of induction
  proof principles.
\end{resenumerate}
These are to be incorporated in the programming language design.  Here, it
is interesting to note that the scheme provided by Mendler's
iterator~\cite{MendlerIter}, that crucially takes advantage of
parametric polymorphism, allows for terminating iteration over recursively
defined datatypes of possibly mixed variance.  Thus, going beyond the
typical inductive types of type theory stemming from Dybjer's Inductive
Families~\cite{DybjerIF}.  It would be interesting to establish a formal
connection between these approaches so that they can inform each other;
in particular in the context of the termination checkers. %~\cite{AbelFoetus}. 

Dually, we will also:
\begin{resenumerate}\setcounter{CC}{4}
\item
  Develop programming primitives corresponding to coinduction proof
  principles.
\end{resenumerate}
which have received less attention.

The above design considerations are to be shaped by the following maxim:
\begin{resenumerate}\setcounter{CC}{5}
\item
  Design language constructs with minimal type annotation supporting
  maximal type inference.
\end{resenumerate}
This is a main open problem in dependently-typed programming language
theory.

Finally, let us mention an intriguing possibility suggested by the
model-theoretic considerations of
Section\pref{MethodologyMathematicalUniversesParagraph}.  The indexing
structure in dependent type theory is discrete.  In many applications,
however, one is interested in indexed types that furthermore relate
structure on indices to structure on the associated indexed family of types.
The question arises as to how to:
\begin{resenumerate}\setcounter{CC}{6}
\item
  Build idioms or language abstractions for defining and then programming
  with indexed types equipped with internal varying structure.
\end{resenumerate}
A specific motivation for this comes from the possibility of directly
programming, as opposed to having to code, presheaf structure;~\eg~as it is
needed in Normalisation by Evaluation~\cite{FiorePPDP}.

\paragraph{Effects.}
\label{ProgrammingEffectsParagraph}

Our main aim here is to consider the model-theoretic investigations of
Section\pref{PolarisationParagraph} from a proof-theoretic viewpoint and
percolate this down to programming language theory.

A first main novelty in our approach is that the proof theory to be
considered is based on sequent calculus, rather than the traditional line
followed so far in programming language foundations based on sequent-style
natural deduction.

The proof-theoretic formalism that we will be adhering to is {\SystemL},
as recalled in Section\pref{SequentCalculiParagraph}.  A first main
question that arises is:
\begin{resenumerate}\setcounter{CC}{0}
\item 
  What is the programming paradigm stemming from sequent calculi in
  general, and {\SystemL} in particular?
\end{resenumerate}
For instance, 
\begin{resenumerate}\setcounter{CC}{1}
\item
  Does the inherent symmetry of {\SystemL} lead to a new programming
  style?

\item
  How can the close connection between {\SystemL} and abstract machines be
  exploited in programming?
\end{resenumerate}
These investigations will also require mathematical principles to be
developed.  Specifically, we will 
\begin{resenumerate}\setcounter{CC}{3}
\item
  Establish the Propositions-As-Types axis corresponding to internal
  languages (see Figure\,\ref{PAT}) relating {\SystemL} to adjoints.
\end{resenumerate}

As for polarisation, we believe that programmers will be able to
intuitively assimilate the eager {\vs}~lazy modes of computation and data
structure underlying it; but, pragmatically,
\begin{resenumerate}\setcounter{CC}{4}
\item
  How will a programmer be able to easily code polarisation in a
  programming language?
\end{resenumerate}

Once the above is sorted, one can consider experimenting with programming
languages based on the various systems of
Section\pref{PolarisationParagraph} further enriched with computational
effects.  To this end, there is a large body of theoretical work on
effects that needs to be examined, evaluated, and reconsidered.  An
interesting stepping stone here is the recent work of Bauer and
Pretnar~\cite{BauerPretnar} on the experimental programming language Eff.  

Eff is based on the algebraic theory of effects and handlers of Plotkin
\etal~\cite{PlotkinPowerAlgOpsAndGenEffs,PlotkinPretnar};
but, for the pragmatics of supporting seemingly non-algebraic control
operators, it goes beyond algebraic models.  This is indeed so with
respect to (first-order) algebraic theories.  However, as it has
transpired in conversation between Fiore and Staton, the consideration of
the richer Second-Order Algebraic Theories of Fiore
\etal~\cite{FioreHur,FioreMahmoud} seems to enrich the class of
specifiable effects to incorporate aspects of control.  Thus, our research
into this topic will also
\begin{resenumerate}\setcounter{CC}{5}
\item
  Investigate Second-Order (and, when they are developed, Polymorphic)
  Algebraic Theories as a mathematical basis for an algebraic theory of
  effects encompassing control.
\end{resenumerate}
Targeted goals here will be program logics for effects and control, and
formal correctness proofs of (continuation and/or abstract machine)
implementations of effects.

\hidenote{Dependent CBPV}

\paragraph{Metaprogramming.}
\label{MetaprogrammingParagraph}

\note{generic programming}

\note{reflection}

\note{TDPE with sums: reflection with delimited control operators}

\note{System $F_\omega^*$: Tillmann Rendel, Klaus Ostermann, Christian
  Hofer (2009),Typed Self-Representation}

\hidenote{\SystemL internalisation of stacks?}

\hidenote{ssreflect?}

\section{\underline{Resources}}
\label{ResourcesSection}

\paragraph*{Team.}

During the 2011-12 academic year, I was on sabbatical leave.  Two
important events for this grant proposal happen then as follows.  

On the one hand, I was awarded a \emph{Research in Paris} grant from the
\emph{programme d'accueil des chercheurs \'etrangers de la Ville de Paris}
to visit Laboratoire PPS at Universit\'e Paris Diderot - Paris~7 for three
months.  There, I mainly interacted with Pierre-Louis Curien and his PhD
student Guillaume Munch-Maccagnoni, establishing the Mathematical Logic
axis of Figure\,\ref{ercTeam}.

On the other hand, Timothy Sheard (Department of Computer Science,
Portland State University), who was also on sabbatical leave during the
2011-12 academic year, visited Microsoft Research Cambridge during
October--December 2011 together with his PhD student Ki Yung Ahn who
instead visited the Computer Laboratory (University of Cambridge).  The
Programming Language axis of Figure\,\ref{ercTeam} was established then.

To complete the team, for the Type Theory axis of Figure\,\ref{ercTeam} it
was most natural to recruit Peter Dybjer (Department of Computer Science
and Engineering, Chalmers University of Technology) and Nicola Gambino
(Dipartimento di Matematica e Applicazioni, Universit\`a degli Studi di
Palermo).  
%held a \ldots at DPMMS, University of Cambridge, during during
%????--???? and we had then collaborated on~\cite{Species}.  
%Since then, we have also kept in contact; most recently regarding our
%common interest on the topic of dependent polynomial
%functors~\cite{GambinoHylad,GambinoKock,FioreICALP} which is also
%relevant to topics of the proposal.  

\note{Makoto --- Marco --- Michael}

\paragraph*{Principal investigator.}

As Principal Investigator, \textbf{\em Marcelo Fiore} would be concerned
with all areas of the project and head the {\erc} team, comprising both
Senior and Junior Researchers.  
\begin{myitemize}
\item
For the Principal Investigator, salary is requested for eight months per
year for the five years of the duration of the project.  
\end{myitemize}
The reduction acknowledges that the Principal Investigator will concentrate on the project while continuing with some teaching to attract new PhD students and still have some departmental administration duties.

\paragraph*{Senior researchers.}

The Senior Researchers Pierre-Louis Curien, Peter Dybjer, and Timothy
Sheard are renown researchers in their respective field of expertise.  
\begin{myitemize}
\item
For the Senior Researchers, funding is requested for annual one-month research
visits to the Computer Laboratory (University of Cambridge) for the five years
of the duration of the project.  
\end{myitemize}
Our collaboration will however go beyond each visit; continuing by email,
conference calls, and/or visits to their respective sites. 
%(for which travel funding is also requested).  
Their academic credentials together with the specific areas of the project
to which they will be engaged follow.

\myparagraph{Pierre-Louis Curien.}\ldots\\
\begin{tabular}{l|l}
Section & Keyword
\\ \hline
(\ref{WiringStructureParagraph}) & operads
\\
(\ref{IntensionalTypeTheoryParagraph}) & dependent types, parametricity
\\
(\ref{PolarisationParagraph}) & polarisation
\\
(\ref{ModalLogicsParagraph}) & Linear Logic
\\
(\ref{ProgrammingEffectsParagraph}) & \SystemL
\\
\end{tabular}

\myparagraph{Peter Dybjer.}\ldots\\
\begin{tabular}{l|l}
Section & Keyword
\\\hline
(\ref{AlgebraicTypeTheoryParagraph}) & Internal Type Theory
\\
(\ref{IntensionalTypeTheoryParagraph}) & Dependent Type Theory
\\
(\ref{MethodologyMathematicalUniversesParagraph}) & NbE, glueing
\\
(\ref{SubstructuralTypeTheoryParagraph}) & induction-recursion
\\
(\ref{IndexedProgrammingParagraph}) & Agda
\\
\end{tabular}

\myparagraph{Timothy Sheard.}\ldots\\
\begin{tabular}{l|l}
Section & Keyword
\\\hline
(\ref{ModalLogicsParagraph}) & MetaML
\\
(\ref{SubstructuralTypeTheoryParagraph}) & iterators
\\
(\ref{IndexedProgrammingParagraph}) & Nax, Omega
\\
(\ref{ProgrammingEffectsParagraph}) & PL design \& implementation
\\
(\ref{MetaprogrammingParagraph}) & MetaML
\\
\end{tabular}

\paragraph*{Junior researchers.}

The Junior Researchers Ki Yung Ahn, Nicola Gambino, and Guillaume
Munch-Maccagnoni will be based at the Computer Laboratory (University of
Cambridge), working in the project full time.  Ahn and Munch-Maccagnoni
are currently finishing their PhD dissertations.  Nicola Gambino is
already an established researcher.  All would be ready to join the project
as it starts.  
\begin{myitemize}
\item 
  Salary is requested to employ each Junior Researcher full-time for the five
  years of duration of the project.
\end{myitemize}

\myparagraph{Ki Yung Ahn.}\ldots\\
\begin{tabular}{l|l}
Section & Keyword
\\\hline
(\ref{SubstructuralTypeTheoryParagraph}) & iterators
\\
(\ref{IndexedProgrammingParagraph}) & Nax
\\
(\ref{ProgrammingEffectsParagraph}) & Haskell
\\
\end{tabular}

\myparagraph{Nicola Gambino.}\ldots\\
\begin{tabular}{l|l}
Section & Keyword
\\\hline
(\ref{WiringStructureParagraph}) & generalised species
\\
(\ref{IntensionalTypeTheoryParagraph}) & HoTT
\\
(\ref{MethodologyMathematicalUniversesParagraph}) & sheaves, Algebraic Set
Theory
\\
(\ref{GeneralisedTypeTheoryParagraph}) & generalised species
\\
(\ref{DirectedTypeTheoryParagraph}) & HoTT
\\
\end{tabular}

\myparagraph{Guillaume Munch-Maccagnoni.} \ldots\\
\begin{tabular}{l|l}
Section & Keyword
\\\hline
(\ref{PolarisationParagraph}) & polarisation
\\
(\ref{ModalLogicsParagraph}) & Linear Logic
\\
(\ref{ProgrammingEffectsParagraph}) & \SystemL, delimited control
\\
\end{tabular}

\paragraph*{Additional costs.}

\begin{myitemize}
\item
Travel funding is requested for attending workshops and 
conferences, \hidefootnote{\Eg~CALCO, CIE, CSL, CT, ICALP, ICFP, LICS, MFPS,
  POPL, PPDP, PSSL, RTA, TLCA, TYPES, WOLLIC.}, 
and for visiting and/or inviting researchers. \hidefootnote{\Eg~Aarhus
  (A\,Kock); 
  Bamberg (Mendler);
  Barcelona (J\,Kock);
  Bath (Power); 
  Berkeley (D\,Scott); 
  Birmingham (Escard\'o, Levy); 
  Bologna (Asperti, Dal Lago);
  Buffalo (Lawvere);
  Copenhagen (Birkedal, Filinski); 
  Darmstadt (Streicher); 
  Dublin (Dotsenko);
  Edinburgh (Leinster, Plotkin, A\,Simpson); 
  Genova (Moggi, Rosolini); 
  G\"oteborg (Coquand, Dybjer);
  Kiryu (Hamana);
  Ljubljana (Bauer, Pretnar);
  London (Oliva);
  Manchester (Aczel);
  Marseille (Girard, Lafont, Regnier); 
  Montreal (Joyal, Makkai, Panangaden);
  Munich (Abel, Hofmann);
  Nice (C\,Simpson, Vallette);
  Nijmegen (Moerdijk);
  Nottingham (Altenkirch);
  Ottawa (P\,Scott); 
  Oxford (Abramsky, Coecke, Doring); 
  Paris (Burroni, Curien, Herbelin, Krivine, Melli\`es, Metayer);
  Philadelphia (Freyd, Pierce, Weirich);
  Pittsburgh (Avigad, Awodey, Harper, Pfenning, Reynolds);
  Portland (Sheard);
  Princeton (Warren);
  Strathclyde (Ghani, McBride);
  Swansea (Beckmann, Berger)
  Sydney (Garner, Lack, Street).}.
\item
Further funds are requested for computer support in the form of laptops
for the Junior Researchers and the cost of broadband for the Principal
Investigator.
\end{myitemize}

{\setstretch{0}\footnotesize
\begin{thebibliography}{000}
\bibitem{AHS}
S.\,Abramsky, E.\,Haghverdi and P.\,Scott (2002). 
\newblock Geometry of Interaction and linear combinatory algebras. 
%\newblock \emph{Mathematical Structures in Computer Science} 12:625--665.
\newblock \emph{MSCS} 12.

\bibitem{AguiarMahajan}
M.\,Aguiar and S.\,Mahajan (2010).
\newblock \emph{Monoidal functors, species, and Hopf algebras}.
\newblock CRM Monograph Series, AMS.

\bibitem{AhnSheard}
K.Y.\,Ahn and T.\,Sheard (2011).
%\newblock A Hierarchy of Mendler style Recursion Combinators: Taming
%  Inductive Datatypes with Negative Occurrences.  
\newblock A Hierarchy of Mendler style Recursion Combinators. 
\newblock In \emph{ICFP'11}.  

\bibitem{Albert}
A.A.\,Albert (1948).
\newblock Power-associative rings.
\newblock \emph{Trans.\ Amer.\ Math.\ Soc.}, 64(3):552--593.

%\bibitem{AltArtemov}
%J.\,Alt and S.\,Artemov (2001).
%\newblock Reflective $\lambda$-calculus.
%\newblock In \emph{PTCS'01}, LNCS 2183, \pp\,22--37.

\bibitem{Andreoli}
J.-M.\,Andreoli (1992).
\newblock Logic programming with focusing proof in linear logic.
%\newblock \emph{Journal of Logic and Computation}, 2(3):297--347.
\newblock \emph{J.\ of Logic and Computation}, 2:297--347.

\bibitem{ArtemovLP}
S.\,Artemov (2001).
\newblock Explicit provability and constructive semantics.
\newblock \emph{Bulletin for Symbolic Logic}, 7(1):1--36.

%\bibitem{ArtemovIemhoff}
%S.\,Artemov and R.\,Iemhoff (2007).
%\newblock The Basic Intuitionistic Logic of Proofs.
%\newblock \emph{Journal of Symbolic Logic}, 72(2):439--451.

\bibitem{SGA4}
M.\,Artin, A.\,Grothendieck, and J.-L.\,Verdier, eds. (1972).
\newblock \emph{SGA~4}, LNM 269. 

\bibitem{AwodeyHofstraWarren}
S.\,Awodey, P.\,Hofstra and M.\,Warren (2012).
\newblock Martin-L\"of complexes.
\newblock In \url{http://arxiv.org/abs/0906.4521}.

\bibitem{Cayenne}
L.\,Augustsson (1998).
\newblock Cayenne --- a language with dependent types.
\newblock In \emph{ICFP'98}, pp\,239--250.

\bibitem{BarberPlotkin}
A.\,Barber (1996).
\newblock Dual Intuitionistic Linear Logic. 
\newblock PhD thesis, University of Edinburgh. 

\bibitem{BauerPretnar}
A,\,Bauer and M.\,Pretnar (2012).
\newblock Programming with Algebraic Effects and Handlers.
%\newblock In \url{http://arxiv.org/abs/1203.1539}.

\bibitem{BeckThesis}
J.\,Beck (1967).
\newblock \emph{Triples, algebras and cohomology}.
\newblock \emph{Reprints in TAC} 2, 2003. 

\bibitem{BeckmannBussFriedman}
A.\,Beckmann, S.\,Buss, and S.\,Friedman (2012).
\newblock Safe Recursive Set Functions.
%\newblock Available on-line.

\bibitem{BeesonBook}
M.\,Beeson (1984).
\newblock Foundations of constructive mathematics.
\newblock Springer Verlag.

\bibitem{BellantoniCook}
S.\,Bellantoni and S.\,Cook (1992).
\newblock A new recursion-theoretic characterisation of the polytime
functions.
\newblock \emph{Computational Complexity}, 2:97--110.

\bibitem{Benabou}
J.~B{\'e}nabou (1967). 
\newblock \emph{Introduction to bicategories}. 
\newblock In LNM 47, pp\,1--77. 

\bibitem{BentonWadler}
N.\,Benton and P.\,Wadler (1996). 
\newblock Linear logic, monads, and the lambda calculus.
\newblock In \emph{11th LICS}.

\bibitem{Birkhoff}
G.\,Birkhoff (1935).
\newblock On the structure of abstract algebras.
\newblock {\em P.\ Camb.\ Philos.\ Soc.}, 31:433--454. 
  
\bibitem{BoehmBerarducci}
C.\,Boehm and A.\,Berarducci (1985). 
\newblock Automatic Synthesis of Typed Lambda-Programs on Term Algebras.
%\newblock \emph{Theoretical Computer Science}, 39:135--154.
\newblock \emph{TCS} 39.

\bibitem{Boole}
G.\,Boole (1847).
\newblock \emph{The Mathematical Analysis of Logic}.%, Being an Essay Towards
  %a Calculus of Deductive Reasoning}. 
\newblock Macmillan, Barclay, \& Macmillan. 

\bibitem{ModalPTS}
T\,Borghuis (1998).
\newblock Modal Pure Type Systems: Type Theory for Knowledge
Representation.  
\newblock \emph{Journal of Logic, Language, and Information}, 7:265--296. 

\bibitem{ChapotonLivernet}
F.\,Chapoton and M.\,Livernet (2001).
\newblock Pre-Lie algebras and the rooted trees operad.
\newblock \emph{Int.\ Math.\ Res.\ Not.}, 8:395--408.

\bibitem{Church1936}
A.\,Church (1936).
\newblock An unsolvable problem of elementary number theory.
\newblock \emph{American Journal of Mathematics}, 58:345--363. 

\bibitem{Church1940}
A.\,Church (1940).
\newblock A formulation of the simple theory of types.
\newblock {\em J.\,Symbolic Logic}, 5:56--68. 
  
\bibitem{Cohen}
P.\,Cohen (1966).
\newblock \emph{Set theory and the continuum hypothesis}.
\newblock W A Benjamin.

\bibitem{CoquandNote}
T.\,Coquand and G.\,Jaber (2012).
\newblock A note on forcing and type theory.
\newblock \emph{Fundamenta Informatica} 100:43--52.

\bibitem{CoquandHuet}
T.\,Coquand and G.\,Huet (1988). 
\newblock The Calculus of Constructions. 
\newblock \emph{Information and Computation}, 76(2--3).

\bibitem{CurienHerbelin}
P.-L.\,Curien and H.\, Herbelin (2000).
\newblock The duality of computation.
\newblock In \emph{ICFP'00}, pp\,233--243.

\bibitem{CurienMunch}
P.-L.\,Curien and G.\,Munch-Maccagnoni (2010).
\newblock The duality of computation under focus.
\newblock In \emph{IFIP TCS}.

\bibitem{Curry1934}
H.\,Curry (1934).
\newblock Functionality in Combinatory Logic.
\newblock In \emph{Proceedings of the National Academy of Sciences},
  20:584--590.

\bibitem{Day}
B.\,Day (1970).
\newblock On closed categories of functors.
%\newblock In LNM 137, pp\,1--38.
\newblock In LNM 137. 

\bibitem{deBruijn}
N.G.\,de~Bruijn (1968).
\newblock Automath, a language for mathematics.
%\newblock In \emph{Automation and Reasoning}, vol\,2, Classical papers on
%Computational Logic 1967--1970, pp\,159--200, 1983.
\newblock In \emph{Automation and Reasoning}, pp\,159--200, 1983.

\bibitem{DybjerIF}
P.~Dybjer (1994).
\newblock Inductive families.
\newblock \emph{Formal Aspects of Computing}, 6:440--465.

\bibitem{DybjerITT}
P.~Dybjer (1996).
\newblock Internal Type Theory.
\newblock In \emph{TYPES}, LNCS 1158, pp\,120--134.

\bibitem{DybjerIR}
P.\,Dybjer (2000).
\newblock A General Formulation of Simultaneous Inductive-Recursive
Definitions in Type Theory. 
%\newblock \emph{J.\ Symb.\ Log.}, 65(2):525--549.
\newblock \emph{JSL} 65.

\bibitem{EEC}
J.\,Egger, R.\,M{\o}gelberg, and A.\,Simpson (2009). 
\newblock Enriching an Effect Calculus with Linear Types. 
\newblock In \emph{CSL'09}.

\bibitem{EilenbergMacLane}
S.\,Eilenberg and S.\,Mac Lane (1945).
\newblock General Theory of Natural Equivalences.
\newblock \emph{Trans.\ Amer.\ Math.\ Soc.},
  58(2):231--294. 

\bibitem{Filinski}
A.\,Filinski (1994).
\newblock Representing monads.
%\newblock In \emph{POPL'94}, pp\,446--457.
\newblock In \emph{POPL'94}.

\bibitem{LiftingKZ}
M.\,Fiore (1995).  
\newblock Lifting as a KZ-doctrine.  
%\newblock In \emph{CTCS'96}, LNCS 953, pp.\,146--158.  
\newblock In \emph{CTCS'96}.

\bibitem{FiorePPDP}
M.\,Fiore (2002). 
\newblock Semantic analysis of Normalisation By Evaluation for typed
lambda calculus.  
\newblock In \emph{PPDP'02}.

\bibitem{FioreFossacs}
M.\,Fiore (2005).  
\newblock Mathematical models of computational and combinatorial
structures.  
%\newblock In \emph{FOSSACS'05}, LNCS 3441, pp\,25--46.  
\newblock In \emph{FOSSACS'05}, LNCS 3441. 

\bibitem{FioreMFPS}
M.\,Fiore (2006).
\newblock On the structure of substitution.  
\newblock MFPS XXII invited address. 

\bibitem{FioreICALP}
M.\,Fiore (2012).   
\newblock Discrete Generalised Polynomial Functors.  
\newblock In \emph{ICALP'12}, LNCS 7392, pp\,214--226.

\bibitem{FioreRemarks}
M.\,Fiore, R.\,Di~Cosmo and V.\,Balat (2002). 
\newblock Remarks on isomorphisms in typed lambda calculi with empty and
sum types.  
\newblock In \emph{LICS'02}, pp\,147-156.  

\bibitem{Species}
M.\,Fiore, N.\,Gambino, M.\,Hyland, and G.\,Winskel (2008).   
\newblock The cartesian closed bicategory of generalised species of
structures.
\newblock \emph{J.\ London Math.\ Soc.}, 77:203-220. 

\bibitem{FioreHur}
M.\,Fiore and C.-K.\,Hur (2010).   
\newblock Second-order equational logic.  
\newblock In \emph{CSL'10}, LNCS 6247, pp\,320--335. 
  
\bibitem{FioreMahmoud}
M.\,Fiore and O.\,Mahmoud (2010).   
\newblock Second-order algebraic theories.  
\newblock In \emph{MFCS'10}, LNCS 6281, pp\,368--380. 

\bibitem{CompleteCuboidalSets}
M.\,Fiore, G.\,Plotkin and A.J.\,Power (1997). 
\newblock Complete cuboidal sets in axiomatic domain theory.  
%\newblock In \emph{LICS'97}, pp.\,268--279.  
\newblock In \emph{LICS'97}.

\bibitem{Frege1879}
G.\,Frege (1879).
\newblock \emph{Begriffsschrift, a formula language, modeled upon that of
  arithmetic, for pure thought}. 
\newblock In \emph{\cite{vanHeijenoort}}, pp\,1--82.

\bibitem{Frege1903}
G.\,Frege (1893/1903).
\newblock 
\emph{Grundgesetze der Arithmetik}.%\footnote{Basic laws of arithmetic.}
%\newblock Vol.\,I Jenna, 1893.  Vol.\,II 1903.
\newblock Vols.\,I/II.

\bibitem{Gentzen1935}
G.\,Gentzen (1935). 
\newblock Untersuchungen \"uber das logische Schlie{\ss}en. 
\newblock Mathematische Zeitschrift 39:176--210.

\bibitem{Gerstenhaber}
M.\,Gerstenhaber (1963).
\newblock The cohomology structure of an associative ring.
\newblock \emph{Annals of Mathematics}, 78(2):267--288.

\bibitem{GirardSystemF}
J.-Y.\,Girard (1972).
\newblock \emph{Interpr\'{e}tation Fonctionnelle et \'{E}li\-mi\-na\-tion des
  Coupures de l'Arithm\'{e}tique d'Ordre Sup\'{e}rieur}.
\newblock Th\`{e}se de doctorat d'\'{e}tat, Universit\'{e} Paris VII. 

\bibitem{GirardLinearLogic}
J.-Y.\,Girard (1987). 
\newblock Linear logic.
%\newblock \emph{Theoretical Computer Science}, 50:1--101.
\newblock \emph{TCS}, 50:1--101.

\bibitem{GirardLC}
J.-Y.\,Girard (1991). 
\newblock A new constructive logic: Classical logic.
\newblock \emph{Math.\ Struct.\ Comp.\ Sci.}, 1(3).

\bibitem{SGA1}
A.\,Grothendieck and M.\,Raynaud (1971).
%\newblock Rev\^etements \'etales et groupe fondamental.
%\newblock LNM 224.
\newblock \emph{SGA~1}, LNM 224.

\bibitem{Hilbert}
D.\,Hilbert (1927).
\newblock The foundations of mathematics.
\newblock In \emph{\cite{vanHeijenoort}}, pp\,464--479.

\bibitem{Hindley1969}
R.\,Hindley (1969).
\newblock The Principal Type-Scheme of an Object in Combinatory Logic.
\newblock \emph{Trans.\ Amer.\ Math.\ Soc.}, 146:29--60.

%\bibitem{Hofmann}
%M.\,Hofmann (1997).
%\newblock Syntax and semantics of dependent types. 
%\newblock In \emph{Semantics and Logics of Computation}, pp\,79--130. 
%%\newblock Cambridge University Press.

\bibitem{HS}
M.\,Hofmann and T.\,Streicher (1998).
\newblock
The groupoid interpretation of type theory. 
\newblock In \emph{Twenty-five years of constructive type theory}, pp\,83--111.
%\newblock Oxford University Press. 

\bibitem{Howard1969}
W.\,Howard (1969).
\newblock The formulae-as-types notion of construction.
\newblock In \emph{\cite{ToHBCurry}}, pp\,479--490. 

\bibitem{TypeTheoryWithForcing}
G.\,Jaber, N.\,Tabareau and M.\,Sozeau (2012).
\newblock Extending Type Theory with Forcing.
\newblock In \emph{LICS'12}, pp\,395--404.

\bibitem{Elephant}
P.\,Johnstone (2002).
\newblock Sketches of an Elephant: A Topos Theory Compendium. 
\newblock Oxford University Press.

\bibitem{JoyalAdvMath}
A.\,Joyal (1981). 
\newblock Une th\'eorie combinatoire des s\'eries formelles. 
\newblock \emph{Advances in Mathematics}, 42:1--82.

\bibitem{JoyalLNM1234}
A.\,Joyal (1986). 
\newblock Foncteurs analytiques et esp\`eces de structures. 
\newblock \emph{Combinatoire \'Enum\'erative}, LNM 1234, pp\,126--159.

\bibitem{Kan}
D.\,Kan (1958).
\newblock Adjoint functors.
%\newblock \emph{Trans.\ Amer.\ Math.\ Soc.}, 87:294--329.
\newblock \emph{Trans.\ Amer.\ Math.\ Soc.} 87.

\bibitem{KellyBook}
G.M.\,Kelly (1982).
\newblock \emph{Basic Concepts of Enriched Category Theory}.
\newblock \emph{Reprints in TAC} 10, 2005.

\bibitem{KellyOperads}
G.M.\,Kelly (1972).
\newblock On the operads of J.P.\,May.
\newblock \emph{Reprints in TAC} 13, 2005.

\bibitem{Kobayashi}
S.\,Kobayashi (1997).
\newblock Monad as modality.
\newblock In \emph{TCS}, 175:29--74.

\bibitem{KrivineRA}
J.-L.\,Krivine (2011).
\newblock Realizability algebras: a program to well order R.
\newblock \emph{LMCS}-7(3:2).

\bibitem{Lafont}
Y.\,Lafont (1990).
\newblock Interaction nets.
\newblock In \emph{POPL'90}.

\bibitem{LambekI}
J.\,Lambek (1968).
\newblock Deductive systems and categories I.
\newblock \emph{J.\ Math.\ Systems Theory}, 2:278--318.

\bibitem{LambekScott}
J.\,Lambek and P.\,Scott (1986).
\newblock \emph{Introduction to Higher Order Categorical Logic}.
\newblock Cambridge University Press.

\bibitem{LawvereThesis}
F.W.\,Lawvere (1963). %\,\&\,1968).
\newblock Functorial Semantics of Algebraic Theories. %and Some Algebraic
  %Problems in the context of Functorial Semantics of Algebraic Theories.
\newblock \emph{Reprints in TAC} 5, 2004.
  
\bibitem{LawvereMetric}
F.W.\,Lawvere (1973).
\newblock Metric spaces, generalized logic and closed categories.
\newblock \emph{Reprints in TAC} 1, 2002.

\bibitem{Leinster}
T.\,Leinster (2004). 
\newblock \emph{Higher Operads, Higher Categories}.
%\newblock London Mathematical Society Lecture Note Series 298, Cambridge
%University Press.
\newblock LMS Lecture Note Series 298, Cambridge
University Press.

\bibitem{LevyCBPV}
P.\,Levy (2005).
\newblock Adjunction models for Call-By-Push-Value with stacks.
\newblock \emph{TAC}, 14(5):75--110.

\bibitem{Linton}
F.\,Linton (1966).
\newblock Some aspects of equational theories.
\newblock In {\em Proc.\ Conf.\ on Categorical Algebra at La Jolla}, pp\,84--95.

\bibitem{Permutads}
J.-L.\,Loday and M.\,Ronco (2012).
\newblock Permutads.
%\newblock In \url{http://arxiv.org/abs/1105.5271}.

\bibitem{LodayVallette}
J.-L.\,Loday and B.\,Vallette (2012).
\newblock \emph{Algebraic Operads}.
%\newblock Grundlehren der mathematischen Wissenschaften, Volume 346. 
\newblock Grundlehren der mathematischen Wissenschaften.

\bibitem{MacLane}
S.\,Mac Lane (1945).
\newblock Categories for the Working Mathematician (Second Edition).
\newblock Springer Verlag, 1998.

\bibitem{Markl}
M.\,Markl (2006). 
\newblock Operads and PROPs.
%\newblock In \url{http://arxiv.org/abs/math/0601129}.

%\bibitem{MartinLofTT}
%P.\,Martin-L{\"o}f (1984).
%\newblock Intuitionistic Type Theory.
%\newblock Bibliopolis.

\bibitem{May}
J.P.\,May (1972).
\newblock \emph{The geometry of iterated loop spaces}.
\newblock LNM 271.

\bibitem{Epigram}
C.\,McBride.
\newblock Epigram: practical programming with dependent types.
\newblock In \emph{AFP'04}, pp\,i130--170.

\bibitem{MelliesCMLL}
P.-A.\,Melli\`es (2009).
\newblock Categorical Semantics of Linear Logic.
\newblock In \emph{Panoramas et synth\`esis}, 27:1--196.

\bibitem{TensorLogic}
P.-A.\,Melli\`es and N.\,Tabareau (2010).
\newblock Resource modalities in tensor logic.
%\newblock \emph{Ann.\ Pure and Appl.\ Logic}, 161(5):632--653.
\newblock \emph{APAL}, 161(5):632--653.

\bibitem{MendlerMMCK}
M.\,Mendler and S.\, Scheele (2011).
\newblock Cut-free Gentzen calculus for multimodal CK.
\newblock \emph{Information and Computation}, 209(12):1465--1490.

\bibitem{MendlerIter}
N.\,Mendler (1991). 
\newblock Inductive types and type constraints in the second-order lambda
calculus. 
%\newblock \emph{Ann.\ Pure Appl.\ Logic}, 51(1--2):159--172.
\newblock \emph{APAL}, 51(1--2):159--172.

\bibitem{Milner1978}
R.\,Milner (1978).
\newblock A Theory of Type Polymorphism in Programming.
%\newblock \emph{Journal of Computer and System Sciences}, 17:348--375.
\newblock \emph{J.\ of Computer and System Sciences}, 17:348--375.

\bibitem{MoggiLambdaC}
E.\,Moggi (1991).
\newblock Notions of computation and monads. 
\newblock \emph{Information And Computation}, 93(1).

\bibitem{Munch}
G.\,Munch-Maccagnoni (2009).
\newblock Focalisation and classical realisability.
\newblock In \emph{CSL'09}, LNCS 5771, pp\,409--423.

\bibitem{ProgMLTT}
B.\,Nordstr\"om, K.\,Petersson, and J.\,Smith (1990).
\newblock Programming in Martin-L\"of Type Theory: An Introduction.
%\newblock Oxford University Press.
\newblock OUP.

\bibitem{ParkIm}
S.\,Park and H.\,Im (2011).
\newblock A modal logic internalizing normal proofs.
\newblock \emph{Information and Computation}, 209:1519--1535.

\bibitem{PlotkinCBVCBN}
G.\,Plotkin (1975).
\newblock Call-by-name, call-by-value and the $\lambda$-calculus.
\newblock \emph{Theoretical Computer Science}, 1:125--159.

\bibitem{PlotkinLCF}
G.\,Plotkin (1977).
\newblock LCF considered as a programming language.
\newblock \emph{Theoretical Computer Science}, 5:223--255.

\bibitem{PlotkinPowerAlgOpsAndGenEffs}
G.\,Plotkin and A.J.\,Power (2003).
\newblock Algebraic operations and generic effects.
\newblock \emph{Applied Categorical Structures} 11.

\bibitem{PlotkinPretnar}
G.\,Plotkin and M.\,Pretnar (2009).
\newblock Handlers of algebraic effects.
\newblock \emph{Programming Languages and Systems}, pp\,80--94.

\bibitem{NatDedForIntNonCommLinLog}
J.\,Polakow and F.\,Pfenning (1999).  
\newblock Natural Deduction for Intuitionistic Non-Commutative Linear
Logic.
\newblock In \emph{TLCA'99}.

\bibitem{Reynolds}
J.\,Reynolds (1983).
\newblock Types, abstraction and parametric polymorphism.
\newblock \emph{Information Processing} 83, pp\,513--523.

\bibitem{RoncoShuffleBialgebras}
M.\,Ronco (2008). 
\newblock Shuffle bialgebras.
%\newblock In \url{http://arxiv.org/abs/math/0703437}.

\bibitem{Russell1902}
B.\,Russell (1902).
\newblock Letter to Frege. 
\newblock In \emph{\cite{vanHeijenoort}}, pp\,124--125. 

\bibitem{Russell1903}
B.\,Russell (1903).
\newblock \emph{The principles of mathematics}.
\newblock Cambridge University Press. 

\bibitem{ScottSolovay}
D.\,Scott (1967).
\newblock A proof of the independence of the continuum hypothesis
\newblock \emph{Theory of Computing Systems}, 1(2):89--111.

\bibitem{ScottTCS}
D.\,Scott (1969).
\newblock A type-theoretical alternative to ISWIM, CUCH, OWHY
%\newblock In \emph{Theoretical Computer Science}, 121(1--2):411--440,
%1993.
\newblock In \emph{TCS}, 121(1--2):411--440, 1993.

\bibitem{Schonfinkel}
M.\,Sch\"{o}nfinkel (1924).
\newblock On the building blocks of mathematical logic.
\newblock In \emph{\cite{vanHeijenoort}}, pp\,355--366.

\bibitem{Seely}
R.\,Seely (1989).
\newblock Linear logic, *-autonomous categories and cofree coalgebras 
\newblock \emph{Contemporary Mathematics} 92. 

\bibitem{Selinger}
P.\,Selinger (2011).
\newblock A survey of graphical languages for monoidal categories.
%\newblock In \emph{New Structures for Physics}, LNP 813, pp\,289--355. 
\newblock In LNP 813, pp\,289--355. 

\bibitem{Omega}
T.\,Sheard (2004).
\newblock Languages of the future.
\newblock In \emph{SIGPLAN Notices}.

\bibitem{Strachey1967}
C.\,Strachey (1967).
\newblock Fundamental Concepts in Programming Languages.  
%\newblock In \emph{Higher-Order and Symbolic Computation}, 13:11--49,
%2000.
\newblock In \emph{HOSC}, 13:11--49, 2000.

\bibitem{Szabo}
M.\,Szabo (1975).
\newblock Polycategories.
\newblock \emph{Communications in Algebra}, 3(8):663--689.

\bibitem{ToHBCurry}
J.\,Seldin and J.R.\,Hindley (1980).
\newblock \emph{To H.B.\,Curry: Essays on Combinatory Logic, Lambda
  Calculus and Formalism}.
%\newblock Academic Press. 
 
\bibitem{Turing}
A.\,Turing (1937). 
\newblock On computable numbers, with an application to the
  Entscheidungsproblem. 
%\newblock \emph{Proc.\ of the London Mathematical Society}, 42:230--265.
\newblock \emph{Proc.\ LMS}, 42:230--265.

\bibitem{vanHeijenoort}
J.\,van Heijenoort, ed.\ (1967).
\newblock \emph{From Frege to G\"odel: A source book in mathematical
  logic, 1879--1931}.
%\newblock Harvard University Press.

\bibitem{EqProofICFP12}
D.\,Vytiniotis, S.\,Peyton~Jones, and J.P.\,Magalhaes (2012). 
\newblock Equality proof and deferred type errors. 
\newblock In \emph{ICFP'12}.

\bibitem{Wadler}
P,\,Wadler (2003).
\newblock Call-by-value is dual to call-by-name.
\newblock In \emph{ICFP'03}, pp\,189--201.

%\bibitem{Haskell}
%The Haskell Programming Language.
%\newblock \url{http://www.haskell.org/haskellwiki/Haskell}.
\end{thebibliography}}

\end{document}
