\documentclass{llncs}
\usepackage{pdfpages}
%% \usepackage[hang,bf]{caption}
\usepackage{color}
\usepackage[fleqn]{amsmath}
\usepackage{amssymb}
\usepackage{bbm}
\usepackage{semantic}
\usepackage{listings}
\usepackage{comment}
\usepackage[numbers]{natbib}

\renewcommand{\qed}{~\hfill$\blacksquare$}

\definecolor{grey}{rgb}{0.8,0.8,0.8}

\newcommand{\Fig}[1]{Fig.\,\ref{fig:#1}}
\newcommand{\Figs}[1]{Figs.\,\ref{fig:#1}}
\newcommand{\newFi}[1]{\colorbox{grey}{\ensuremath{#1}}}

\newcommand{\ie}{{i.e.}}
\newcommand{\eg}{{e.g.}}
\newcommand{\aka}{{a.k.a.}$\,$}

\newcommand{\Fi}{\ensuremath{\mathsf{F}_i}}
\newcommand{\Fw}{\ensuremath{\mathsf{F}_\omega}}
\newcommand{\fix}{\mathsf{fix}}
\newcommand{\Fix}{\mathsf{Fix}}
\newcommand{\Fixw}{\ensuremath{\Fix_{\omega}}}
\newcommand{\Fixi}{\ensuremath{\Fix_{i}}}

\renewcommand{\l}{\lambda}

\newcommand{\dom}{\mathop{\mathsf{dom}}}
\newcommand{\FV}{\mathop{\mathrm{FV}}}

\newcommand{\Eq}{\mathtt{Eq}}
\newcommand{\LEq}{\mathtt{LEq}}
\newcommand{\Ran}{\mathtt{Ran}}
\newcommand{\Lan}{\mathtt{Lan}}
\newcommand{\s}[1]{\{#1\}}

\renewcommand{\S}[0]{Sect.\;}

\title{System \Fi}
\subtitle{a Higher-Order Polymorphic $\lambda$-Calculus\\
        with Erasable Term-Indices}
\titlerunning{System \Fi}
\author{Ki Yung Ahn\inst{1} \and Tim Sheard\inst{1} \and
        Marcelo Fiore\inst{2} \and Andrew M. Pitts\inst{2} }
\institute{
        Portland State University, Portland, Oregon, USA
        \thanks{supported by NSF grant 0910500.}
	\\ \email{\{kya,sheard\}@cs.pdx.edu}
        \and
        University of Cambridge, Cambridge, UK
        \\ \email{\{Marcelo.Fiore,Andrew.Pitts\}@cl.cam.ac.uk}
        }

\begin{document}
\maketitle
\begin{abstract}
We introduce a foundational lambda calculus,
System~\Fi, for studying programming languages with
term-indexed datatypes -- higher-kinded datatypes whose indices range
over data such as 
Natural Numbers %~(e.g. {\tt Z} and {(\tt S Z})) 
or 
Lists. %~(e.g. {\tt Nil} and ({\tt Cons Z Nil})).
System~\Fi\ is a minimal extension of System~\Fw\ that 
introduces the minimal features needed to support term-indexing. We
show that System~\Fi\ provides a theory for analysing programs with term-indexed
types and also argue that it constitutes a basis for the design of logically-sound
light-weight dependent programming languages.
In terms of expressivity, System~\Fi\ sits in between System~\Fw\ and the
Implicit Calculus of Constructions (ICC). Indeed, we relate System~\Fi\ to
System~\Fw\ and ICC as follows.  We establish erasure properties of
\Fi-types that capture the idea that term-indices are discardable in
that they are irrelevant for computation.  Index erasure projects typing
in System~\Fi\ to typing in System~\Fw; so System~\Fi\ inherits the
strong-normalisation property from System~\Fw.  The logical consistency of
System~\Fi\ is established by embedding it into a subset of ICC.


\keywords{term-indexed data types, generalized algebraic data types,
higher-order polymorphism, type-constructor polymorphism, higher-kinded types,
impredicative encoding, strong normalization, logical consistency}
\end{abstract}

\input{intro}

\input{motiv}

\input{fi}

\input{theory}

\input{data}

\input{relwork}

\input{concl}

\bibliographystyle{splncsnat}
\bibliography{main}

%% \newpage
%% \appendix

\end{document}
