\section{Related Work} \label{sec:relwork}
System~\Fi\ is most closely related to
Curry-style System~\Fw \cite{AbeMatUus05,GHR93}
and the Implicit Calculus of Constructions (ICC) \cite{Miquel01}.
All terms typable in a Curry-style System \Fw\ are typable (with the same type) in System \Fi\ 
and all terms typable in \Fi\ are typable (with the same type\footnote{The $*$ kind in \Fw\ and \Fi\ corresponds
        to \textsf{Set} in ICC}) in ICC.

As mentioned in \S\ref{ssec:sn}, we can derive strong normalization of \Fi\ 
from System~\Fw, and derive logical consistency of \Fi\ from certain
restrictions of ICC \cite{Miquel00,BarrasB08}.
In fact, ICC is more than just an extension of System~\Fi\ 
with dependent types and stratified universes, since ICC includes
$\eta$-reduction and $\eta$-equivalence.
We do not foresee any problems adding
$\eta$-reduction and $\eta$-equivalence to System~\Fi.
Although System~\Fi\ accepts fewer terms than ICC, it enjoys simpler
erasure properties (Theorem \ref{thm:ierasetypingall} and
Corollary \ref{cor:ierasetypingifree}) just by looking at the syntax
of kinds and types, which ICC cannot enjoy due to its support for
full dependent types.  In System \Fi, term-indices appearing in types
(\eg,~$s$ in $F\{s\}$) are always erasable.  \citet{LingerS08} generalized the ICC framework to one which describes
erasure on arbitrary Church-style calculi~(EPTS) and Curry-style
calculi~(IPTS), but only consider $\beta$-equivalence for type conversion.


In the practical setting of programming language implementation,
\citet{YorgeyWCJVM12}, inspired by \citet{SHE}, recently designed an extension
to Haskell's GADTs by allowing datatypes to be used as kinds. For instance,
\texttt{Bool} is promoted to a kind (\ie, $\texttt{Bool}:\square$) and its
data constructors \texttt{True} and \texttt{False} are promoted to types.
They extended System $F_{\!C}$ (the Glasgow Haskell Compiler's
intermediate core language) to support \emph{datatype promotion}
and named it System~$F_{\!C}^\uparrow$. The key difference between
$F_{\!C}^\uparrow$ and \Fi\ is in their kind syntax: %%, as illustrated below:
\vspace*{-5pt}
\[\qquad\quad
\begin{array}{ll}
F_{\!C}^\uparrow\,\textbf{kinds}\quad &
\kappa ::= * \mid \kappa -> \kappa \mid F \vec{\kappa} \mid \mathcal{X} \mid \forall \mathcal{X}.\kappa \mid \cdots \\
\,\Fi\,\,\textbf{kinds}\quad &
\kappa ::= * \mid \kappa -> \kappa \mid A -> \kappa \phantom{A^{A^A}}
\end{array}  
\] \vskip-1.5ex\noindent
In $F_{\!C}^\uparrow$, all type constructors ($F$) are promotable to the 
kind level and become kinds when fully applied to other kinds
($F\vec\kappa$). On the other hand, in \Fi,  a type can only appear
as the domain of an index arrow kind ($A-> \kappa$).
The ramifications of this difference is that $F_{\!C}^\uparrow$ can
express type level data structues but not nested term-indices,
while \Fi\ supports the converse. Intuitively,
a type constructor like $\texttt{List}:* -> *$ is promoted to a kind constructor
$\texttt{List}:\square-> \square$, which enables type-level data structures
such as $\mathtt{[Nat,Bool,Nat-> Bool]:List\,*}$. Type-level
data structures motivate type-level computations over promoted data.
This is made possible by type families\footnote{
        A GHC extension to define type-level functions in Haskell.}.
The promotion of polymorphic types naturally motivates
kind polymorphism ($\forall \mathcal{X}.\kappa$).
Kind polymorphism of arbitrary rank is known to
break strong normalization and cause logical inconsistency \cite{Girard72}.
In a {\em programming language}, inconsistency is not an issue.
However, when studying logically consistent systems, we need
a more conservative approach, as in \Fi.

