\section{The Nax language}\label{sec:Nax}

To validate the second part of my thesis -- the Mendler-style can support 
a simple and expressive programming language, we have designed the
programming language Nax.

Nax supports non-recursive datatypes, the recursive type operator
($\mu^\kappa$), and the Mendler-style recursion combinators (\MIt, \MPr,
\McvIt, \McvPr, \MsfIt). The calculi, \Fi\ and \Fixi, discussed
in the previous sections, are theoretically powerful enough to capture
the semantics of both recursive types and Mendler-style recursion schemes.
However, they are far from practical programming languages. Even very simple
datatypes, such as the boolean type, must be defined via impredicative 
Church-style encodings (\eg, an encoding for the boolean type is
$\forall a . a -> a -> a$). Fortunately, it is possible to provide
a more user friendly layer of abstraction over these calculi. Nax supports
datatypes and Mendler-style recursion combinators as primitive constructs of
the language. I will introduce other programmer friendly features of
the Nax language in \S\ref{sec:Nax:tysym}. Then, in \S\ref{sec:Nax:theory},
I will discuss the desired properties of Nax, which will be demonstrated
in the thesis.

\subsection{Type synonyms and fixpoint derivations}\label{sec:Nax:tysym}

Nax was designed with two goals in mind. First it must be directly definable
in terms of these calculi, and it must exhibit a user friendly interface to
programmers. Towards these goals, Nax has a Haskell-like syntax, and
conservatively extends Hindley-Milner type inference. It also supports
type synonyms and fixpoint derivations, which are useful for defining
recursive types more conveniently. The examples in \S\ref{sec:mendler} are
really reformatted Nax programs. If you know Haskell, you will see that
the Nax syntax is similar to Haskell. We have already discussed how Nax
performs type inference (or, type reconstruction) based on user provided
index transformers attached to Mendler-style recursion combinators. I also
mentioned that Nax supports type synonyms, but did not show any details of
the concrete syntax in \S\ref{sec:mendler}. In Figure \ref{fig:naxderiving},
I illustrate the syntax for type synonyms and fixpoint derivations, which make
recursive type definitions more convenient. In the figure I give three versions.
A Haskell version for reference, and the other two versions of Nax programs.
The first Nax version makes explicit the type synonyms (\textit{List})
and constructor functions (\textit{nil} and \textit{cons}).
In the second Nax version we use the fixpoint derivation mechanism.
The two versions define equivalent programs.

In the type synonym version, we abbreviate the fixpoint $\mu^{*}(L\;a)$
using the type synonym $List$, and define constructor functions
$\textit{nil} : a -> \textit{List}\;a$ and
$\textit{cons} : a -> \textit{List}~a -> \textit{List}~a$.
Recall, that we define recursive types using two levels in Nax.  We take
the fixpoint of the base structure (\eg, $L$) to define a recursive type.
The type synonym (\textit{List}), and its constructor functions
\textit{nil} and \textit{cons}, correspond to the Haskell definition of
the recursive datatype (\textit{List}) and its data constructors
(\textit{Nil} and \textit{Cons}).

In the other version using the fixpoint derivation, the derivation syntax
(\textbf{deriving fixpoint} \textit{List}) attached to the definition of
the base datatype ($L$) instructs the Nax implementation to automatically
generate the type synonym (\textit{List}) for the recursive type and
its constructor functions (\textit{nil} and \textit{cons}) with names
derived by dropping the case of the initial alphabetic character.

\begin{figure}
\begin{tabular}{c||c|c}
Haskell  & Nax with synonyms &  Nax with derivation \\
\hline

\begin{minipage}[t]{.28\linewidth}
\small\vspace{.1em}
\textbf{data} \textit{List} $a$\\
$~~~$ $=$ \textit{Nil}\\ 
$~~~$ $\;|\;$ \textit{Cons} $a$ (\textit{List} a)
\vspace{4.9em}\\
$x = \textit{Cons}~3~(\textit{Cons}~2~\textit{Nil}\,)$
\end{minipage} 

& 

\begin{minipage}[t]{.32\linewidth}
\small\vspace{.1em}
\textbf{data} $L : * -> * -> *$ \textbf{where}\\
$~~~$  \textit{Nil}$~~\; :$ $L\;a\;r$\\
$~~~$  \textit{Cons}    $:$ $a -> r -> L\;a\;r$
\vspace{.3em}\\
\textbf{synonym} \textit{List} $a = \mu^{*} (L\;a)$
\vspace{.3em}\\
\textit{nil}$~~~~~~~~~= \In^{*} \textit{Nil}$ \\
\textit{cons}  $x\;xs = \In^{*} (Cons\;x\;xs)$
\vspace{.5em}\\
$x = \textit{cons}~3~(\textit{cons}~2~\textit{nil}\,)$
\end{minipage}

&

\begin{minipage}[t]{.32\linewidth}
\small\vspace{.1em}
\textbf{data} $L : * -> * -> *$ \textbf{where}\\
$~~~$  \textit{Nil}$~~\; :$ $L\;a\;r$\\
$~~~$  \textit{Cons}    $:$ $a -> r -> L\;a\;r$\\
$~~$ \textbf{deriving} \textbf{fixpoint} \textit{List}
\vspace{3.5em}\\
$x = \textit{cons}~3~(\textit{cons}~2~\textit{nil}\,)$
\end{minipage}

\end{tabular}
\caption{Two versions of a recursive type definition of \textit{List} in Nax --
         with a \textit{type synonym} and with a \emph{derivation macro}
         (\cf\ Haskell definition of $List$)}
\label{fig:naxderiving}
\end{figure}

\subsection{Properties of Nax}\label{sec:Nax:theory}

Nax is designed to demonstrate my thesis. As such, the Nax language must
exhibit a number of properties. I list a few of them here.

\paragraph{Type safety:}
Nax, being a typed programming language, should be type safe.
I will prove type safety of Nax in the thesis.

\paragraph{Normalization:}
I will prove normalization of Nax by a reduction preserving embedding into
either \Fi\ or \Fixw, depending on which families of Mendler-style recursion
combinators we decide to include in Nax. Since it is likely that 
primitive recursion is useful in most cases, the eventual target of
the embedding will be \Fixw. But, for simplicity of illustration, I will
initially illustrate an embedding of a subset of Nax, with \MIt\ and \MsfIt\
families only, into \Fi. Then, I will illustrate an embedding of Nax,
including other families, into \Fixw, with gradually increasing complexity.

\paragraph{Type inference:}
There are several properties to show regarding type inference --
soundness and completeness, and most general typing.

I will describe the type inference algorithm for Nax, which is similar to
the $\mathcal{W}$-algorithm for the Hindley-Milner type system, and show
that the type inference algorithm is sound and complete with respect to
the type system of Nax. The proof of soundness and completeness shall be
done in two steps. First, I will define a syntax directed version of
the type system, where the conversion rule is inlined, and show that
the syntax directed system is equivalent to the standard version of
the Nax type system. Then, I will show that the inference algorithm is
sound and complete with respect to the syntax directed system.

I will also show that the type inference algorithm infers the most general type
of a given program with respect to the index transformer annotations provided by
the programmer.


